---
title: "Dynamic Linear Models"
subtitle: "FISH 507 – Applied Time Series Analysis"
author: "Mark Scheuerell"
date: "5 Feb 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(MARSS)
set.seed(123)
data("Nile")

mod_list <- list(B = "identity", U = "zero", Q = matrix("q"),
                 Z = "identity", A = matrix("a"), R = matrix("r"))

fit <- MARSS(matrix(Nile, nrow = 1), mod_list)

```

## Topics for today




## Simple linear regression

Let's begin with a linear regression model

$$
y_i = \alpha + \beta x_i + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$

The index $i$ has no explicit meaning in that shuffling ($y_i,x_i$) pairs has no effect on parameter estimation


## Simple linear regression

We can write the model in matrix form

$$
y_i = \alpha + \beta x_i + e_i \\
\Downarrow \\
y_i = 
\begin{bmatrix}
  1 & x_i
\end{bmatrix}
\begin{bmatrix}
  \alpha \\
  \beta
\end{bmatrix} +
e_i
$$


## Simple linear regression

We can write the model in matrix form

$$
y_i = \alpha + \beta x_i + e_i \\
\Downarrow \\
y_i = 
\begin{bmatrix}
  1 & x_i
\end{bmatrix}
\begin{bmatrix}
  \alpha \\
  \beta
\end{bmatrix} +
e_i \\
\Downarrow \\
y_i = \mathbf{X}^{\top}_i \boldsymbol{\theta} + e_i
$$

with

$\mathbf{X}^{\top}_i = \begin{bmatrix} 1 & x_i \end{bmatrix}$ and $\boldsymbol{\theta} = \begin{bmatrix} \alpha & \beta \end{bmatrix}^{\top}$ 


## Dynamic linear model (DLM)

In a _dynamic_ linear model, the regression parameters change over time, so we write

$$
y_i = \mathbf{X}^{\top}_i \boldsymbol{\theta} + e_i ~~~~~~~ \text{(static)}
$$

as

$$
y_t = \mathbf{X}^{\top}_t \boldsymbol{\theta}_t + e_t ~~~~~~~ \text{(dynamic)}
$$


## Dynamic linear model (DLM)

There are 2 important points here:

$$
y_\boxed{t} = \mathbf{X}^{\top}_t \boldsymbol{\theta}_t + e_t
$$

1. Subscript $t$ explicitly acknowledges implicit info in the time ordering of the data in $\mathbf{y}$  


## Dynamic linear model (DLM)

There are 2 important points here:

$$
y_t = \mathbf{X}^{\top}_t \boldsymbol{\theta}_\boxed{t} + e_t
$$

1. Subscript $t$ explicitly acknowledges implicit info in the time ordering of the data in $\mathbf{y}$  

2. The relationship between $\mathbf{y}$ and $\mathbf{X}$ is unique for every $t$


## Constraining a DLM

Close examination of the DLM reveals an apparent problem for parameter estimation

$$
y_t = \mathbf{X}^{\top}_t \boldsymbol{\theta}_t + e_t
$$


## Constraining a DLM

Close examination of the DLM reveals an apparent problem for parameter estimation

$$
y_t = \mathbf{X}^{\top}_t \boldsymbol{\theta}_t + e_t
$$

We only have 1 data point per time step (ie, $y_t$ is a scalar)

__Thus, we can only estimate 1 parameter (with no uncertainty)!__


## Constraining a DLM

To address this issue, we'll constrain the regression parameters to be dependent from $t$ to $t+1$

$$
\boldsymbol{\theta}_t = \mathbf{G}_t \boldsymbol{\theta}_{t-1} + \mathbf{w}_t ~ \text{with} ~ \mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
$$


## Constraining a DLM

In practice, we often make $\mathbf{G}_t$ time invariant 

$$
\boldsymbol{\theta}_t = \mathbf{G} \boldsymbol{\theta}_{t-1} + \mathbf{w}_t
$$

or assume $\mathbf{G}_t$ is an $m \times m$ identity matrix $\mathbf{I}_m$

$$
\begin{align}
  \boldsymbol{\theta}_t &= \mathbf{I}_m \boldsymbol{\theta}_{t-1} + \mathbf{w}_t \\
                    &= \boldsymbol{\theta}_{t-1} + \mathbf{w}_t
\end{align}
$$

In the latter case, the parameters follow a random walk over time


## DLM in state-space form

Observation model relates covariates to data

$$
y_t = \mathbf{X}^{\top}_t \boldsymbol{\theta}_t + e_t
$$

State model determines how parameters "evolve" over time

$$
\boldsymbol{\theta}_t = \mathbf{G} \boldsymbol{\theta}_{t-1} + \mathbf{w}_t
$$


## DLM in MARSS notation | Full state-space form

$$
y_t = \mathbf{X}^{\top}_t \boldsymbol{\theta}_t + e_t \\
\boldsymbol{\theta}_t = \mathbf{G} \boldsymbol{\theta}_{t-1} + \mathbf{w}_t \\
\Downarrow \\
y_t = \mathbf{Z}_t \mathbf{x}_t + v_t \\
\mathbf{x}_t = \mathbf{B} \mathbf{x}_{t-1} + \mathbf{w}_t
$$


## Contrast in covariate effects

__Note__: DLMs include covariate effect in the observation eqn much differently than other forms of MARSS models

DLM: $\mathbf{Z}$ is covariates, $\mathbf{x}$ is parameters 

$$
y_t = \boxed{\mathbf{Z}_t \mathbf{x}_t} + v_t \\
$$

Others: $\mathbf{d}$ is covariates, $\mathbf{D}$ is parameters

$$
y_t = \mathbf{Z}_t \mathbf{x}_t + \boxed{\mathbf{D} \mathbf{d}_t} +v_t \\
$$


## Other forms of DLMs

The regression model is but one type

Others include:

* stochastic "level" (intercept)

* stochastic "growth" (trend, bias)

* seasonal effects (fixed, harmonic)


## The most simple DLM | Stochastic level

$$
y_t = \alpha_t + e_t \\
\alpha_t = \alpha_{t-1} + w_t
$$


## The most simple DLM | Stochastic level = random walk with obs error

$$
y_t = \alpha_t + e_t \\
\alpha_t = \alpha_{t-1} + w_t \\
\Downarrow \\
y_t = x_t + v_t \\
x_t = x_{t-1} + w_t
$$


## Ex of stochastic level model

```{r}
data("Nile")

par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 0, 0, 0))

plot.ts(Nile, las = 1,
        xlab = "Year", ylab = "Flow of the River Nile")
```


## Ex of stochastic level model


```{r}
par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 0, 0, 0))

plot.ts(Nile, las = 1, lwd = 2,
        xlab = "Year", ylab = "Flow of the River Nile")
lines(seq(start(Nile)[1], end(Nile)[1]),
       lwd = 2, t(fit$states), col = "blue")
lines(seq(start(Nile)[1], end(Nile)[1]), t(fit$states + 2*fit$states.se),
       lwd = 2, lty = "dashed", col = "blue")
lines(seq(start(Nile)[1], end(Nile)[1]), t(fit$states - 2*fit$states.se),
       lwd = 2, lty = "dashed", col = "blue")
```


## Univariate DLM for level & growth

Stochastic "level" $\alpha_t$ with deterministic "growth" $\eta$

$$
y_t = \alpha_t + e_t \\
\alpha_t = \alpha_{t-1} + \eta + w_t \\
$$


## Univariate DLM for level & growth

Stochastic "level" $\alpha_t$ with deterministic "growth" $\eta$

$$
y_t = \alpha_t + e_t \\
\alpha_t = \alpha_{t-1} + \eta + w_t \\
\Downarrow \\
y_t = x_t + v_t \\
x_t = x_{t-1} + u + w_t
$$

This is just a random walk with bias $u$


## Univariate DLM for level & growth

Stochastic "level" $\alpha_t$ with stochastic "growth" $\eta_t$

$$
\begin{align}
  y_t &= \alpha_t + e_t \\
  \alpha_t &= \alpha_{t-1} + \eta_{t-1} + w_{\alpha,t} \\
  \eta_t &= \eta_{t-1} + w_{\eta,t} \\
\end{align}
$$

Now the "growth" term $\eta_t$ evolves as well


## Univariate DLM for level & growth

Evolution of $\alpha_t$ and $\eta_t$

$$
\begin{align}
  \alpha_t &= \alpha_{t-1} + \eta_{t-1} + w_{\alpha,t} \\
  \eta_t &= \eta_{t-1} + w_{\eta,t} \\
  & \Downarrow \\
  \alpha_t &= 1 \alpha_{t-1} + 1 \eta_{t-1} + w_{\alpha,t} \\
  \eta_t &= 0 \alpha_{t-1} + 1 \eta_{t-1} + w_{\eta,t}
\end{align}
$$


## Univariate DLM for level & growth

Evolution of $\alpha_t$ and $\eta_t$

$$
\begin{align}
  \alpha_t &= \alpha_{t-1} + \eta_{t-1} + w_{\alpha,t} \\
  \eta_t &= \eta_{t-1} + w_{\eta,t} \\
  & \Downarrow \\
  \alpha_t &= \underline{1} \alpha_{t-1} + \underline{1} \eta_{t-1} + w_{\alpha,t} \\
  \eta_t &= \underline{0} \alpha_{t-1} + \underline{1} \eta_{t-1} + w_{\eta,t} \\
  & \Downarrow \\
  \underbrace{\begin{bmatrix}
    \alpha_t \\
    \eta_t
  \end{bmatrix}}_{\boldsymbol{\theta}_t} &=
  \underbrace{\begin{bmatrix}
    \underline{1} & \underline{1} \\
    \underline{0} & \underline{1}
  \end{bmatrix}}_{\mathbf{G}}
  \underbrace{\begin{bmatrix}
    \alpha_{t-1} \\
    \eta_{t-1}
  \end{bmatrix}}_{\boldsymbol{\theta}_{t-1}} +
  \underbrace{\begin{bmatrix}
    w_{\alpha,t} \\
    w_{\eta,t}
  \end{bmatrix}}_{\mathbf{w}_t} 
\end{align}
$$


## Univariate DLM for level & growth

Observation model for stochastic "level" & stochastic "growth"

$$
\begin{align}
  y_t &= \alpha_t + v_t \\
  & \Downarrow \\
  y_t &= \underline{1} \alpha_t + \underline{0} \eta_t + v_t \\
  & \Downarrow \\
  y_t &= 
  \underbrace{\begin{bmatrix}
    \underline{1} & \underline{0}
  \end{bmatrix}}_{\mathbf{X}^{\top}_t}
  \underbrace{\begin{bmatrix}
    \alpha_t \\ \eta_t 
  \end{bmatrix}}_{\boldsymbol{\theta}_t}
  + v_t
\end{align}
$$


## Univariate DLM for regression

Stochastic intercept and slope

$$
\begin{align}
  y_t &= \alpha_t + \beta_t x_t + v_t \\
  & \Downarrow \\
  y_t &= \underline{1} \alpha_t + \underline{x_t} \beta_t + v_t \\
  & \Downarrow \\
  y_t &= 
  \underbrace{\begin{bmatrix}
    \underline{1} & \underline{x_t}
  \end{bmatrix}}_{\mathbf{X}^{\top}_t}
  \underbrace{\begin{bmatrix}
    \alpha_t \\ \beta_t 
  \end{bmatrix}}_{\boldsymbol{\theta}_t}
  + v_t
\end{align}
$$


## Univariate DLM for regression

Parameter evolution follows a random walk

$$
\begin{align}
  \alpha_t &= \alpha_{t-1} + w_{\alpha,t} \\
  \beta_t &= \beta_{t-1} + w_{\beta,t} \\
  & \Downarrow \\
  \underbrace{\begin{bmatrix}
    \alpha_t \\
    \beta_t
  \end{bmatrix}}_{\boldsymbol{\theta}_t} &=
  \underbrace{\begin{bmatrix}
    \alpha_{t-1} \\
    \beta_{t-1}
  \end{bmatrix}}_{\boldsymbol{\theta}_{t-1}} +
  \underbrace{\begin{bmatrix}
    w_{\alpha,t} \\
    w_{\beta,t}
  \end{bmatrix}}_{\mathbf{w}_t} 
\end{align}
$$


## Univariate DLM with seasonal effect

Dynamic linear regression with fixed seasonal effect

$$
y_t = \alpha_t + \beta_t x_t + \gamma_{qtr} + e_t \\
\gamma_{qtr} = 
\begin{cases}
  \gamma_{1} & \text{if } qtr = 1 \\
  \gamma_{2} & \text{if } qtr = 2 \\
  \gamma_{3} & \text{if } qtr = 3 \\
  \gamma_{4} & \text{if } qtr = 4
\end{cases}
$$


## Univariate DLM with seasonal effect

$$
y_t = \alpha_t + \beta_t x_t + \gamma_{qtr} + e_t \\
\Downarrow \\
y_t = 
\begin{bmatrix}
  1 & x_t & 1
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_{qtr}
\end{bmatrix}
+ e_t
$$


## Univariate DLM with seasonal effect

$$
y_t = 
\begin{bmatrix}
  1 & x_t & 1
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_{qtr}
\end{bmatrix}
+ e_t \\
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_{qtr}
\end{bmatrix} =
\begin{bmatrix}
  \alpha_{t-1} \\ \beta_{t-1} \\ ?
\end{bmatrix} +
\begin{bmatrix}
  w_{\alpha,t} \\ w_{\beta,t} \\ ?
\end{bmatrix}
$$

How should we model the fixed effect of $\gamma_{qtr}$?


## Univariate DLM with seasonal effect

$$
y_t = 
\begin{bmatrix}
  1 & x_t & 1
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_{qtr}
\end{bmatrix}
+ e_t \\
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_{qtr}
\end{bmatrix} =
\begin{bmatrix}
  \alpha_{t-1} \\ \beta_{t-1} \\ \gamma_{qtr}
\end{bmatrix} +
\begin{bmatrix}
  w_{\alpha,t} \\ w_{\beta,t} \\ 0
\end{bmatrix}
$$

We don't want the effect of quarter to evolve


## Univariate DLM with seasonal effect

$$
y_t = 
\begin{bmatrix}
  1 & x_t & 1
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_{qtr}
\end{bmatrix}
+ e_t \\
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_{qtr}
\end{bmatrix} =
\begin{bmatrix}
  \alpha_{t-1} \\ \beta_{t-1} \\ \gamma_{qtr}
\end{bmatrix} +
\begin{bmatrix}
  w_{\alpha,t} \\ w_{\beta,t} \\ 0
\end{bmatrix}
$$

OK, but how do we select the right quarterly effect?


## Univariate DLM with seasonal effect

Let's separate out the quarterly effects

$$
y_t = \alpha_t + \beta_t x_t + \gamma_1 + \gamma_2 + \gamma_3 + \gamma_4 + e_t \\
\Downarrow \\
y_t = 
\begin{bmatrix}
  1 & x_t & 1 & 1 & 1 & 1
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_1 \\ \gamma_2 \\ \gamma_3 \\ \gamma_4
\end{bmatrix}
$$

But how do we select only the current quarter?


## Univariate DLM with seasonal effect

We can set some values in $\mathbf{x}_t$ to 0 ($qtr$ = 1)

$$
y_t = 
\begin{bmatrix}
  1 & x_t & 1 & 0 & 0 & 0
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_1 \\ \gamma_2 \\ \gamma_3 \\ \gamma_4
\end{bmatrix} \\
\Downarrow \\
y_t = \alpha_t + \beta_t x_t + \gamma_1 + e_t \\
$$


## Univariate DLM with seasonal effect

We can set some values in $\mathbf{x}_t$ to 0 ($qtr$ = 2)

$$
y_t = 
\begin{bmatrix}
  1 & x_t & 0 & 1 & 0 & 0
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_1 \\ \gamma_2 \\ \gamma_3 \\ \gamma_4
\end{bmatrix} \\
\Downarrow \\
y_t = \alpha_t + \beta_t x_t + \gamma_2 + e_t \\
$$


## Univariate DLM with seasonal effect

But _how_ would we set the correct 0/1 values?

$$
\mathbf{X}^{\top}_t =
\begin{bmatrix}
  1 & x_t & ? & ? & ? & ?
\end{bmatrix}
$$


## Univariate DLM with seasonal effect

We could instead reorder the $\gamma_i$ within $\boldsymbol{\theta}_t$ ($qtr$ = 1)

$$
y_t = 
\begin{bmatrix}
  1 & x_t & 1 & 0 & 0 & 0
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_1 \\ \gamma_2 \\ \gamma_3 \\ \gamma_4
\end{bmatrix} \\
\Downarrow \\
y_t = \alpha_t + \beta_t x_t + \gamma_1 + e_t \\
$$


## Univariate DLM with seasonal effect

We could instead reorder the $\gamma_i$ within $\boldsymbol{\theta}_t$ ($qtr$ = 2)

$$
y_t = 
\begin{bmatrix}
  1 & x_t & 1 & 0 & 0 & 0
\end{bmatrix}
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_2 \\ \gamma_3 \\ \gamma_4 \\ \gamma_1
\end{bmatrix} \\
\Downarrow \\
y_t = \alpha_t + \beta_t x_t + \gamma_2 + e_t \\
$$


## Univariate DLM with seasonal effect

But _how_ would we shift the $\gamma_i$ within $\boldsymbol{\theta}_t$?

$$
\boldsymbol{\theta}_t =
\begin{bmatrix}
  \alpha_t \\ \beta_t \\ ? \\ ? \\ ? \\ ?
\end{bmatrix}
$$



## Example of non-diagonal $\mathbf{G}$

We can use a non-diagonal $\mathbf{G}$ to get the correct quarter effect

$$
\mathbf{G} = 
\begin{bmatrix}
  1 & 0 & 0 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 0 & 1 & 0 \\
  0 & 0 & 0 & 0 & 0 & 1 \\
  0 & 0 & 1 & 0 & 0 & 0
\end{bmatrix}
$$


## Evolving parameters

$$
\underbrace{\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_2 \\ \gamma_3 \\ \gamma_4 \\ \gamma_1
\end{bmatrix}}_{\boldsymbol{\theta}_t} =
\underbrace{\begin{bmatrix}
  1 & 0 & 0 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 0 & 1 & 0 \\
  0 & 0 & 0 & 0 & 0 & 1 \\
  0 & 0 & 1 & 0 & 0 & 0
\end{bmatrix}}_{\mathbf{G}}
\underbrace{\begin{bmatrix}
  \alpha_{t-1} \\ \beta_{t-1} \\ \gamma_1 \\ \gamma_2 \\ \gamma_3 \\ \gamma_4
\end{bmatrix}}_{\boldsymbol{\theta}_{t-1}} + 
\underbrace{\begin{bmatrix}
  w_{\alpha,t} \\ w_{\beta,t}  \\ 0  \\0 \\ 0 \\ 0
\end{bmatrix}}_{\mathbf{w}_t}
$$


## Evolving parameters

$$
\underbrace{\begin{bmatrix}
  \alpha_t \\ \beta_t \\ \gamma_3 \\ \gamma_4 \\ \gamma_1 \\ \gamma_2
\end{bmatrix}}_{\boldsymbol{\theta}_t} =
\underbrace{\begin{bmatrix}
  1 & 0 & 0 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 0 & 1 & 0 \\
  0 & 0 & 0 & 0 & 0 & 1 \\
  0 & 0 & 1 & 0 & 0 & 0
\end{bmatrix}}_{\mathbf{G}}
\underbrace{\begin{bmatrix}
  \alpha_{t-1} \\ \beta_{t-1} \\ \gamma_2 \\ \gamma_3 \\ \gamma_4 \\ \gamma_1
\end{bmatrix}}_{\boldsymbol{\theta}_{t-1}} + 
\underbrace{\begin{bmatrix}
  w_{\alpha,t} \\ w_{\beta,t}  \\ 0  \\0 \\ 0 \\ 0
\end{bmatrix}}_{\mathbf{w}_t}
$$


## Forecasting with a DLM

DLMs are often used in a forecasting context where we want a prediction for time $t$ based on the data up through time $t-1$


## Forecasting with a DLM

Pseudo-code

1. get estimate of today's parameters from yesterday's

2. make prediction based on today's parameters & covariates

3. get observation for today

4. update parameters and repeat


## Forecasting with a DLM

1. Define the parameters at time $t = 0$

$$
\boldsymbol{\theta}_0 | y_0 = \mathbf{\Pi} + \mathbf{w}_1 ~ \text{with} ~ \mathbf{w}_1 \sim \text{MVN}(\mathbf{0}, \mathbf{\Lambda}) \\
$$


## Forecasting with a DLM

1. Define the parameters at time $t = 0$

$$
\boldsymbol{\theta}_0 | y_0 = \mathbf{\Pi} + \mathbf{w}_1 ~ \text{with} ~ \mathbf{w}_1 \sim \text{MVN}(\mathbf{0}, \mathbf{\Lambda}) \\
\Downarrow \\
\text{E}(\boldsymbol{\theta}_0) = \mathbf{\Pi} \\
\text{Var}(\boldsymbol{\theta}_0) = \mathbf{\Lambda} \\
\Downarrow \\
\boldsymbol{\theta}_0 | y_0 \sim \text{MVN}(\mathbf{\Pi}, \mathbf{\Lambda})
$$


## Forecasting with a DLM

2. Define the parameters at time $t = 1$

$$
\boldsymbol{\theta}_1 | y_0 = \mathbf{G} \boldsymbol{\theta}_0 + \mathbf{w}_1 ~ \text{with} ~ \mathbf{w}_1 \sim \text{MVN}(\mathbf{0}, \mathbf{Q}) \\
$$


## Forecasting with a DLM

2. Define the parameters at time $t = 1$

$$
\boldsymbol{\theta}_1 | y_0 = \mathbf{G} \boldsymbol{\theta}_0 + \mathbf{w}_1 ~ \text{with} ~ \mathbf{w}_1 \sim \text{MVN}(\mathbf{0}, \mathbf{Q}) \\
\Downarrow \\
  \text{E}(\boldsymbol{\theta}_1) = \mathbf{G} \boldsymbol{\theta}_0 \\
  \text{E}(\boldsymbol{\theta}_1) = \mathbf{G} \mathbf{\Pi} \\
\text{and} \\
  \text{Var}(\boldsymbol{\theta}_1) = \mathbf{G} \text{Var}(\boldsymbol{\theta}_0) \mathbf{G}^{\top} + \mathbf{Q} \\
  \text{Var}(\boldsymbol{\theta}_1) = \mathbf{G} \mathbf{\Lambda} \mathbf{G}^{\top} + \mathbf{Q} \\
\Downarrow \\
\boldsymbol{\theta}_1 | y_0 \sim \text{MVN}(\mathbf{G} \mathbf{\Pi}, \mathbf{G} \mathbf{\Lambda} \mathbf{G}^{\top} + \mathbf{Q})
$$


## Forecasting with a DLM

3. Make a forecast of $y_t$ at time $t = 1$

$$
y_1 | y_0 = \mathbf{X}^{\top}_1 \boldsymbol{\theta}_1 + e_1 ~ \text{with} ~ e_1 \sim \text{N}(0, R)  \\
\Downarrow \\
\text{E}(y_1) = \mathbf{X}^{\top}_1 \boldsymbol{\theta}_1 \\
\text{E}(y_1) = \mathbf{X}^{\top}_1 [\mathbf{G} \mathbf{\Pi}] \\
\text{and} \\
\text{Var}(y_1) = \mathbf{X}^{\top}_1 \text{Var}(\boldsymbol{\theta}_1) \mathbf{X}_1 + R \\
\text{Var}(y_1) = \mathbf{X}^{\top}_1 [\mathbf{G} \mathbf{\Lambda} \mathbf{G}^{\top} + \mathbf{Q}] \mathbf{X}_1 + R \\
\Downarrow \\
y_1 | y_0 \sim \text{N}(\mathbf{X}^{\top}_1 [\mathbf{G} \mathbf{\Pi}], \mathbf{X}^{\top}_1 [\mathbf{G} \mathbf{\Lambda} \mathbf{G}^{\top} + \mathbf{Q}] \mathbf{X}_1 + R)
$$


## Forecasting with a DLM

Putting it all together

$$
\begin{align}
\boldsymbol{\theta}_0 | y_0 & \sim \text{MVN}(\mathbf{\Pi}, \mathbf{\Lambda}) \\
\boldsymbol{\theta}_t | y_{t-1} & \sim \text{MVN}(\mathbf{G} \mathbf{\Pi}, \mathbf{G} \mathbf{\Lambda} \mathbf{G}^{\top} + \mathbf{Q}) \\
y_t | y_{t-1} & \sim \text{N}(\mathbf{X}^{\top}_t [\mathbf{G} \mathbf{\Pi}], \mathbf{X}^{\top}_t [\mathbf{G} \mathbf{\Lambda} \mathbf{G}^{\top} + \mathbf{Q}] \mathbf{X}_t + R)
\end{align}
$$


## Diagnostics for DLMs

Just as with other models, we'd like to know if our fitted DLM meets its underlying assumptions 

We can calcuate the forecast error $e_t$ as

$$
e_t = y_t - \hat{y}_t
$$
and check if

$$
\begin{align}
  (1) & ~~ e_t \sim \text{N}(0, \sigma) \\
  (2) & ~~ \text{Cov}(e_t, e_{t-1}) = 0
\end{align}
$$

with a QQ-plot (1) and an ACF (2)


## {.flexbox .vcenter}

<font size="10">MULTIVARIATE DLMs</font>


## The most simple multivariate DLM | Multiple observations of a stochastic level

$$
\begin{matrix}
  \mathbf{y}_t = \mathbf{Z} \alpha_t + \mathbf{v}_t & ~~ \mathbf{y}_t ~ \text{is } n \times 1 \\
  \alpha_t = \alpha_{t-1} + w_t & ~~ \alpha_t ~ \text{is } 1 \times 1 
\end{matrix}
$$

with

$$
\mathbf{Z} =
\begin{bmatrix}
  1 \\
  1 \\
  \vdots \\
  1
\end{bmatrix}
$$


## Another simple multivariate DLM | Multiple observations of multiple levels

$$
\begin{matrix}
  \mathbf{y}_t = \mathbf{Z} \boldsymbol{\alpha}_t + \mathbf{v}_t & ~~ \mathbf{y}_t ~ \text{is } n \times 1 \\
  \boldsymbol{\alpha}_t = \boldsymbol{\alpha}_{t-1} + \mathbf{w}_t & ~~ \boldsymbol{\alpha}_t ~ \text{is } n \times 1 \\
\end{matrix}
$$

with

$$
\mathbf{Z} =
\mathbf{I}_n =
\begin{bmatrix}
  1 & 0 & \dots & 0 \\
  0 & 1 & \ddots & 0 \\
  \vdots & \ddots & \ddots & 0 \\
  0 & \dots & 0 & 1 \\
\end{bmatrix}
$$


## Multivariate DLMs | Regression model

Our univariate model 

$$
y_t = \mathbf{X}^{\top}_t \boldsymbol{\theta}_t + e_t ~ \text{with} ~ e_t \sim \text{N}(0,R)
$$

becomes

$$
\mathbf{y}_t = (\mathbf{X}^{\top}_t \otimes \mathbf{I}_n) \boldsymbol{\theta}_t + \mathbf{e}_t ~ \text{with} ~ \mathbf{e}_t \sim \text{MVN}(\mathbf{0},\mathbf{R})
$$


## Kronecker products

If $\mathbf{A}$ is an $m \times n$ matrix and $\mathbf{B}$ is a $p \times q$ matrix

then $\mathbf{A} \otimes \mathbf{B}$ will be an $mp \times nq$ matrix

$$
\mathbf{A} \otimes \mathbf{B} = 
\begin{bmatrix}
 a_{11} \mathbf{B} & \dots & a_{1n} \mathbf{B} \\
 \vdots & \ddots & \vdots \\
 a_{m1} \mathbf{B} & \dots & a_{mn} \mathbf{B} \\
\end{bmatrix}
$$

## Kronecker products

For example

$$
\mathbf{A} = 
\begin{bmatrix}
 1 & 2 \\
 3 & 4
\end{bmatrix}
~ \text{and} ~
\mathbf{B} = 
\begin{bmatrix}
 2 & 4 \\
 6 & 8
\end{bmatrix}
$$

so

$$
\mathbf{A} \otimes \mathbf{B} = 
\begin{bmatrix}
 1 \begin{bmatrix}
  2 & 4 \\
  6 & 8
 \end{bmatrix} & 
 2 \begin{bmatrix}
  2 & 4 \\
  6 & 8
 \end{bmatrix} \\
 3 \begin{bmatrix}
  2 & 4 \\
  6 & 8
 \end{bmatrix} &
 4 \begin{bmatrix}
  2 & 4 \\
  6 & 8
 \end{bmatrix}
\end{bmatrix} = 
\begin{bmatrix}
  2 & 4 & 4 & 8\\
  6 & 8 & 12 & 16\\
  6 & 12 & 8 & 16\\
  18 & 24 & 24 & 32
\end{bmatrix}
$$


## Multivariate DLMs | Regression model with $n = 2$

$$
\begin{align}
\mathbf{y}_t &= (\mathbf{X}^{\top}_t \otimes \mathbf{I}_n) \boldsymbol{\theta}_t + \mathbf{e}_t \\
\begin{bmatrix}
 y_{1,t} \\
 y_{2,t}
\end{bmatrix} &= 
\left(
\begin{bmatrix}
 1 & x_t
\end{bmatrix}
\otimes
\begin{bmatrix}
 1 & 0 \\
 0 & 1
\end{bmatrix}
\right)
\begin{bmatrix}
 \alpha_{1,t} \\
 \alpha_{2,t} \\
 \beta_{1,t} \\
 \beta_{2,t}
\end{bmatrix} + 
\begin{bmatrix}
 e_{1,t} \\
 e_{2,t}
\end{bmatrix}
\end{align}
$$


## Multivariate DLMs

$$
\begin{bmatrix}
 y_{1,t} \\
 y_{2,t}
\end{bmatrix} = 
\left(
\begin{bmatrix}
 1 & x_t
\end{bmatrix}
\otimes
\begin{bmatrix}
 1 & 0 \\
 0 & 1
\end{bmatrix}
\right)
\begin{bmatrix}
 \alpha_{1,t} \\
 \alpha_{2,t} \\
 \beta_{1,t} \\
 \beta_{2,t}
\end{bmatrix} + 
\begin{bmatrix}
 e_{1,t} \\
 e_{2,t}
\end{bmatrix} \\
\Downarrow \\
\begin{bmatrix}
 y_{1,t} \\
 y_{2,t}
\end{bmatrix} = 
\begin{bmatrix}
 1 & 0 & x_t & 0 \\
 0 & 1 & 0 & x_t
\end{bmatrix}
\begin{bmatrix}
 \alpha_{1,t} \\
 \alpha_{2,t} \\
 \beta_{1,t} \\
 \beta_{2,t}
\end{bmatrix} + 
\begin{bmatrix}
 e_{1,t} \\
 e_{2,t}
\end{bmatrix}
$$


## Multivariate DLMs | Covariance of observation errors

$$
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma & 0 & 0 & 0 \\
 0 & \sigma & 0 & 0 \\
 0 & 0 & \sigma & 0 \\
 0 & 0 & 0 & \sigma
\end{bmatrix}
~\text{or}~~
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma_1 & 0 & 0 & 0 \\
 0 & \sigma_2 & 0 & 0 \\
 0 & 0 & \sigma_3 & 0 \\
 0 & 0 & 0 & \sigma_4
\end{bmatrix}
$$

$$
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma & \gamma & \gamma & \gamma \\
 \gamma & \sigma & \gamma & \gamma \\
 \gamma & \gamma & \sigma & \gamma \\
 \gamma & \gamma & \gamma & \sigma
\end{bmatrix}
~\text{or}~~
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma_1 & 0 & 0 & 0 \\
 0 & \sigma_2 & 0 & \gamma_{2,4} \\
 0 & 0 & \sigma_3 & 0 \\
 0 & \gamma_{2,4} & 0 & \sigma_4
\end{bmatrix}
$$

## Multivariate DLMs | Parameter evolution

$$
\boldsymbol{\theta}_t = \mathbf{G} \boldsymbol{\theta}_{t-1} + \mathbf{w}_t ~ \text{with} ~ \mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
$$

becomes

$$
\boldsymbol{\theta}_t = \left( \mathbf{G} \otimes \mathbf{I}_n \right) \boldsymbol{\theta}_{t-1} + \mathbf{w}_t ~ \text{with} ~ \mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
$$


## Multivariate DLMs | Parameter evolution

If we have 2 regression parameters and $n = 2$, then

$$
\boldsymbol{\theta}_t =
\begin{bmatrix}
  \alpha_{1,t} \\
  \alpha_{2,t} \\
  \beta_{1,t} \\
  \beta_{2,t} \\
\end{bmatrix}
~~ \text{and} ~~
\mathbf{G} = 
\begin{bmatrix}
 1 & 0 \\
 0 & 1
\end{bmatrix} =
\mathbf{I}_2 
$$


## Multivariate DLMs | Parameter evolution

$$
\begin{align}
\boldsymbol{\theta}_t  &=
\left( \mathbf{G} \otimes \mathbf{I}_n \right) \boldsymbol{\theta}_{t-1} + \mathbf{w}_t \\
& \Downarrow \\
\boldsymbol{\theta}_t  &=
\left( \mathbf{I}_2 \otimes \mathbf{I}_2 \right) \boldsymbol{\theta}_{t-1} + \mathbf{w}_t
\end{align}
$$


## Multivariate DLMs | $\mathbf{I}_m \otimes \mathbf{I}_n = \mathbf{I}_{mn}$

$$
\begin{align}
\mathbf{I}_2 \otimes \mathbf{I}_2 &= 
\begin{bmatrix}
 1 \begin{bmatrix}
  1 & 0 \\
  0 & 1
 \end{bmatrix} & 
 0 \begin{bmatrix}
  1 & 0 \\
  0 & 1
 \end{bmatrix} \\
 0 \begin{bmatrix}
  1 & 0 \\
  0 & 1
 \end{bmatrix} &
 1 \begin{bmatrix}
  1 & 0 \\
  0 & 1
 \end{bmatrix} \\
\end{bmatrix} \\
 &=
\begin{bmatrix}
 1 & 0 & 0 & 0 \\
 0 & 1 & 0 & 0 \\
 0 & 0 & 1 & 0 \\
 0 & 0 & 0 & 1
\end{bmatrix}
\end{align}
$$


## Multivariate DLMs | Parameter evolution

$$
\begin{align}
\boldsymbol{\theta}_t  &=
\left( \mathbf{G} \otimes \mathbf{I}_n \right) \boldsymbol{\theta}_{t-1} + \mathbf{w}_t \\
\boldsymbol{\theta}_t  &=
\left( \mathbf{I}_2 \otimes \mathbf{I}_2 \right) \boldsymbol{\theta}_{t-1} + \mathbf{w}_t \\
\begin{bmatrix}
  \alpha_{1,t} \\
  \alpha_{2,t} \\
  \beta_{1,t} \\
  \beta_{2,t} \\
\end{bmatrix} &=
\begin{bmatrix}
 1 & 0 & 0 & 0 \\
 0 & 1 & 0 & 0 \\
 0 & 0 & 1 & 0 \\
 0 & 0 & 0 & 1
\end{bmatrix}
\begin{bmatrix}
  \alpha_{1,t-1} \\
  \alpha_{2,t-1} \\
  \beta_{1,t-1} \\
  \beta_{2,t-1} \\
\end{bmatrix} + 
\begin{bmatrix}
  w_{\alpha_1,t} \\
  w_{\alpha_2,t} \\
  w_{\beta_1,t} \\
  w_{\beta_2,t}
\end{bmatrix} \\
\boldsymbol{\theta}_t  &=
 \boldsymbol{\theta}_{t-1} + \mathbf{w}_t
\end{align}
$$





