---
title: "Week 9 lab: Spatiotemporal modeling"
author: "Eric"
date: "3/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(mgcv)
```

For this lab, we're going to be working off of a couple datasets from lecture. 

## 1. Spatiotemporal modeling of tick presence absence

The first exercise is based on the NEON dataset on tick counts, collected 
throughout the US. We'll subset the data to the core sites, and years with the 
most data, and only look at modeling a single genus (IXOSP2) and 
life stage (larvae).

### Loading and cleaning the tick data

```{r tickdata}
# this data is from the neon tick github repo, 
# https://github.com/eco4cast/neon4cast-ticks
dat.raw <- readRDS("tck_longform.Rdata")
dat = dplyr::filter(dat.raw, Species_LifeStage == "IXOSP2_Larva") 
# only use sites 
dat = dplyr::filter(dat, decimalLongitude > -100, decimalLongitude < -80, 
                    decimalLatitude > 30, decimalLatitude < 40)

# sum up spp g
dat_summary = dplyr::group_by(dat, plotID, collectDate) %>% 
  dplyr::summarize(ticks = sum(IndividualCount),
                   lon = decimalLongitude[1],
                   lat = decimalLatitude[1])

# add in date
dat_summary$month = lubridate::month(dat_summary$collectDate)
dat_summary$year = lubridate::year(dat_summary$collectDate)
dat_summary$week = ceiling(lubridate::yday(dat_summary$collectDate)/7)

# filter out years 2016-2018, with most data
dat_summary = dplyr::filter(dat_summary, year %in% 2016:2018)
# create a presence-absence variable
dat_summary$presence <- ifelse(dat_summary$ticks > 0, 1, 0)
```

### Fitting Spatial GAMs to model occurrence

Throughout the model fitting, it's a good idea to rememember some of our
basic tools for diagnostics. Things to look at are using functions like

```{r, eval=FALSE}
acf(residuals(fit))
pacf(residuals(fit))
qqnorm(residuals(fit))
```

A non-spatial null model can be built for these data, using a simple GLM. 
Ignoring site,

```{r}
null_fit <- glm(presence ~ as.factor(year), 
                data = dat_summary, 
                family = "binomial")
```

But this model ignores the seasonal variation (week) and spatial variation. 

Next, we can try to add a non-spatial GAM that has a smooth term over the week 
effects,

```{r}
fit <- gam(presence ~ as.factor(year) + s(week), 
           data = dat_summary, 
           family = "binomial")
```

Using AIC, does this model do any better than the null model? 

How would you modify this code to test a hypothesis about changing seasonal 
effects across years? [There's no right answer for this -- it depends in part 
whether you want to model the effect as being linear, non-linear, and whether 
or not random effects are included]

With seasonality generally improving the predictions of tick occurrence, let's 
try to implement a spatial model. Starting with a 2-d model, with a single 
spatial field,

```{r}
spatial_fit <- gam(presence ~ as.factor(year) + s(week) + s(lon,lat), 
           data = dat_summary, 
           family = "binomial")
```

Next, we can try adding separate spatial fields by year -- we do this with the 
'by' argument,
```{r}
spatial_fit2 <- gam(presence ~ as.factor(year) + s(week) + 
                     s(lon,lat,by=as.factor(year)), 
           data = dat_summary, 
           family = "binomial", 
           control=list(maxit=1000))
```

Is there evidence of the occurence varying by year? 

A slight modification of the spatial fields by year is to include spatial fields
by year and constrain the wiggliness of each be the same,

```{r}
spatial_fit3 <- gam(presence ~ as.factor(year) + s(week) + 
                     s(lon,lat) + 
                      s(lon,lat,as.factor(year),bs="fs",m=2), 
           data = dat_summary, 
           family = "binomial",
           control=list(maxit=1000))
```

Question: how else can you improve the model to get a lower AIC value? 


## 2. Spatiotemporal modeling of snow data in Washington state

For this exercise, we're going to switch gears and try to develop a spatio-temporal 
model using sdmTMB and INLA. 

We'll be working with the SNOTEL data from lecture, and using only data from February.

```{r}
d = read.csv("SNOTEL_Washington_1981_2013.csv")
d = d[,c("Station.Id","Station.Name","Water.Year","Feb")]
meta = read.csv("SNOTEL_metadata.csv")
d = dplyr::left_join(d, meta) %>% 
  dplyr::filter(!is.na(Water.Year), !is.na(Feb))
```

To fit a model with sdmTMB, first we need to generate a mesh object

```{r eval=FALSE}
library(sdmTMB)
mesh <- make_mesh(d, xy_cols = c("Longitude","Latitude"), n_knots = 50)
```

In practice, this may not be an ideal mesh -- but we'll use it for this lab
```{r eval = FALSE}
plot(mesh)
```

### Fitting a spatiotemporal model with sdmTMB

First, we'll start with a model that doesn't include a spatiotemporal effect

```{r eval=FALSE}
fit <- sdmTMB(Feb ~ 1, spde = mesh, time = NULL, data = d)
```

We can extend this model to include spatiotemporal processes -- first as an 
independent process, 

```{r eval=FALSE}
fit <- sdmTMB(Feb ~ 1, spde = mesh, time = Water.Year, data = d)
```

Or secondly, as an autoregressive spatial process, 

```{r eval=FALSE}
fit <- sdmTMB(Feb ~ 1, spde = mesh, time = Water.Year, data = d, ar1_fields=TRUE)
```

Is there generally support for spatiotemporal effects in this model? What about the AR(1) parameter? 


### Diagnostics

Finally, we'll examine whether there's residual spatial autocorrelation in the residuals -- this may
indicate a covariate that should have been included that wasn't. 

We'll use a statistic called Moran's I to test for spatial autocorrelation,

```{r eval=FALSE}
install.packages("ape")
library(ape)
?Morans.I
```

So for our model of snowfall, the first thing we'd want to do is construct a matrix 'm' of distances (starting with data for
just 1 year). It's important that the diagonal of this matrix be 0. The general form  may be something like this, where

```{r eval=FALSE}
d$resid = residuals(fit)
df = dplyr::filter(d,Water.Year==2003)
```

And then our matrix of weights could be computed like this,

```{r eval =FALSE}
m <- as.matrix(dist(df[,c("Longitude","Latitude")]))
m <- 1/m
diag(m) <- 0
Moran.I(df$resid, weight=m)
```

Taking a subset of years, e.g. 2002-2005, is there support for residual autocorrelation?