---
title: "Forecating with ARIMA models"
author: "Eli Holmes"
date: "14 Jan 2021"
output:
  beamer_presentation: default
  ioslides_presentation:
    css: lecture_slides.css
subtitle: FISH 507 – Applied Time Series Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(R.options=list(width=60))
set.seed(123)
```


```{r load_data, message=FALSE, warning=FALSE, echo=FALSE}
load("landings.RData")
landings$log.metric.tons = log(landings$metric.tons)
landings = subset(landings, Year <= 1989)
anchovy = subset(landings, Species=="Anchovy")$log.metric.tons
anchovyts = ts(anchovy, start=c(1964,1), frequency = 1)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(tseries)
library(urca)
```

## Forecasting with an ARIMA model

The basic idea of forecasting with an ARIMA model to estimate the parameters and forecast forward.

For example, let's say we want to forecast with this ARIMA(2,1,0) model:
$$y_t = \mu + \beta_1 y_{t-1} + \beta_2 y_{t-2} + e_t$$
where $y_t = x_t - x_{t-1}$, the first difference.

`Arima()` would write this model:
$$(y_t-m) = \beta_1 (y_{t-1}-m) + \beta_2 (y_{t-2}-m) + e_t$$
The relationship between $\mu$ and $m$ is $\mu = m(1 - \beta_1 - \beta_2)$.

##

Let's estimate the $\beta$'s for this model from the anchovy data.

```{r}
fit <- forecast::Arima(anchovyts, order=c(2,1,0), include.constant=TRUE)
coef(fit)
```

```{r}
mu <- coef(fit)[3]*(1-coef(fit)[1]-coef(fit)[2])
mu
```

##

So we will forecast with this model:

$$y_t = 0.1065807-0.53850433 y_{t-1} - 0.44732522 y_{t-2} + e_t$$

To get our forecast for 1990, we do this

$$(x_{90}-x_{89}) = 0.106 - 0.538 (x_{89}-x_{88}) - 0.447 (x_{88}-x_{87})$$

Thus

$$x_{90} = x_{89}+0.106 -0.538  (x_{89}-x_{88}) - 0.447 (x_{88}-x_{87})$$

##

Here is R code to do that:

```{r}
anchovyts[26]+mu+coef(fit)[1]*(anchovyts[26]-anchovyts[25])+
  coef(fit)[2]*(anchovyts[25]-anchovyts[24])
```

## Forecasting with `forecast()`

`forecast(fit, h=h)` automates the forecast calculations for us and computes the upper and lower prediction intervals.  Prediction intervals include uncertainty in parameter estimates plus the process error uncertainty.

```{r}
fr <- forecast::forecast(fit, h=5)
fr
```

## Plotting our forecasts

```{r}
plot(fr, xlab="Year")
```

## Missing values

Missing values are allowed for `Arima()` and `arima()`. We can produce forecasts with the same code.

```{r fig.show=FALSE}
anchovy.miss <- anchovyts
anchovy.miss[10:11] <- NA
anchovy.miss[20:21] <- NA
fit <- forecast::auto.arima(anchovy.miss)
fr <- forecast::forecast(fit, h=5)
fr
```

##

```{r}
plot(fr)
```


## Forecasting with a Seasonal model

### Load the chinook salmon data

```{r}
load("chinook.RData")
chinookts <- ts(chinook$log.metric.tons, start=c(1990,1), 
                frequency=12)
```

## Plot seasonal data

```{r}
plot(chinookts)
```

## Seasonal ARIMA model

Seasonally differenced data, e.g. chinook data January 1990 - chinook data January 1989.

$$z_t = x_t - x_{t+s} - m$$

Basic structure of a seasonal AR model

$z_t$ = AR(p) + AR(season) + AR(p+season)

e.g. $z_t$ = AR(1) + AR(12) + AR(1+12)


Example AR(1) non-seasonal part + AR(1) seasonal part

$$z_t = \phi_1 z_{t-1} + \Phi_1 z_{t-12} - \phi_1\Phi_1 z_{t-13}$$

## Notation

ARIMA (p,d,q)(ps,ds,qs)S


ARIMA (1,0,0)(1,1,0)[12]

## `auto.arima()` for seasonal ts

`auto.arima()` will recognize that our data has season and fit a seasonal ARIMA model to our data by default.  We will define the training data up to 1998 and use 1999 as the test data.

```{r}
traindat <- window(chinookts, c(1990,10), c(1998,12))
testdat <- window(chinookts, c(1999,1), c(1999,12))
fit <- forecast::auto.arima(traindat)
fit
```

## Forecast using seasonal model

```{r}
fr <- forecast::forecast(fit, h=12)
plot(fr)
points(testdat)
```

## Missing values

Missing values are ok when fitting a seasonal ARIMA model

```{r echo=FALSE}
plot(fr)
```
