---
title: "Intro to ARIMA models"
author: "Eli Holmes"
date: "6 Apr 2023"
output:
  beamer_presentation: default
  ioslides_presentation:
    css: lecture_slides.css
subtitle: |
  FISH 550 – Applied Time Series Analysis \ 
  [Download Rmd](https://atsa-es.github.io/atsa/Lectures/Week%202/lec_04_Intro_ARIMA.Rmd)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(R.options=list(width=60))
set.seed(123)
```

## Big picture

Let's imagine that we can describe our data as a combination of the mean trend ($m_t$) and error.

$$x_t = m_t + e_t$$

Fisheries biologists and ecologists often want to know $m_t$. *In fact, that is often our main and sole goal*.  

But let's say we don't care about $m_t$. Our only goal is to predict $x_{t+1}$.  

How could we do this?

## Approach 1 which we will use in the rest of the course

1. We write a model for $m_t$ and estimate that model from the data.  
2. Once we have an estimated model for $m_t$, we have an estimate of $e_t$.  
3. We can model that error (think AR and MA)
4. With 2 and 3, we can predict the future $x_t$ from $x_{t-1}$.

## Approach 2 Box-Jenkins

The Box-Jenkins approach (ARIMA models) is **totally** different. 

1. Keep  differencing the data to until you get a new transformed stationary time series $\Delta^d x_t$
2. Any stationary time series can be modeled as a ARMA process (Wold Decomposition). So now fit a ARMA model to $\Delta^d x_t$
3. Using the estimated ARMA, predict $\Delta^d x_{t+1}$
4. Using $\Delta^d x_{t+1}$, $x_t$, $x_{t-1}$, $x_{t-2}$, etc, you can compute $x_{t+1}$
5. That's the prediction!

## What's $\Delta^d x_{t+1}$

$$\Delta^1 x_t = x_t - x_{t-1}$$
$$\Delta^2 x_t = \Delta^1 x_t - \Delta^1 x_{t-1}$$
$$\Delta^3 x_t = \Delta^2 x_t - \Delta^2 x_{t-1}$$

## In Box-Jenkin's approach $m_t$ is gone

In this approach to predicting $x_{t+1}$, we remove $m_t$ from our data using differencing.  

We don't have to worry about a model for $m_t$ because we have removed it!!  

## How is that possible?

You can remove any wiggly trend with enough differencing.

```{r echo=FALSE, message=FALSE}
library(tidyr)
library(ggplot2)
t <- seq(-3,2,0.01)
m <- (t+3)*(t-2)^2*(t+1)^3
df <- data.frame(x = 1:(length(t)-2), d0=m[1:(length(t)-2)], d1 = diff(m)[1:(length(t)-2)], d2 = diff(m, differences = 2))
df <- df %>% pivot_longer(cols = d0:d2, values_to = "value")
ggplot(df, aes(x=x, y=value)) + geom_line() + facet_wrap(~name)
```

## A biological interpretation is hard for ARIMA models

The error structure of $\Delta^d x_{t+1}$ is NOT the same as $e_t$. 

$$\Delta^d x_{t} = \phi_1\Delta^d x_{t-1} + \phi_2\Delta^d x_{t-2} + \dots + z_t$$

$z_t$ is the error of the differences.  And the $\phi$ in the AR part are for the differences not the original $x_t$.  

But remember, the objective was to predict $x_{t+1}$ not to fit a model with a biological interpretation.


## Box-Jenkins method for fitting ARIMA model

1. Make data stationary by differencing the data
2. Fit a ARMA model to the differenced data
3. Estimate the ARMA parameters
4. Predict $x_{t+1}$
5. Assess the residuals for problems

## Other approaches for non-stationary time series data

ARIMA models are one approach for fitting data that have underlying trends.

Other approaches

* Regression (we won't cover this)
* Dynamic Linear Regression (we will cover this)
* Stochastic level models (we will do a lot of variants of this in class)
* ARMAX models: $x_t = b x_{t-1} + \beta \text{covariates} + \text{error}$ (we will do some of this)



