---
title: "Seasonal ARIMA models"
author: "Eli Holmes"
date: "6 Apr 2023"
output:
  beamer_presentation: default
  ioslides_presentation:
    css: lecture_slides.css
subtitle: |
  FISH 550 – Applied Time Series Analysis \ 
  Download [Rmd](https://atsa-es.github.io/atsa/Lectures/Week%202/lec_04_Seasonal_ARIMA_models.Rmd)  [pdf](https://atsa-es.github.io/atsa/Lectures/Week%202/lec_04_Seasonal_ARIMA_models.[pdf])
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(R.options=list(width=60))
set.seed(123)
# ref
# https://tsmatz.wordpress.com/2017/08/04/time-series-arima-r-tutorial-02-unit-root-seasonal-arima/
```


```{r load_data, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
library(gridExtra)
library(reshape2)
library(tseries)
library(urca)
```


## Seasonality

### Load the chinook salmon data set

```{r}
load("chinook.RData")
head(chinook)
```

##

The data are monthly and start in January 1990.  To make this into a ts object do

```{r}
chinookts <- ts(chinook$log.metric.tons, start=c(1990,1), 
                frequency=12)
```
`start` is the year and month and frequency is the number of months in the year.  

Use `?ts` to see more examples of how to set up ts objects.

## Plot seasonal data

```{r}
plot(chinookts)
```

## Seasonal ARIMA model SARIMA

Seasonally differenced data, e.g. chinook data January 1990 - chinook data January 1989.

$$z_t = x_t - x_{t+s} - m$$

Basic structure of a seasonal AR model

seasonal differenced data = AR(seasonal) + AR(non-seasonal) + AR(p+season)

##

For example, SARIMA model could capture this process

1. AR(seasonal) January in $t$ is correlated with January in $t-1$
2. AR(non-seasonal) January differences are correlated with February differences
3. AR(p+season) appears because of 1 and 2

##



$z_t$ = AR(1) + AR(12) + AR(1+12)

$$z_t = \phi_1 z_{t-1} + \Phi_1 z_{t-12} - \phi_1\Phi_1 z_{t-13}$$

## Notation

ARIMA (p,d,q)(ps,ds,qs)S

ARIMA (non-seasonal part)(seasonal part)Frequency

ARIMA (non-seasonal) means $t$ correlated with $t-1$

ARIMA (seasonal) means $t$ correlated with $t-s$


## Examples

ARIMA (1,0,0)(1,0,0)[12]

What data are we modeling? Get the differences

$$z_t = x_t - m$$

Write out the AR parts with $z_t$

$$z_t = \phi_1 z_{t-1} + \Phi_1 z_{t-12} - \phi_1\Phi_1 z_{t-13} + w_t$$

Write out the MA parts, the $w_t$. No MA in this model.

$$w_t = e_t$$

##

ARIMA (1,0,0)(1,1,0)[12]

Figure out $z_t$. Just a seasonal difference.

$$z_t = x_t - x_{t-12} - m$$

Write out the AR parts with $z_t$

$$z_t = \phi_1 z_{t-1} + \Phi_1 z_{t-12} - \phi_1\Phi_1 z_{t-13} + w_t$$

Write out the MA parts, the $w_t$. No MA in this model. $w_t$ is white noise.

$$w_t = e_t$$


## Seasonal random walk model

ARIMA(0,0,0)(0,1,0)[12]

expected January 1990 = January 1989 + constant mean

Figure out $z_t$. $m$ is the mean seasonal difference.

$$z_t = x_t - x_{t-12} - m$$

Write out the AR parts with $z_t$. No AR part.

$$z_t = w_t$$

Write out the MA parts, the $w_t$.

$$w_t = e_t$$

## Seasonal random walk model with random trend

ARIMA(0,1,0)(0,1,0)[12]

expected Feb 1990 = Feb 1989 + (Jan 1990 - Jan 1989)

Figure out $z_t$. $m$ is the mean seasonal difference.

$$z_t = (x_t - x_{t-12}) - (x_{t-1} - x_{t-13}) - m$$

Write out the AR parts with $z_t$. No AR part.

$$z_t = w_t$$

Write out the MA parts, the $w_t$.

$$w_t = e_t$$


## airline model

ARIMA(0, 1, 1)(0, 1, 1)[12]

Figure out $z_t$. 

$$z_t = (x_t - x_{t-12}) - (x_{t-1} - x_{t-13}) - m$$

Write out the AR parts with $z_t$. No AR part.

$$z_t = w_t$$

Write out the MA parts, the $w_t$.

$$w_t = e_t - \theta_1 e_{t-1} - \Theta_1 e_{t-12} + \theta_1\Theta_1 e_{t-13}$$

## Example with longer lags

ARIMA (2,0,1)(1,0,2)[12]

What data are we modeling? Get the differences

$$z_t = x_t - m$$

Write out the AR parts with $z_t$

$$z_t = \phi_1 z_{t-1} + \phi_2 z_{t-2} + \Phi_1 z_{t-12} - (\text{cross products}) + w_t$$

Write out the MA parts, the $w_t$. 

$$w_t = e_t - \theta_1 e_{t-1} - \Theta_1 e_{t-12} - \Theta_2 e_{t-24} + (\text{cross products})$$

## `auto.arima()` for seasonal ts

`auto.arima()` will recognize that our data has season and fit a seasonal ARIMA model to our data by default.  We will define the training data up to 1998 and use 1999 as the test data.

```{r}
traindat <- window(chinookts, c(1990,10), c(1998,12))
testdat <- window(chinookts, c(1999,1), c(1999,12))
fit <- forecast::auto.arima(traindat)
fit
```


## Summary for seasonal models

Basic steps for identifying a seasonal model.  **forecast** automates most of this.

* Check that you have specified your season correctly in your ts object.

* Plot your data.  Look for trend, seasonality and random walks.

## Summary

* Use differencing to remove season and trend.
    * Season and no trend.  Take a difference of lag = season
    * No seasonality but a trend.  Try a first difference
    * Both. Do both types of differences
    * Neither. No differencing
    * Random walk. First difference
    * Parametric looking curve. Transform

## Summary

* Examine the ACF and PACF of the differenced data.
    * Look for patterns (spikes) at seasonal lags

* Estimate likely models and compare with model selection criteria (or cross-validation). Use `TRACE=TRUE`

* Do residual checks 

## Forecasting with a Seasonal model

### Load the chinook salmon data

```{r}
load("chinook.RData")
chinookts <- ts(chinook$log.metric.tons, start=c(1990,1), 
                frequency=12)
```

## Plot seasonal data

```{r}
plot(chinookts)
```

## Seasonal ARIMA model

Seasonally differenced data, e.g. chinook data January 1990 - chinook data January 1989.

$$z_t = x_t - x_{t+s} - m$$

Basic structure of a seasonal AR model

$z_t$ = AR(p) + AR(season) + AR(p+season)

e.g. $z_t$ = AR(1) + AR(12) + AR(1+12)


Example AR(1) non-seasonal part + AR(1) seasonal part

$$z_t = \phi_1 z_{t-1} + \Phi_1 z_{t-12} - \phi_1\Phi_1 z_{t-13}$$

## Notation

ARIMA (p,d,q)(ps,ds,qs)S


ARIMA (1,0,0)(1,1,0)[12]

## `auto.arima()` for seasonal ts

`auto.arima()` will recognize that our data has season and fit a seasonal ARIMA model to our data by default.  We will define the training data up to 1998 and use 1999 as the test data.

```{r}
traindat <- window(chinookts, c(1990,10), c(1998,12))
testdat <- window(chinookts, c(1999,1), c(1999,12))
fit <- forecast::auto.arima(traindat)
fit
```

## Forecast using seasonal model

```{r}
fr <- forecast::forecast(fit, h=12)
plot(fr)
points(testdat)
```

## Missing values

Missing values are ok when fitting a seasonal ARIMA model

```{r echo=FALSE}
plot(fr)
```


