---
title: "Fitting and Selecting ARIMA models"
subtitle: "FISH 507 – Applied Time Series Analysis"
author: "Eli Holmes"
date: "17 Jan 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(123)
```


```{r load_data, message=FALSE, warning=FALSE, echo=FALSE}
load("landings.RData")
landings$log.metric.tons = log(landings$metric.tons)
landings = subset(landings, Year <= 1989)
anchovy = subset(landings, Species=="Anchovy")$log.metric.tons
sardine = subset(landings, Species=="Sardine")$log.metric.tons

library(ggplot2)
library(gridExtra)
library(reshape2)
library(tseries)
library(urca)
```


## Box-Jenkins method

A. Model form selection

  1. Evaluate stationarity and seasonality
  2. Selection of the differencing level (d)
  3. Selection of the AR level (p)
  4. Selection of the MA level (q)

B. Parameter estimation

C. Model checking

## Good news

Much of the Box-Jenkins method will be automated with the forecast package functions, which we will use in the lab.

## Stationarity

Stationarity means 'not changing in time' in the context of time-series models.  Typically we test the trend and variance, however more generally all statistical properties of a time-series is time-constant if the time series is 'stationary'.

## Example

Many ARMA models exhibit stationarity.  White noise is one type:
$$x_t = e_t, e_t \sim N(0,\sigma)$$

```{r fig.stationarity, fig.height = 3.5, fig.width = 7, fig.align = "center", echo=FALSE}
require(gridExtra)
require(reshape2)

TT=100
y = rnorm(TT)
dat = data.frame(t=1:TT, y=y)
p1 = ggplot(dat, aes(x=t, y=y)) + geom_line() + 
  ggtitle("White Noise") + xlab("") + ylab("value")
ys = matrix(rnorm(TT*10),TT,10)
ys = data.frame(ys)
ys$id = 1:TT

ys2=melt(ys, id.var="id")
p2 = ggplot(ys2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("The variance of a white noise process is steady")
grid.arrange(p1, p2, ncol = 1)
```

## Example

An AR-1 process with $-1<b<1$
$$x_t = b x_{t-1} + e_t$$
is also stationary.

```{r fig.stationarity2, fig.height = 3.5, fig.width = 7, fig.align = "center", echo=FALSE}
require(ggplot2)
require(reshape2)
theta=0.8
nsim=10
ar1=as.vector(arima.sim(TT, model=list(ar=theta)))
dat = data.frame(t=1:TT, y=ar1)
p1 = ggplot(dat, aes(x=t, y=y)) + geom_line() + 
  ggtitle("AR-1") + xlab("") + ylab("value")
ys = matrix(0,TT,nsim)
for(i in 1:nsim) ys[,i]=as.vector(arima.sim(TT, model=list(ar=theta)))
ys = data.frame(ys)
ys$id = 1:TT

ys2=melt(ys, id.var="id")
p2 = ggplot(ys2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("The variance of an AR-1 process is steady")
grid.arrange(p1, p2, ncol = 1)
```

## Stationarity around non-zero mean

We can also have stationarity around a non-zero level or around a linear trend.

```{r fig.stationarity3, fig.height = 4.5, fig.width = 7, fig.align = "center", echo=FALSE}
require(ggplot2)
require(gridExtra)
intercept = .5
trend=.1
dat = data.frame(t=1:TT, wn=rnorm(TT))
dat$wni = dat$wn+intercept
dat$wnti = dat$wn + trend*(1:TT) + intercept
p1 = ggplot(dat, aes(x=t, y=wn)) + geom_line() + ggtitle("White noise")
p2 = ggplot(dat, aes(x=t, y=wni)) + geom_line() + ggtitle("with non-zero mean")
p3 = ggplot(dat, aes(x=t, y=wnti)) + geom_line() + ggtitle("with linear trend")

ar1 = rep(0,TT)
err = rnorm(TT)
for(i in 2:TT) ar1[i]=theta*ar1[i-1]+err[i]
dat = data.frame(t=1:TT, ar1=ar1)
dat$ar1i = dat$ar1+intercept
dat$ar1ti = dat$ar1 + trend*(1:TT) + intercept
p4 = ggplot(dat, aes(x=t, y=ar1)) + geom_line() + ggtitle("AR1")
p5 = ggplot(dat, aes(x=t, y=ar1i)) + geom_line() + ggtitle("with non-zero mean")
p6 = ggplot(dat, aes(x=t, y=ar1ti)) + geom_line() + ggtitle("with linear trend")

grid.arrange(p1, p4, p2, p5, p3, p6, ncol = 2)
```

## Mathematically it looks like this

AR-1

1. Non-zero mean adds $\mu$: $x_t = \mu + b x_{t-1} + e_t$
2. Linear trend adds $at$: $x_t = \mu + at + b x_{t-1} + e_t$

White noise ($b=0$)

1. Non-zero mean: $x_t = \mu + e_t$
2. Linear trend: $x_t = \mu + at + e_t$


## Non-stationarity

One of the most common forms of non-stationarity that is tested for is that the process is a random walk $x_t = x_{t-1} + e_t$.  A test for an underlying random walk is called a 'unit root' test.


```{r fig.nonstationarity, fig.height = 4, fig.width = 8, fig.align = "center", echo=FALSE}
require(ggplot2)
require(reshape2)

rw = rep(0,TT)
for(i in 2:TT) rw[i]=rw[i-1]+err[i]
dat = data.frame(t=1:TT, rw=rw)
p1 = ggplot(dat, aes(x=t, y=rw)) + geom_line() + 
  ggtitle("Random Walk") + xlab("") + ylab("value")
rws = apply(matrix(rnorm(TT*nsim),TT,nsim),2,cumsum)
rws = data.frame(rws)
rws$id = 1:TT

rws2=melt(rws, id.var="id")
p2 = ggplot(rws2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("The variance of a random walk process grows in time")
grid.arrange(p1, p2, ncol = 1)
```

## Random walk with $\mu$ and $at$ added

Similar to the way we added an intecept and linear trend to the stationarity process equations, we can do the same to the random walk equation.

1. Non-zero mean or intercept: $x_t = \mu + x_{t-1} + e_t$

2. Linear trend: $x_t = \mu + at + x_{t-1} + e_t$

## Random walk with $\mu$ and $at$ added

The effects are fundamentally different however.  The addition of $\mu$ leads to a upward mean linear trend while the addition of $at$ leads to exponential growth (or decline).

```{r fig.stationarity4, fig.height = 4, fig.width = 7, fig.align = "center", echo=FALSE}
require(ggplot2)
require(gridExtra)
dat = data.frame(t=1:TT, y=cumsum(rnorm(TT)))
dat$yi = cumsum(rnorm(TT,intercept,1))
dat$yti = cumsum(rnorm(TT,intercept+trend*1:TT,1))
p1 = ggplot(dat, aes(x=t, y=y)) + geom_line() + ggtitle("Random Walk")
p2 = ggplot(dat, aes(x=t, y=yi)) + geom_line() + ggtitle("with non-zero mean added")
p3 = ggplot(dat, aes(x=t, y=yti)) + geom_line() + ggtitle("with linear trend added")

grid.arrange(p1, p2, p3, ncol = 1)
```

## Testing for stationarity

Why is evaluating stationarity important? 

- Many AR models have a flat level or trend and time-constant variance.  If your data do not have those properties, you are fitting a model that is fundamentally inconsistent with your data.
- Many standard algorithms for fitting ARIMA models assume stationarity.  Note, you can fit ARIMA models without making this assumption, but you need to use the appropriate algorithm.

## Testing for stationarity

We will discuss three common approaches to evaluating stationarity:

- Visual test
- (Augmented) Dickey-Fuller test
- KPSS test


## Visual test

The visual test is simply looking at a plot of the data versus time.  Look for

- Change in the level over time.  Is the time series increasing or decreasing?  Does it appear to cycle?
- Change in the variance over time.  Do deviations away from the mean change over time, increase or decrease?


## Example anchovy and sardine catch in Greek waters

```{r fig.vis, fig.height = 4, fig.width = 8, fig.align = "center", echo=FALSE}
require(ggplot2)
dat = subset(landings, Species %in% c("Anchovy", "Sardine") & 
               Year <= 1989)
dat$log.metric.tons = log(dat$metric.tons)
ggplot(dat, aes(x=Year, y=log.metric.tons)) +
  geom_line()+facet_wrap(~Species)
```

## Dickey-Fuller test

The Dickey=Fuller test (and Augmented Dickey-Fuller test) look for evidence that the time series has a unit root.  

The **null hypothesis** is that the time series has a unit root, that is, it has a random walk component.  

The **alternative hypothesis** is some variation of stationarity.  The test has three main verisons. 

## Dickey-Fuller nulls and alternatives

It is hard to see but in the panels on the left, the variance around the trend is increasing and on the right, it is not.

```{r fig.df, fig.height = 4.5, fig.width = 8, fig.align = "center", echo=FALSE}
require(ggplot2)
require(gridExtra)
#####
ys = matrix(0,TT,nsim)
for(i in 2:TT) ys[i,]=ys[i-1,]+rnorm(nsim)
rws = data.frame(ys)
rws$id = 1:TT
library(reshape2)
rws2=melt(rws, id.var="id")
p1 = ggplot(rws2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("Null Non-stationary", subtitle="Random walk")
ys = matrix(0,TT,nsim)
for(i in 2:TT) ys[i,]=theta*ys[i-1,]+rnorm(nsim)
ar1s = data.frame(ys)
ar1s$id = 1:TT
library(reshape2)
ar1s2=melt(ar1s, id.var="id")
p2 = ggplot(ar1s2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("Alternate Stationary", subtitle="AR1")
#####
ys = matrix(intercept,TT,nsim)
for(i in 2:TT) ys[i,]=intercept+ys[i-1,]+rnorm(nsim)
rws = data.frame(ys)
rws$id = 1:TT
library(reshape2)
rws2=melt(rws, id.var="id")
p3 = ggplot(rws2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("", subtitle="Random walk + drift")
ys = matrix(intercept/(1-theta),TT,nsim)
for(i in 2:TT) ys[i,]=intercept+theta*ys[i-1,]+rnorm(nsim)
ar1s = data.frame(ys)
ar1s$id = 1:TT
library(reshape2)
ar1s2=melt(ar1s, id.var="id")
p4 = ggplot(ar1s2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("Alternate", subtitle="AR1 + non-zero level")
#####
ys = matrix(intercept+trend*1,TT,nsim)
for(i in 2:TT) ys[i,]=intercept+trend*i+ys[i-1,]+rnorm(nsim)
rws = data.frame(ys)
rws$id = 1:TT
library(reshape2)
rws2=melt(rws, id.var="id")
p5 = ggplot(rws2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("", subtitle="Random walk + exponential drift")
ys = matrix((intercept+trend*1)/(1-theta),TT,nsim)
for(i in 2:TT) ys[i,]=intercept+trend*i+theta*ys[i-1,]+rnorm(nsim)
ar1s = data.frame(ys)
ar1s$id = 1:TT
library(reshape2)
ar1s2=melt(ar1s, id.var="id")
p6 = ggplot(ar1s2, aes(x=id,y=value,group=variable)) +
  geom_line() + xlab("") + ylab("value") +
  ggtitle("Alternate", subtitle="AR1 + linear trend")
#####
grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2)
```

## Dickey-Fuller null and alternatives

In math, here are the null and alternative hypotheses as equations.  In each, we are testing if $\delta=0$.

1. Null $x_t = x_{t-1}+e_t$

Alternative $x_t = \delta x_{t-1}+e_t$

2. Null $x_t = \mu + x_{t-1} + e_t$

Alternative $x_t = \mu + \delta x_{t-1} + e_t$

3. Null $x_t = \mu + at + x_{t-1} + e_t$

Alternative $x_t = \mu + at + \delta x_{t-1} + e_t$


## Example: Dickey-Fuller tests on the anchovy time series

`adf.test()` in the tseries package will apply the Augemented Dickey-Fuller and report the p-value.  We want to reject the Dickey=Fuller null hypothesis of non-stationarity.  We will set `k=0` to apply the Dickey-Fuller test which tests for AR(1) stationarity.  The Augmented Dickey-Fuller tests for more general lag-p stationarity.

```
adf.test(x, alternative = c("stationary", "explosive"),
         k = trunc((length(x)-1)^(1/3)))
```

Here is how to apply this test to the anchovy data

```{r adf.test.anchovy1}
adf.test(anchovy, k=0)
```

The null hypothesis is not rejected.  That is not what we want.

## Example: Dickey-Fuller tests on the anchovy time series

The `urca` R package can also be used to apply the Dickey-Fuller tests.  Use `lags=0` for Dickey-Fuller which tests for AR(1) stationarity.  We will set `type="trend"` to deal with the trend seen in the anchovy data.  Note, `adf.test()` uses this type by default.

```
ur.df(y, type = c("none", "drift", "trend"), lags = 0)
```

```{r dickey.fuller, message=FALSE, warning=FALSE}
require(urca)
test = ur.df(anchovy, type="trend", lags=0)
test
```

## Example: Dickey-Fuller tests on the anchovy time series

`ur.df()` will report the test statistic.  You can look up the values of the test statistic for different $\alpha$ levels using `summary(test)` or `attr(test, "cval")`.  If the test statistic is less than the critical value for $\alpha$=0.05 ('5pct' in cval), it means the null hypothesis of non-stationarity is rejected.  For the Dickey-Fuller test, you do want to reject the null hypothesis.

The test statistic is 
```{r teststat}
attr(test, "teststat")
```
and the critical value at $\alpha = 0.05$ is 
```{r cval}
attr(test,"cval")
```
The statistic is larger than the critical value and thus the null hypothesis of non-stationarity is not rejected. That's not what we want.

## Augmented Dickey-Fuller test

The Dickey-Fuller test assumes that the stationary process is AR(1) (autoregressive lag-1).  The Augmented Dickey-Fuller test allows a general stationary process.  The idea of the test however is the same.

We can apply the Augmented Dickey-Fuller test with the `ur.df()` function or the `adf.test()` function in the `tseries` package.

```
adf.test(x, alternative = c("stationary", "explosive"),
         k = trunc((length(x)-1)^(1/3)))
```

The alternative is either stationary like $x_t = \delta x_{t-1} + \eta_t$ with $\delta<1$ or 'explosive' with $\delta>1$.  `k` is the number of lags which determines the number of time lags allowed in the autoregression.  `k` is generally determined by the length of your time series.

## Example: Augmented Dickey-Fuller tests with adf.test()

With the `tseries` package, we apply the Augmented Dickey-Fuller test with `adf.test()`.  This function uses the test where the alternative model is stationary around a linear trend: $x_t = \mu + at + \delta x_{t-1} + e_t$.

```{r dickey.fuller2, message=FALSE, warning=FALSE}
require(tseries)
adf.test(anchovy)
```

In both cases, we do not reject the null hypothesis that the data have a random walk.  Thus there is not support for these time series being stationary.


## Example: Augmented Dickey-Fuller tests with ur.df()

With the `urca` package, we apply the Augmented Dickey-Fuller test with `ur.df()`.  The defaults for `ur.df()` are different than for `adf.test()`.  

`ur.df()` allows you to specify which of the 3 alternative hypotheses you want: none (stationary around 0), drift (stationary around a non-zero intercept), trend (stationary around a linear trend).

Another difference is that by default, `ur.df()` uses a fixed lag of 1 while by default `adf.test()` selects the lag based on the length of the time series.

## Example: Augmented Dickey-Fuller tests with ur.df()

We will specify "trend" to make the test similar to `adf.test()`.  We will set the lags like `adf.test()` does also.

```{r dickey.fuller.ur.df, message=FALSE, warning=FALSE}
require(urca)
k = trunc((length(anchovy)-1)^(1/3))
test = ur.df(anchovy, type="trend", lags=k)
test
```

The test statistic values are the same, but we need to look up the critical values with `summary(test)`.

## KPSS test

In the Dickey-Fuller test, the null hypothesis is the unit root, i.e. random walk.  Often times, there is not enough power to reject the null hypothesis.  A null hypothesis is accepted unless there is strong evidence against it.  

The Kwiatkowski–Phillips–Schmidt–Shin (KPSS) test has as the null hypothesis that a time series is stationary around a level trend (or a linear trend). The alternative hypothesis for the KPSS test is a random walk.

The stationarity assumption is general; it does not assume a specific type of stationarity such as white noise.

If both KPSS and Dickey-Fuller tests support non-stationarity, then the stationarity assumption is not supported.

## Example: KPSS tests

```{r kpss.test, message=FALSE, warning=FALSE}
require(tseries)
kpss.test(anchovy, null="Trend")
```

Here `null="Trend"` was included to account for the increasing trend in the data.  The null hypothesis of stationarity is rejected.  Thus both the KPSS and Dickey-Fuller tests support the hypothesis that the anchovy time series is non-stationary.  That's not what we want.


## Differencing the data to make the mean stationary

Differencing means to create a new time series  $z_t = x_t - x_{t-1}$.  First order differencing means you do this once (so $z_t$) and second order differencing means you do this twice (so $z_t - z_{t-1}$).

The `diff()` function takes the first difference:

```{r diff1}
x <- diff(c(1,2,4,7,11))
x
```

The second difference is the first difference of the first difference.

```{r diff2}
diff(x)
```

## Anchovy catch first differenced

Here is a plot of the anchovy data and its first difference.

```{r diff1.plot, fig.align="center", fig.height = 4, fig.width = 8}
par(mfrow=c(1,2))
plot(anchovy, type="l")
title("Anchovy")
plot(diff(anchovy), type="l")
title("Anchovy first difference")
```


## Stationarity of anchovy catch first differences

Let's test the anchovy data with one difference using the KPSS test.

```{r diff.anchovy.test}
diff.anchovy = diff(anchovy)
kpss.test(diff.anchovy)
```

The null hypothesis of stationairity is not rejected. That is good.

## Stationarity of anchovy catch first differences

Let's test the first difference of the anchovy data using the Augmented Dickey-Fuller test.  We do the default test and allow it to chose the number of lags.

```{r test.dickey.fuller.diff1}
adf.test(diff.anchovy)
```

The null hypothesis of non-stationarity is not rejected.  That is not what we want.  However, we differenced which removed the trend thus we are testing against a more general model than we need.  Let's test with an alternative hypothesis that has a non-zero mean.  We can do this with `ur.df()` and `type='drift'`.

```{r test.dickey.fuller.diff2}
test <- ur.df(diff.anchovy, type="drift", lags=2)
```

## Stationarity of anchovy catch first differences

The null hypothesis of NON-stationairity IS rejected. That is good.

The test statistic is 
```{r teststat.diff}
attr(test, "teststat")
```

and the critical value at $\alpha = 0.05$ is 
```{r cval.diff}
attr(test,"cval")
```

## Summary

Test stationarity before you fit a ARMA model.

- Visual test: is the time series flutuating about a level or a linear trend?

Yes or maybe?  Apply a "unit root" test.

- (Augmented) Dickey-Fuller test
- KPSS test

No or fails the unit root test.

- Apply differencing again and re-test.

Still not passing?

- Try a second difference.

Still not passing?

- ARMA model might not be the best choice. Or you may need to an adhoc detrend.

