---
title: "Forecasting with ARIMA models"
author: "Eli Holmes"
date: "10 Apr 2025"
output:
  ioslides_presentation:
    widescreen: true
    css: lecture_slides.css
  beamer_presentation: default
subtitle: |
  FISH 550 – Applied Time Series Analysis \ 
  Download [Rmd](https://atsa-es.github.io/atsa/Lectures/Week%202/lec_04_Forecasting_ARIMA.Rmd) [pdf](https://atsa-es.github.io/atsa/Lectures/Week%202/lec_04_Forecasting_ARIMA.pdf)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(R.options=list(width=60))
set.seed(123)
```


```{r load_data, message=FALSE, warning=FALSE, echo=FALSE}
load("landings.RData")
landings$log.metric.tons = log(landings$metric.tons)
landings = subset(landings, Year <= 1989)
anchovy = subset(landings, Species=="Anchovy")$log.metric.tons
anchovyts = ts(anchovy, start=c(1964,1), frequency = 1)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(tseries)
library(urca)
```

## Forecasting with an ARIMA model

The basic idea of forecasting with an ARIMA model to estimate the parameters and forecast forward.

For example, let's say we want to forecast with a ARIMA(2,1,0) model with drift:
$$z_t = \mu + \beta_1 z_{t-1} + \beta_2 z_{t-2} + e_t$$
where $z_t = x_t - x_{t-1}$, the first difference.

`Arima()` would write this model:
$$(z_t-m) = \beta_1 (z_{t-1}-m) + \beta_2 (z_{t-2}-m) + e_t$$
The relationship between $\mu$ and $m$ is $\mu = m(1 - \beta_1 - \beta_2)$.

##

Let's estimate the $\beta$'s for this model from the anchovy data.

```{r}
fit <- forecast::Arima(anchovyts, order=c(2,1,0), include.constant=TRUE)
coef(fit)
```

```{r}
mu <- coef(fit)[3]*(1-coef(fit)[1]-coef(fit)[2])
mu
```

##

So we will forecast with this model:

$$z_t = 0.1065807-0.53850433 z_{t-1} - 0.44732522 z_{t-2} + e_t$$

To get our forecast for 1990, we do this

$$(x_{90}-x_{89}) = 0.106 - 0.538 (x_{89}-x_{88}) - 0.447 (x_{88}-x_{87})$$

Thus

$$x_{90} = x_{89}+0.106 -0.538  (x_{89}-x_{88}) - 0.447 (x_{88}-x_{87})$$

## Forecasting with `forecast()`

`forecast(fit, h=h)` automates the forecast calculations for us and computes the upper and lower prediction intervals.  Prediction intervals include uncertainty in parameter estimates plus the process error uncertainty.

```{r}
fr <- forecast::forecast(fit, h=5)
fr
```

## Plotting our forecasts

```{r}
plot(fr, xlab="Year")
```

## Missing values

Missing values are allowed for `forecast::Arima()`. We can produce forecasts with the same code.

```{r fig.show=FALSE}
anchovy.miss <- anchovyts
anchovy.miss[10:11] <- NA
anchovy.miss[20:21] <- NA
fit <- forecast::Arima(anchovy.miss, order=c(2,1,0), include.constant=TRUE)
fr <- forecast::forecast(fit, h=5)
fr
```

##

```{r}
plot(fr)
```

## Using auto.arima()

We can let forecast to select the ARIMA model:

```{r fig.show=FALSE}
anchovy.miss <- anchovyts
anchovy.miss[10:11] <- NA
anchovy.miss[20:21] <- NA
fit <- forecast::auto.arima(anchovy.miss)
fit
```

##

```{r}
fr <- forecast::forecast(fit, h=5)
plot(fr)
```


## Seasonal ARIMA Models


```{r setup2, include=FALSE}
set.seed(123)
```


### Load the chinook salmon data set

```{r}
load("chinook.RData")
head(chinook)
```

##

The data are monthly and start in January 1990.  To make this into a ts object do

```{r}
chinookts <- ts(chinook$log.metric.tons, start=c(1990,1), 
                frequency=12)
```
`start` is the year and month and frequency is the number of months in the year.  

Use `?ts` to see more examples of how to set up ts objects.

## Plot seasonal data

```{r}
plot(chinookts)
```

## Seasonal ARIMA model: SARIMA

Basic structure of a seasonal AR model

differenced data = AR(non-seasonal) + AR(seasonal) + cross-terms + MA

## Differenced chinook data

* No differencing $x_t$. We are modeling the data.
* $x_t - x_{t-1}$. We are modeling the monthly change.
* $x_t - x_{t-12}$. We are modeling the seasonal change. Jan vs Jan last year, Feb vs Feb last year
* $(x_t - x_{t-12}) - (x_{t-1} - x_{t-13})$. We are modeling the seasonal rate of monthly change.
* $(x_t - x_{t-1}) - (x_{t-12} - x_{t-13})$. We are modeling the seasonal monthly change.



## Notation

ARIMA (p,d,q)(ps,ds,qs)S

ARIMA (non-seasonal part)(seasonal part)Frequency

ARIMA (non-seasonal) means $t$ correlated with $t-1$

ARIMA (seasonal) means $t$ correlated with $t-s$


## Seasonal autoregressive

ARIMA (1,0,0)(1,0,0)[12]

*This month depends on last month plus the same month last year*

What data are we modeling? Zero differences, so

$$z_t = x_t$$

Write out the AR parts with $z_t$

$$z_t = \phi_1 z_{t-1} + \Phi_1 z_{t-12} - \text{cross-term} + w_t$$

No MA in this model so $w_t = e_t$ is white noise.


## Autocorrelated seasonal differences

ARIMA (1,0,0)(1,1,0)[12]

*The seasonal difference this month depends on the seasonal difference last month plus the seasonal difference last year*


$z_t$ is just a seasonal difference.

$$z_t = x_t - x_{t-12}$$

Write out the AR parts with $z_t$

$$z_t = \phi_1 z_{t-1} + \Phi_1 z_{t-12} - \text{cross-term} + w_t$$

No MA in this model so $w_t = e_t$ is white noise.


## Seasonal random walk model

ARIMA(0,0,0)(0,1,0)[12]

*This month’s value is the same as last year’s, plus some random noise*

expected January 1990 = January 1989 + constant mean

$$z_t = x_t - x_{t-12}$$

No AR nor MA

$$z_t = w_t = x_t - x_{t-12}$$
$$x_t = x_{t-12} + w_t$$


## Seasonal random walk model with random trend

ARIMA(0,1,0)(0,1,0)[12]

expected Feb 1990 = Feb 1989 + (Jan 1990 - Jan 1989)

$$z_t = (x_t - x_{t-12}) - (x_{t-1} - x_{t-13})$$

No AR or MA part.

$$z_t = w_t = (x_t - x_{t-12}) - (x_{t-1} - x_{t-13})$$
$$x_t = x_{t-12} + (x_{t-1} - x_{t-13}) + w_t$$

## airline model

ARIMA(0, 1, 1)(0, 1, 1)[12]

Seasonal random walk model with random trend + autocorrelated noise

$$z_t = (x_t - x_{t-12}) - (x_{t-1} - x_{t-13})$$

No AR part. Write out the MA parts, the $w_t$.

$$w_t = e_t - \theta_1 e_{t-1} - \Theta_1 e_{t-12} + \text{cross-term}$$

## `auto.arima()` for seasonal ts

`auto.arima()` will recognize that our data has season and fit a seasonal ARIMA model to our data by default.  We will define the training data up to 1998 and use 1999 as the test data.

```{r}
traindat <- window(chinookts, c(1990,10), c(1998,12))
testdat <- window(chinookts, c(1999,1), c(1999,12))
fit <- forecast::auto.arima(traindat)
fit
```

## Forecasting with a Seasonal model

### Load the chinook salmon data

```{r}
library(atsalibrary)
chinook <- chinook.month %>% subset(State == "WA")
chinookts <- ts(chinook$log.metric.tons, start=c(1990,1), 
                frequency=12)
```

## Plot seasonal data

```{r}
plot(chinookts)
```

## Fit

Missing values are ok when fitting a seasonal ARIMA model

```{r}
traindat <- window(chinookts, c(1990,10), c(1998,12))
testdat <- window(chinookts, c(1999,1), c(1999,12))
fit <- forecast::auto.arima(traindat)
fit
```

## Forecast using seasonal model

```{r}
fr <- forecast::forecast(fit, h=12)
plot(fr)
points(testdat)
```


