---
title: "HW 2 Box-Jenkins Chapter Key"
author: "EE Holmes"
date: "1/28/2019"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 2
---

```{r mlr-setup, include=FALSE, purl=FALSE}
knitr::opts_knit$set(unnamed.chunk.label = "bjsolns-")
```

Here are the answers for the homework problems based on the material in [Box-Jenkins method in R](https://nwfsc-timeseries.github.io/atsa-labs/chap-labboxjenkins-.html).  You can download the Rmd file for this key from [here](HW_2_box-jenkins_Key.Rmd).


# Solutions Chapter 5
\chaptermark{Solutions Ch 5}

## Question 1

_Because the information on how to do a Dickey-Fuller test was buried in the material that I gave you and not in the help file for these functions, I did not grade problem 1._


**Repeat the stationarity tests for another species in the landings data set.  Sardine, Chub.mackerel, and Horse.mackerel have enough data. Here is how to set up the data for another species.**
```{r get.another.species}
load("landings.RData")
datdf <- subset(landings, Species=="Chub.mackerel" & Year<=1989)
dat <- ts(datdf$log.metric.tons, start=1964)
```

a. **Do a Dickey-Fuller test using `ur.df()` and `adf.test()`. What do you need to set the lags to to achieve that?**
  
    You need to set the lags to 0 to do a Dickey-Fuller test. The Dickey-Fuller test assumes an AR(1) process for the stationary process while an Augmented Dickey-Fuller test allows AR($k+1$) where $k$ is the lag value that you pass into the test function.  The $k$ is referring to the $\delta$ lags in the test equation:
    $$\Delta y_t = \alpha + \beta t + \gamma y_{t-1} + \delta_1 \Delta y_{t-1} + \delta_2 \Delta y_{t-2} + \dots$$
    Setting lags = 0 will be this model and we are testing if $\gamma = 0$.

    $$\Delta y_t = \alpha + \beta t + \gamma y_{t-1}$$
    ```{r}
    tseries::adf.test(dat, k=0)
    ```
    The null hypothesis of non-stationarity is not rejected, i.e. we cannot reject that $\gamma=0$.

    Getting the information on whether the null is rejected is harder for `ur.df()`.
    ```{r}
    library(urca)
    test <- ur.df(dat, lags=0, type="trend")
    summary(test)
    ```
    Ignore the part in the summary with the p-value.  You need to look at the test-statistic and see if the `tau` statistic is greater than the critical value at 5pct.  The first test statistic is `tau3`, the second `phi2` and the third is `phi3`.  The test statistic for `tau3` is -2.89 and is greater than -3.50. Thus the null of $\gamma=0$ is not rejected at $\alpha=0.05$.
    
    For our purposes, we just need to focus on the `tau`. The `phi` are for other null hypotheses involving the presence of absence of the drift and trend terms.
    
b. **Do an Augmented Dickey-Fuller test using `ur.df()`. How did you choose to set the lags?**
    
    There are a few different approaches to use to set the lags. You could use the rule of thumb that `adf.test()` uses which is `trunc((length(x)-1)^(1/3))` which is 2 for 26 data points.  
    
    ```{r}
    urca::ur.df(dat, lags=2, type="trend")
    ```
    
    You could also let `ur.df()` select the best lag based on AIC or BIC. It chooses 2 also.  Look for the number of `z.diff.lag`s in the summary table.

    ```{r}
    library(urca)
    test<-ur.df(dat, lags=8, type="trend", selectlags="AIC")
    summary(test)
    ```

c. **Do a KPSS test using `kpss.test()`.**

    ```{r}
    tseries::kpss.test(dat, null="Trend")
    ```
    The null of trend stationary is rejected.  Again the data do pass the stationarity tests.

## Question 2

**Repeat the stationarity tests and differencing tests for anchovy using the full data.  The full data go to 2007.** 
```{r read_data_prob2}
load("landings.RData")
datdf <- subset(landings, Species=="Anchovy")
anchovy <- ts(datdf$log.metric.tons, start=1964)
```

a. Do the conclusions regarding stationarity and the amount of differencing needed change? 

b. Repeat with only the data after 1987.  Are the conclusions different for only the recent data?

c. Fit the using `auto.arima()` to the full data (1964-2007) and recent data (1987 to present).  Do the selected models change?

## Question 3

**Fit each of the models listed under the `auto.arima()` trace using `Arima()` and show that you can produce the same AICc value that is shown in the trace table.**

```{r results='hide'}
forecast::auto.arima(anchovy, trace=TRUE)
```

## Question 4

**Use `auto.arima()` with `stepwise=FALSE` to find the top 3 models (according to AICc) for the anchovy.**

a. **Use `Arima()` to fit the top 3 models to the data.**

b. **Create a 5-year forecast for each of the top 3 models using the fits from (a).**

c. **Do the forecasts differ?**
    
## Question 5

a. **Fit a seasonal model to the Chinook data up to 1999 using `auto.arima()`.**

b. **Then create a forecast through 2015.**

c. **Plot the forecast with the 2014 and 2015 actual landings added as data points.**
