---
title: 'Homework #1'
author: "Your Name"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

`r format(Sys.time(), '%d %b %Y')`


# An example Knitr/R Markdown document

This is a super simple template to show you how to use knitr and rmarkdown to create a html file with your R code and output. Look at http://rmarkdown.rstudio.com/authoring_basics.html to see how to make lists, tables, italics, bold, etc.

```{r knitr_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=12, fig.height=4, fig.path='./', warning=FALSE, message=FALSE)
```

## Problem 1

This is my solution to problem 1: add together 2 numbers. I decided to do $1+1$. It was not very hard. 
```{r basic_addition}
1+1
```

## Problem 2

This question asked me to add together the numbers `r 1:9`, so $\sum_{i=1}^9 i$. This question was also easy using 
 [Smith et al., Adv Numerics 10:5&ndash;12,
2002]

```{r add_numbers}
sum(1:9)
```

## Problem 3

Use the `lm` function to do a linear regression using the example in `?lm`.

```{r lm}
## Annette Dobson (1990) "An Introduction to Generalized Linear Models".
## Page 9: Plant Weight Data.
ctl = c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
trt = c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
group = gl(2, 10, 20, labels = c("Ctl","Trt"))
weight = c(ctl, trt)
lm.D9 = lm(weight ~ group)
```

We can use `summary` to get a summary.

```{r summary_lm}
summary(lm.D9)
```

## Problem 4

Use `plot()` to get a summary plot.

```{r summary_plot, fig.height=8}
opar <- par(mfrow = c(2,2), oma = c(0, 0, 1.1, 0))
plot(lm.D9, las = 1)      # Residuals, Fitted, ...
par(opar)
```
## Problem 5

This problem asked me to write a function to do $$\sqrt{b^2 - 4ac}$$.

```{r function}
myfun = function(a,b,c){return(sqrt(b^2-4*a*c))}
myfun(1,3,1)
```

## Problem 6

This problem asked me to write out a matrix equation $\mathbf{A}\mathbf{B}$ with $\mathbf{A}$ as a $3 \times 2$ matrix and $\mathbf{B}$ as a $2 \times 2$ matrix.  I chose this equation:
$$
\mathbf{A}\mathbf{B}=
\begin{bmatrix} 1 & 4\\ 2 & 5 \\ 3 & 6\end{bmatrix}
\begin{bmatrix} 3 & 0 \\ 0 & 3 \end{bmatrix}
$$
Here's my R code:

```{r matrix_math}
A=matrix(1:6,3,2)
B=diag(3,2)
A%*%B
```
