---
title: "Dynamic Factor Analysis"
author: "Mark Scheuerell"
date: "20 April 2023"
output:
  ioslides_presentation:
    css: lecture_slides.css
subtitle: FISH 550 – Applied Time Series Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(123)
```

## Topics for today

Deterministic vs stochastic elements

Regression with autocorrelated errors

Regression with temporal random effects

Dynamic Factor Analysis (DFA)

* Forms of covariance matrix  

* Constraints for model fitting

* Interpretation of results


## Code for today

You can find the R code for these lecture notes and other related exercises [here](lec_08_intro_to_DFA.R).


## A very simple model

Consider this simple model, consisting of a mean $\mu$ plus error

$$
y_i = \mu + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$


## A very simple model

The right-hand side of the equation is composed of _deterministic_ and _stochastic_ pieces

$$
y_i = \underbrace{\mu}_{\text{deterministic}} + \underbrace{e_i}_{\text{stochastic}}
$$


## A very simple model

Sometime these pieces are referred to as _fixed_ and _random_

$$
y_i = \underbrace{\mu}_{\text{fixed}} + \underbrace{e_i}_{\text{random}}
$$


## A very simple model

This can also be seen by rewriting the model 

$$
y_i = \mu + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$

as

$$
y_i \sim \text{N}(\mu,\sigma^2)
$$


## Simple linear regression

We can expand the deterministic part of the model, as with linear regression

$$
y_i = \underbrace{\alpha + \beta x_i}_{\text{mean}} + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$

so

$$
y_i \sim \text{N}(\alpha + \beta x_i,\sigma^2)
$$


## A simple time series model

Consider a simple model with a mean $\mu$ plus white noise

$$
y_t = \mu + e_t ~ \text{with} ~ e_t \sim \text{N}(0,\sigma^2)
$$


## Time series model with covariates

We can expand the deterministic part of the model, as before with linear regression

$$
y_t = \underbrace{\alpha + \beta x_t}_{\text{mean}} + e_t ~ \text{with} ~ e_t \sim \text{N}(0,\sigma^2)
$$

so

$$
y_t \sim \text{N}(\alpha + \beta x_t,\sigma^2)
$$


## Example of linear model

```{r ts_with_AR_errors}
set.seed(123)
tt <- 30
ee <- rnorm(tt)
ww <- rnorm(tt)
phi <- 0.7
for(t in 2:tt) {
  ee[t] <- phi * ee[t-1] + ww[t]
}

aa <- 5
bb <- 1
xx <- rnorm(tt) + 2

yy <- aa + bb * xx + ee

par(mai = c(0.9, 0.9, 0.3, 0.1), omi = c(0, 0, 0, 0))

plot.ts(yy, ylim = c(min(xx,yy), max(xx, yy)), las = 1,
        col = "dodgerblue", lwd = 2,
        ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(xx, col = "blue", lwd = 2)
text(xx[1], expression(italic(x[t])), pos = 3, col = "blue")
text(yy[1], expression(italic(y[t])), pos = 3, col = "dodgerblue")
mtext(side = 3, expression(italic(y[t]) == 1~+~5~italic(x[t])~+~italic(e[t])),
      line = 0.4, adj = 0)
```


## Model residuals

```{r model_residuals}

par(mai = c(0.9, 0.9, 0.3, 0.1), omi = c(0, 0, 0, 0))

plot.ts(ee, col = "darkred", lwd = 2,
        ylab = expression(italic(e[t])))
```

These do _not_ look like white noise!


## ACF of model residuals

```{r ACF_model_residuals}

par(mai = c(0.9, 0.9, 0.3, 0.1), omi = c(0, 0, 0, 0))

acf(ee)
```

There is significant autocorrelation at lag = 1


## Model with autocorrelated errors

We can expand the stochastic part of the model to have autocorrelated errors

$$
y_t = \alpha + \beta x_t + e_t \\
e_t = \phi e_{t-1} + w_t
$$

with $w_t \sim \text{N}(0,\sigma^2)$


## Model with autocorrelated errors

We can expand the stochastic part of the model to have autocorrelated errors

$$
y_t = \alpha + \beta x_t + e_t \\
e_t = \phi e_{t-1} + w_t
$$

with $w_t \sim \text{N}(0,\sigma^2)$

We can write this model as our standard state-space model


## State-space model | Observation equation

$$
\begin{align}
  y_t &= \alpha + \beta x_t + e_t \\
      &= e_t + \alpha + \beta x_t\\
      &\Downarrow \\
  y_t &= x_t + a + D d_t + v_t\\
\end{align}
$$

with

$x_t = e_t$, $a = \alpha$, $D = \beta$, $d_t = x_t$, $v_t = 0$


## State-space model | State equation


$$
\begin{align}
  e_t &= \phi e_{t-1} + w_t \\
      &\Downarrow \\
  x_t &= B x_t + w_t\\
\end{align}
$$

with

$x_t = e_t$ and $B = \phi$


## State-space model | Full form

$$
y_t = \alpha + \beta x_t + e_t \\
e_t = \phi e_{t-1} + w_t \\
\Downarrow \\
y_t = a + D d_t + x_t\\
x_t = B x_t + w_t
$$


## State-space model | Observation model in `MARSS()`

$$
y_t = a + D d_t + x_t \\
\Downarrow \\
y_t = Z x_t + a + D d_t + v_t
$$

```{r MARSS_DFA_setup, echo = TRUE, eval = FALSE}
y = data         ## [1 x T] matrix of data
a = matrix("a")  ## intercept
D = matrix("D")  ## slope
d = covariate    ## [1 x T] matrix of measured covariate
Z = matrix(1)    ## no multiplier on x 
R = matrix(0)    ## v_t ~ N(0,R); want v_t = 0 for all t
```


## State-space model | State model in `MARSS()`

$$
x_t = B x_t + w_t \\
\Downarrow \\
x_t = B x_t + u + C c_t + w_t
$$

```{r MARSS_AR1_setup, echo = TRUE, eval = FALSE}
B = matrix("b")  ## AR(1) coefficient for model errors
Q = matrix("q")  ## w_t ~ N(0,Q); var for model errors
u = matrix(0)    ## u = 0
C = matrix(0)    ## C = 0
c = matrix(0)    ## c_t = 0 for all t
```


## {.flexbox .vcenter}

<font size="10">MORE RANDOM EFFECTS</font>


## Expanding the random effect

Recall our simple model

$$
y_t = \underbrace{\mu}_{\text{fixed}} + \underbrace{e_t}_{\text{random}}
$$


## Expanding the random effect

We can expand the random portion

$$
y_t = \underbrace{\mu}_{\text{fixed}} + ~ \underbrace{f_t + e_t}_{\text{random}}
$$

$$
e_t \sim \text{N}(0, \sigma) \\
f_t \sim \text{N}(f_{t-1}, \gamma)
$$


## Expanding the random effect

We can expand the random portion

$$
y_t = \underbrace{\mu}_{\text{fixed}} + ~ \underbrace{f_t + e_t}_{\text{random}}
$$

$$
e_t \sim \text{N}(0, \sigma) \\
f_t \sim \text{N}(f_{t-1}, \gamma)
$$

This is simply a random walk observed with error


## Random walk observed with error

$$
y_t = \mu + f_t + e_t ~ \text{with} ~ e_t \sim \text{N}(0, \sigma) \\
f_t = f_{t-1} + w_t ~ \text{with} ~ w_t \sim \text{N}(0, \gamma) \\
\Downarrow \\
y_t = a + x_t + v_t ~ \text{with} ~ v_t \sim \text{N}(0, R) \\
x_t = x_{t-1} + w_t ~ \text{with} ~ w_t \sim \text{N}(0, Q)
$$


## Expanding fixed & random effects

We can expand the fixed portion

$$
y_t = \underbrace{\alpha + \beta x_t}_{\text{fixed}} + ~ \underbrace{f_t + e_t}_{\text{random}}
$$

$$
e_t \sim \text{N}(0, \sigma) \\
f_t \sim \text{N}(f_{t-1}, \gamma)
$$


## Fixed & random effects | In familiar state-space form

$$
y_t = \alpha + \beta x_t + f_t + e_t ~ \text{with} ~ e_t \sim \text{N}(0, \sigma) \\
f_t = f_{t-1} + w_t ~ \text{with} ~ w_t \sim \text{N}(0, \gamma) \\
\Downarrow \\
y_t = a + D d_t + x_t + v_t ~ \text{with} ~ v_t \sim \text{N}(0, R) \\
x_t = x_{t-1} + w_t ~ \text{with} ~ w_t \sim \text{N}(0, Q)
$$

## {.flexbox .vcenter}

<font size="10">MULTIPLE TIME SERIES</font>


## Simple model for 2+ time series | Random walk observed with error

$$
y_{i,t} = x_{i,t} + a_i + v_{i,t} \\
x_{i,t} = x_{i,t-1} + w_{i,t}
$$

with

$v_{i,t} \sim \text{N}(0, R)$

$w_{i,t} \sim \text{N}(0, Q)$


## Random walk observed with error

$$
y_{1,t} = x_{1,t} + a_1 + v_{1,t} \\
y_{2,t} = x_{2,t} + a_2 + v_{2,t} \\
\vdots \\
y_{n,t} = x_{n,t} + a_2 + v_{n,t} \\
$$

$$
x_{1,t} = x_{1,t-1} + w_{1,t} \\
x_{2,t} = x_{2,t-1} + w_{2,t} \\
\vdots \\
x_{n,t} = x_{n,t-1} + w_{n,t}
$$


## Random walk observed with error | In matrix form

$$
\mathbf{y}_t = \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

with 

$\mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R})$

$\mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})$


##

```{r plot_many_ts, echo = FALSE, fig.align = 'center'}
NN <- 25
TT <- 30
MM <- 3
 
## MM x TT matrix of innovations
ww <- matrix(rnorm(MM * TT, 0, 1), MM, TT)
ww[,1] <- rnorm(MM, 0, sqrt(5))
## MM x TT matrix of scaled latent trends
xx <- t(scale(apply(ww, 1, cumsum)))

## loadings matrix
ZZ <- matrix(runif(NN * MM, -1, 1), NN, MM)
diag(ZZ) <- rev(sort(abs(diag(ZZ))))
ZZ[upper.tri(ZZ)] <- 0
ZZ <- round(ZZ, 2)

## obs var
obs_var <- 0.2^2
## obs errors
ee <- t(MASS::mvrnorm(TT, matrix(0, NN, 1), diag(obs_var, NN, NN)))
## NN x TT matrix of observed data
yy <- ZZ %*% xx + ee

clr <- viridis::plasma(NN, alpha = 0.7, end = 0.8)

vv <- sample(seq(NN), NN)

par(mfrow = c(5, 5), mai = c(0.1, 0.1, 0, 0), omi = c(0, 0, 0, 0)) 

for(i in 1:NN) {
	plot.ts(yy[vv[i],], lwd = 2,
	        xlab = "", xaxt = "n", ylab = "", yaxt = "n",
	        col = clr[i], bty = "n")
}
```


## Environmental time series

We often observe covariance among environmental time series, especially for those collected close to one another in space


##

```{r plot_many_ts_2, echo = FALSE, fig.align='center'}
par(mfrow = c(5,5), mai = c(0.1,0.1,0,0), omi = c(0, 0, 0, 0)) 

for(i in 1:NN) {
	plot.ts(yy[vv[i],], lwd = 2,
	        xlab = "", xaxt = "n", ylab = "", yaxt = "n",
	        col = clr[i], bty = "n")
}
```

Are there some common patterns here?


## Common patterns in time series

```{r plot_dfa_trends, fig.align="center"}
## plot the trends
par(mfrow = c(1,3), mai = c(1.2, 0, 0, 0), omi = rep(0.1, 4))
clr <- viridis::plasma(MM, end = 0.8)
for(i in 1:3) {
	plot.ts(xx[i,], lwd = 3,
	        xlab = "", xaxt = "n", ylab = "", yaxt = "n",
	        col = clr[i], bty = "n")
}
```


## State-space model | Ex: population structure

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

We can make (test) assumptions by specifying different forms for $\mathbf{Z}$


## State-space model | Ex: Harbor seal population structure

$$
\begin{bmatrix}
 y_1 \\
 y_2 \\
 y_3 \\
 y_4 \\
 y_5 
\end{bmatrix}_t =
\begin{bmatrix}
 1 & 0 & 0 \\
 0 & 1 & 0 \\
 0 & 1 & 0 \\
 0 & 0 & 1 \\
 0 & 0 & 1 \\
\end{bmatrix} \times
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_t +
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 a_4 \\
 a_5 
\end{bmatrix} +
\begin{bmatrix}
 v_1 \\
 v_2 \\
 v_3 \\
 v_4 \\
 v_5 
\end{bmatrix}_t
$$

$$
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_t =
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_{t-1} +
\begin{bmatrix}
 w_{JF} \\
 w_N \\
 w_S
\end{bmatrix}_t
$$


## Finding common patterns

What if our observations were instead a mixture of 2+ states?

For example, we sampled haul-outs located between several breeding sites


## Mixtures of states

$$
\begin{bmatrix}
 y_1 \\
 y_2 \\
 y_3 \\
 y_4 \\
 y_5 
\end{bmatrix}_t =
\begin{bmatrix}
 0.8 & 0.2 & 0 \\
 0.2 & 0.7 & 0.1 \\
 0 & 0.9 & 0.1 \\
 0 & 0.3 & 0.7 \\
 0 & 0.1 & 0.9 \\
\end{bmatrix} \times
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_t +
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 a_4 \\
 a_5 
\end{bmatrix} +
\begin{bmatrix}
 v_1 \\
 v_2 \\
 v_3 \\
 v_4 \\
 v_5 
\end{bmatrix}_t
$$

$$
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_t =
\begin{bmatrix}
 x_{JF} \\
 x_N \\
 x_S 
\end{bmatrix}_{t-1} +
\begin{bmatrix}
 w_{JF} \\
 w_N \\
 w_S
\end{bmatrix}_t
$$


## Finding common patterns

What if our observations were a mixture of states, but we didn't know how many or the weightings?

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

What are the dimensions of $\mathbf{Z}$?

What are the elements within $\mathbf{Z}$?


## Dynamic Factor Analysis (DFA)

DFA is a _dimension reduction_ technique, which models $n$ observed time series as a function of $m$ hidden states (patterns), where $n \gg m$


## Dynamic Factor Analysis (DFA) | State-space form

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

data: $\mathbf{y}_t$ is $n \times 1$

loadings: $\mathbf{Z}$ is $n \times m$ with $n > m$

states: $\mathbf{x}_t$ is $m \times 1$


## Dimension reduction | Principal Components Analysis (PCA)

Goal is to reduce some large number of correlated variates into a few uncorrelated factors


## Principal Components Analysis (PCA)

Calculating the principal components requires us to estimate the covariance of the data

$$
\text{PC} = \text{eigenvectors}(\text{cov}(\mathbf{y}))
$$

There will be $n$ principal components (eigenvectors) for an $n \times T$ matrix $\mathbf{y}$

We reduce the dimension by selecting a subset of the components that explain much of the variance (eg, the first 2)


## Principal Components Analysis (PCA)

```{r ex_corr_ts}
nn <- 100
xx <- rnorm(nn, 10, 1)
yy <- xx + rnorm(nn, 5)
xx <- xx - mean(xx)
yy <- yy - mean(yy)

par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 0, 0, 0))

plot(xx, yy, pch = 16, las = 1,
     ylab = "Y", xlab = "X")

text(-2, 2,
     substitute(rho == x, list(x = round(cor(cbind(xx, yy))[2,1], 2))))
```


## Principal Components Analysis (PCA)

```{r ex_PCA}
pca <- prcomp(cbind(xx, yy))

pc1 <- pca$rotation[,1]
pc2 <- matrix(c(0, -1, 1, 0), 2, 2) %*% matrix(pc1, 2, 1)

par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 0, 0, 0))

plot(xx, yy, pch = 16, las = 1,
     ylab ="Y", xlab = "X")
arrows(0, 0, pc1[1], pc1[2],
       col = "blue", lwd = 3, length = 0.1)
text(pc1[1], pc1[2], "PC1",
     pos = 3, col = "blue")
arrows(0, 0, pc2[1], pc2[2],
       col = "darkred", lwd = 3, length = 0.1)
text(pc2[1], pc2[2], "PC2",
     pos = 1, col = "darkred")
```


## Principal Components Analysis (PCA)

```{r ex_PCA_rotated}
par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 0, 0, 0))

plot(pca$x[,1], pca$x[,2], pch = 16, las = 1,
     ylab = "PC2", xlab = "PC1")

text(-4, 1.2,
     substitute(rho == x, list(x = round(cor(pca$x)[2,1], 2))))
```


## Relationship between PCA & DFA

We need to estimate the covariance in the data $\mathbf{y}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t, ~ \text{with} ~ \mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R})
$$

so

$$
\text{cov}(\mathbf{y}_t) = \mathbf{Z} \text{cov}(\mathbf{x}_t) \mathbf{Z}^\top + \mathbf{R}
$$
In PCA, we require $\mathbf{R}$ to be diagonal, but not so in DFA


## Principal Components Analysis (PCA) | Forms for $\mathbf{R}$ with $n = 4$

$$
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma & 0 & 0 & 0 \\
 0 & \sigma & 0 & 0 \\
 0 & 0 & \sigma & 0 \\
 0 & 0 & 0 & \sigma
\end{bmatrix}
~\text{or}~~
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma_1 & 0 & 0 & 0 \\
 0 & \sigma_2 & 0 & 0 \\
 0 & 0 & \sigma_3 & 0 \\
 0 & 0 & 0 & \sigma_4
\end{bmatrix}
$$


## Dynamic Factor Analysis (DFA) | Forms for $\mathbf{R}$ with $n = 4$

$$
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma & 0 & 0 & 0 \\
 0 & \sigma & 0 & 0 \\
 0 & 0 & \sigma & 0 \\
 0 & 0 & 0 & \sigma
\end{bmatrix}
~\text{or}~~
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma_1 & 0 & 0 & 0 \\
 0 & \sigma_2 & 0 & 0 \\
 0 & 0 & \sigma_3 & 0 \\
 0 & 0 & 0 & \sigma_4
\end{bmatrix}
$$

$$
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma & \gamma & \gamma & \gamma \\
 \gamma & \sigma & \gamma & \gamma \\
 \gamma & \gamma & \sigma & \gamma \\
 \gamma & \gamma & \gamma & \sigma
\end{bmatrix}
~\text{or}~~
\mathbf{R} \stackrel{?}{=}
\begin{bmatrix}
 \sigma_1 & 0 & 0 & 0 \\
 0 & \sigma_2 & 0 & \gamma_{2,4} \\
 0 & 0 & \sigma_3 & 0 \\
 0 & \gamma_{2,4} & 0 & \sigma_4
\end{bmatrix}
$$


## Dynamic Factor Analysis (DFA)

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

What form should we use for $\mathbf{Z}$?

$$
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_1 \\
 z_2 \\
 z_3 \\
 \vdots \\
 z_5
\end{bmatrix}
~\text{or}~~
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_{1,1} & z_{2,1} \\
 z_{1,2} & z_{2,2} \\
 z_{1,3} & z_{2,3} \\
 \vdots & \vdots \\
 z_{1,n} & z_{2,n}
\end{bmatrix}
~\text{or}~~
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_{1,1} & z_{2,1} & z_{3,1} \\
 z_{1,2} & z_{2,2} & z_{3,2} \\
 z_{1,3} & z_{2,3} & z_{3,3} \\
 \vdots & \vdots & \vdots \\
 z_{1,n} & z_{2,n} & z_{3,n}
\end{bmatrix}
$$


## Dynamic Factor Analysis (DFA)

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

What form should we use for $\mathbf{Z}$?

$$
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_1 \\
 z_2 \\
 z_3 \\
 \vdots \\
 z_5
\end{bmatrix}
~\text{or}~~
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_{1,1} & z_{2,1} \\
 z_{1,2} & z_{2,2} \\
 z_{1,3} & z_{2,3} \\
 \vdots & \vdots \\
 z_{1,n} & z_{2,n}
\end{bmatrix}
~\text{or}~~
\mathbf{Z} \stackrel{?}{=}
\begin{bmatrix}
 z_{1,1} & z_{2,1} & z_{3,1} \\
 z_{1,2} & z_{2,2} & z_{3,2} \\
 z_{1,3} & z_{2,3} & z_{3,3} \\
 \vdots & \vdots & \vdots \\
 z_{1,n} & z_{2,n} & z_{3,n}
\end{bmatrix}
$$

We'll use model selection criteria to choose (eg, AICc)


## Fitting DFA models

Unless $\mathbf{Z}$ is unconstrained in some manner, there are an infinite number of combinations of $\mathbf{Z}$ and $\mathbf{x}$ that will equal $\mathbf{y}$

Therefore we need to impose some constraints on the model


## Constraints on DFA models | 1) The offset $\mathbf{a}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{a} =
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 \vdots \\
 a_n
\end{bmatrix}
$$


## Constraints on DFA models | 1) The offset $\mathbf{a}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{a} =
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 \vdots \\
 a_n
\end{bmatrix}
$$

We will set the first $m$ elements of $\mathbf{a}$ to 0


## Constraints on DFA models | 1) The offset $\mathbf{a}$

For example, if $n = 5$ and $m = 2$

$$
\mathbf{a} =
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 a_4 \\
 a_5
\end{bmatrix}
\Rightarrow
\mathbf{a} =
\begin{bmatrix}
 0 \\
 0 \\
 a_3 \\
 a_4 \\
 a_5
\end{bmatrix}
$$


## Constraints on DFA models | 1) The offset $\mathbf{a}$

For example, if $n = 5$ and $m = 2$

$$
\mathbf{a} =
\begin{bmatrix}
 a_1 \\
 a_2 \\
 a_3 \\
 a_4 \\
 a_5
\end{bmatrix}
\Rightarrow
\mathbf{a} =
\begin{bmatrix}
 0 \\
 0 \\
 a_3 \\
 a_4 \\
 a_5
\end{bmatrix}
\Rightarrow
\mathbf{a} =
\begin{bmatrix}
 0 \\
 0 \\
 0 \\
 0 \\
 0
\end{bmatrix}
$$

Note, however, that this causes problems for the EM algorithm so we will often de-mean the data and set $a_i = 0$ for all $i$


## Constraints on DFA models | 2) The loadings $\mathbf{Z}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{Z} =
\begin{bmatrix}
 z_{1,1} & z_{2,1} & \dots & z_{m,1} \\
 z_{1,2} & z_{2,2} & \dots & z_{m,2} \\
 z_{1,3} & z_{2,3} & \dots & z_{m,3} \\
 \vdots & \vdots & \ddots & z_{m,4} \\
 z_{1,n} & z_{2,n} & \dots & z_{m,n}
\end{bmatrix}
$$


## Constraints on DFA models | 2) The loadings $\mathbf{Z}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{Z} =
\begin{bmatrix}
 z_{1,1} & z_{2,1} & \dots & z_{m,1} \\
 z_{1,2} & z_{2,2} & \dots & z_{m,2} \\
 z_{1,3} & z_{2,3} & \dots & z_{m,3} \\
 \vdots & \vdots & \ddots & z_{m,4} \\
 z_{1,n} & z_{2,n} & \dots & z_{m,n}
\end{bmatrix}
$$

We will set the upper right triangle of $\mathbf{Z}$ to 0


## Constraints on DFA models | 2) The loadings $\mathbf{Z}$

For example, if $n = 5$ and $m = 3$

$$
\mathbf{Z} =
\begin{bmatrix}
 z_{1,1} & 0 & 0 \\
 z_{1,2} & z_{2,2} & 0 \\
 z_{1,3} & z_{2,3} & z_{3,3} \\
 z_{1,4} & z_{2,3} & z_{3,4} \\
 z_{1,5} & z_{2,5} & z_{3,5}
\end{bmatrix}
$$

For the first $m - 1$ rows of $\mathbf{Z}$, $z_{i,j} = 0$ if $j > i$


## Constraints on DFA models | 2) The loadings $\mathbf{Z}$

An additional constraint is necessary in a Bayesian context

$$
\mathbf{Z} =
\begin{bmatrix}
 \underline{z_{1,1}} & 0 & 0 \\
 z_{1,2} & \underline{z_{2,2}} & 0 \\
 z_{1,3} & z_{2,3} & \underline{z_{3,3}} \\
 z_{1,4} & z_{2,3} & z_{3,4} \\
 z_{1,5} & z_{2,5} & z_{3,5}
\end{bmatrix}
$$

Diagonal of $\mathbf{Z}$ is positive: $z_{i,j} > 0$ if $i = j$


## Constraints on DFA models | 3) The state variance $\mathbf{Q}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
$$


## Constraints on DFA models | 3) The state variance $\mathbf{Q}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})
$$

We will set $\mathbf{Q}$ equal to the Identity matrix $\mathbf{I}$


## Constraints on DFA models | 3) The state variance $\mathbf{Q}$

For example, if $m = 4$

$$
\mathbf{Q} =
\begin{bmatrix}
 1 & 0 & 0 & 0 \\
 0 & 1 & 0 & 0 \\
 0 & 0 & 1 & 0 \\
 0 & 0 & 0 & 1
\end{bmatrix}
$$

This allows our random walks to have a _lot_ of flexibility


## Dynamic Factor Analysis (DFA) | Including $p$ covariates

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \underline{\mathbf{D} \mathbf{d}_t} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$\mathbf{d}_t$ is a $p \times 1$ vector of covariates at time $t$

$\mathbf{D}$ is an $n \times p$ matrix of covariate effects


## Dynamic Factor Analysis (DFA) | Form for $\mathbf{D}$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \underline{\mathbf{D}} \mathbf{d}_t + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

Careful thought must be given _a priori_ as to the form for $\mathbf{D}$

Should the effect(s) vary by site, species, etc?


## Dynamic Factor Analysis (DFA) | Form for $\mathbf{D}$

For example, given 2 covariates, $\text{Temp}$ and $\text{Salinity}$

$$
\mathbf{D} =
\underbrace{
\begin{bmatrix}
 d_{\text{Temp}} & d_{\text{Salinity}} \\
 d_{\text{Temp}} & d_{\text{Salinity}} \\
 \vdots & \vdots \\
 d_{\text{Temp}} & d_{\text{Salinity}} \\
\end{bmatrix}
}_{\text{effects same by site/species}}
~~~ \text{or} ~~~
\mathbf{D} =
\underbrace{
\begin{bmatrix}
 d_{\text{Temp}, 1} & d_{\text{Salinity}, 1} \\
 d_{\text{Temp}, 2} & d_{\text{Salinity}, 2} \\
 \vdots & \vdots \\
 d_{\text{Temp}, n} & d_{\text{Salinity}, n} \\
\end{bmatrix}
}_{\text{effects differ by site/species}}
$$


## A note on model selection

Earlier we saw that we could use model selection criteria to help us choose among the different forms for $\mathbf{Z}$

However, caution must be given when comparing models with and without covariates, and varying numbers of states


## A note on model selection

Think about the DFA model form

$$
\mathbf{y}_t = \mathbf{Z} \underline{\mathbf{x}_t} + \mathbf{a} + \mathbf{D} \underline{\mathbf{d}_t} + \mathbf{v}_t \\
$$

$\mathbf{x}_t$ are _undetermined_ random walks

$\mathbf{d}_t$ are _predetermined_ covariates


## An example with 3 times series

Model 1 has 2 trends and no covariates

$$
\begin{bmatrix}
 y_1 \\
 y_2 \\
 y_3 
\end{bmatrix}_t =
\begin{bmatrix}
 z_{1,1} & z_{2,1} \\
 z_{1,2} & z_{2,2} \\
 z_{1,3} & z_{2,3}
\end{bmatrix}
\begin{bmatrix}
 x_1 \\
 x_2 
\end{bmatrix}_t +
\begin{bmatrix}
 v_1 \\
 v_2 \\
 v_3 
\end{bmatrix}_t
$$

Model 2 has 1 trend and 1 covariate

$$
\begin{bmatrix}
 y_1 \\
 y_2 \\
 y_3 
\end{bmatrix}_t =
\begin{bmatrix}
 z_1 \\
 z_2 \\
 z_3 
\end{bmatrix}
\begin{bmatrix}
 x
\end{bmatrix}_t +
\begin{bmatrix}
 D_1 \\
 D_2 \\
 D_3 
\end{bmatrix}
\begin{bmatrix}
 d
\end{bmatrix}_t +
\begin{bmatrix}
 v_1 \\
 v_2 \\
 v_3 
\end{bmatrix}_t
$$


## An example with 3 times series

Model 1 has 2 trends and no covariates

$$
\begin{bmatrix}
 y_1 \\
 y_2 \\
 y_3 
\end{bmatrix}_t =
\begin{bmatrix}
 z_{1,1} & z_{2,1} \\
 z_{1,2} & z_{2,2} \\
 z_{1,3} & z_{2,3}
\end{bmatrix}
\begin{bmatrix}
 x_1 \\
 x_2 
\end{bmatrix}_t +
\begin{bmatrix}
 v_1 \\
 v_2 \\
 v_3 
\end{bmatrix}_t
$$

Model 2 has 1 trend and 1 covariate

$$
\begin{bmatrix}
 y_1 \\
 y_2 \\
 y_3 
\end{bmatrix}_t =
\begin{bmatrix}
 z_1 \\
 z_2 \\
 z_3 
\end{bmatrix}
\begin{bmatrix}
 x
\end{bmatrix}_t +
\begin{bmatrix}
 D_1 \\
 D_2 \\
 D_3 
\end{bmatrix}
\begin{bmatrix}
 d
\end{bmatrix}_t +
\begin{bmatrix}
 v_1 \\
 v_2 \\
 v_3 
\end{bmatrix}_t
$$

Unless $\mathbf{d}$ is *highly correlated* with $\mathbf{y}$, Model 1 will be favored


## A note on model selection | For models with covariates

* fit the *most complex model you can envision* based on all of your possible covariates and random factors (states)

* keep the covariates fixed and choose the number of trends (states) using AICc

* keep the covariates & states fixed and choose the form for $\mathbf{R}$

* sort out the covariates while keeping the states & $\mathbf{R}$ fixed


## Interpreting DFA results

Recall that we had to constrain the form of $\mathbf{Z}$ to fit the model

$$
\mathbf{Z} =
\begin{bmatrix}
 z_{1,1} & 0 & \dots & 0 \\
 z_{1,2} & z_{2,2} & \ddots & 0 \\
 \vdots & \vdots & \ddots & 0 \\
 \vdots & \vdots & \vdots & z_{m,m}  \\
 \vdots & \vdots & \vdots & \vdots \\
 z_{1,n} & z_{2,n} & z_{3,n} & z_{m,n} 
\end{bmatrix}
$$

So, the 1st common factor is determined by the 1st variate, the 2nd common factor by the first two variates, etc.


## Interpreting DFA results

To help with this, we can use a _basis rotation_ to maximize the loadings on a few factors

If $\mathbf{H}$ is an $m \times m$ non-singular matrix, these 2 DFA models are equivalent

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{D} \mathbf{d}_t + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

$$
\mathbf{y}_t = \mathbf{Z} \mathbf{H}^{-1} \mathbf{x}_t + \mathbf{a} + \mathbf{D} \mathbf{d}_t + \mathbf{v}_t \\
\mathbf{H} \mathbf{x}_t = \mathbf{H} \mathbf{x}_{t-1} + \mathbf{H} \mathbf{w}_t
$$

How should we choose $\mathbf{H}$?


## Basis rotation | Varimax

A _varimax_ rotation will maximize the variance of the loadings in $\mathbf{Z}$ along a few of the factors


## PCA of 5 wines with 8 attributes

```{r ex_PCA_wines}
dat <- read.csv("wine_data_for_PCA.csv")
colnames(dat)[2:3] <- c("For meat", "For dessert")

pca <- prcomp(dat, scale. = TRUE)

loads <- pca$rotation
if(loads[1,2] < 0) { loads[,2] <- -1 * loads[,2] }

load1 <- loads[,1]
load2 <- loads[,2]

par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 1.5, 0, 1.5))

poz <- c(3, 1, 3, 3, 3, 4, 2)
  
plot(load1, load2, pch = 16,
     xlim = c(min(load1), max(load1)),
     ylim = c(min(load2), max(load2)),
     xaxt = "n", yaxt = "n", bty = "n",
     xlab = "", ylab = "")
abline(h = 0, col = "darkgray")
abline(v = 0, col = "darkgray")
for(i in 1:ncol(dat)) {
  text(load1[i], load2[i], colnames(dat)[i],
       pos = poz[i], xpd = NA)
}
text(0, min(load2) * 1.05, "X2", pos = 1, xpd = NA, col = "darkgray")
text(min(load1) * 1.05, 0, "X1", pos = 2, xpd = NA, col = "darkgray")
```


## Rotated loadings

```{r ex_PCA_wines_rotated}
theta <- 165 * pi / 180
basis_rot <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2)
  
new_x <- t(basis_rot %*% matrix(c(1, 0, -1, 0), 2, 2))
new_y <- t(basis_rot %*% matrix(c(0, 1, 0, -1), 2, 2))

par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 1.5, 0, 1.5))

plot(load1, load2, pch = 16,
     xlim = c(min(load1), max(load1)),
     ylim = c(min(load2), max(load2)),
     xaxt = "n", yaxt = "n", bty = "n",
     xlab = "", ylab = "")
abline(h = 0, col = "darkgray")
abline(v = 0, col = "darkgray")
## new axes
lines(new_x[,1], new_x[,2], col = "blue")
lines(new_y[,1], new_y[,2], col = "blue")
## labels
for(i in 1:ncol(dat)) {
  text(load1[i], load2[i], colnames(dat)[i],
       pos = poz[i], xpd = NA)
}
text(0, min(load2) * 1.05, "X2", pos = 1, xpd = NA, col = "darkgray")
text(min(load1) * 1.05, 0, "X1", pos = 2, xpd = NA, col = "darkgray")

theta <- 345 * pi / 180
basis_rot <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2)
text((basis_rot %*% matrix(c(0,par()$usr[3]),2,1))[1], min(load2) * 1.05,
     "R2", pos = 1, xpd = NA, col = "blue")
text(min(load1) * 1.05, (basis_rot %*% matrix(c(par()$usr[1],0),2,1))[2],
     "R1", pos = 2, xpd = NA, col = "blue")

```



## Rotated loadings

```{r ex_PCA_loadings}
theta <- 15 * pi / 180
basis_rot <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2)
  
rot_loads <- t(basis_rot %*% t(loads[,1:2]))
rload1 <- rot_loads[,1]
rload2 <- rot_loads[,2]

  
par(mai = c(0.9, 0.9, 0.1, 0.1), omi = c(0, 1.5, 0, 1.5))

plot(load1, load2, pch = 16, col = "darkgray",
     xlim = c(min(load1,rload1), max(load1,rload1)),
     ylim = c(min(load2,rload2), max(load2,rload2)),
     xaxt = "n", yaxt = "n", bty = "n",
     xlab = "", ylab = "")
abline(h = 0, col = "darkgray")
abline(v = 0, col = "darkgray")
for(i in 1:ncol(dat)) {
  text(rload1[i], rload2[i], colnames(dat)[i],
       pos = poz[i], xpd = NA)
}
text(0, min(load2) * 1.05, "R2", pos = 1, xpd = NA, col = "darkgray")
text(min(load1) * 1.05, 0, "R1", pos = 2, xpd = NA, col = "darkgray")

points(rload1, rload2, pch = 16)
```


## Topics for today

Deterministic vs stochastic elements

Regression with autocorrelated errors

Regression with temporal random effects

Dynamic Factor Analysis (DFA)

* Forms of covariance matrix  

* Constraints for model fitting

* Interpretation of results

