---
title: "Hidden Markov Models and Bayesian MARSS models"
author: "Eli Holmes, Eric Ward"
date: "17 February 2021"
output:
  html_document:
    theme: cosmo
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r hw6p1-setup, echo=FALSE}
knitr::opts_knit$set(unnamed.chunk.label = "hw6-",
                     message=FALSE,
                     warning=FALSE)
```

<br>

# Part 1. Hidden Markov Models {#sec-hmm}

**Identifying regimes using Hidden Markov Models (HMMs)**

For the first part of the homework, we'll use data from the Pacific Decadal Oscillation (PDO) to ask questions about identifying regimes. This dataset can be accessed via the `rsoi` package. First, let's grab the data. Run the `install.packages()` code if you need the `rsoi` package.

```{r read-data, message=FALSE, warning=FALSE}
library(dplyr)
#install.packages("rsoi")
pdo <- rsoi::download_pdo()
```
We will look at the winter PDO only. We need to shift the year for Oct-Dec by 1 since Oct-Feb spans 2 calendar years.
```{r read-data2}
pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] <- pdo$Year[which(pdo$Month%in%c("Oct","Nov","Dec"))] + 1
pdo <- dplyr::group_by(pdo, Year) %>%
  dplyr::summarize(winter_pdo = mean(PDO[which(Month %in% c("Oct","Nov","Dec","Jan","Feb"))])) %>% 
  dplyr::select(winter_pdo, Year)
# The first year will be missing Oct-Dec
pdo <- pdo[-1,]
```

Use `pdo` for your analyses. You will be modeling `winter_pdo`. Use the `depmixS4` package discussed in the [HMM lecture](https://atsa-es.github.io/atsa/Lectures/Week%205/lec_10_hmm.html#1) to answer the questions in in Part 1.


1. Fit a 2-state HMM to the annual indices of winter PDO. Assume Gaussian errors (default). See the lecture on HMMs and/or section 3 in the [depmixS4 vignette](https://cran.r-project.org/web/packages/depmixS4/vignettes/depmixS4.pdf)
    
2. Try fitting the model 10-20 times. Does the likelihood seem reasonably stable? (Note `logLik()` gets you the log-likelihood from model fits in R).
    
3. What is the transition matrix for the best model? What are the persistence probabilities (e.g. probabilities of staying in the same state from step $t$ to $t+1$)?
    
4. Using slide 15 of the HMM lecture as a template and the matrix in Q3, write out your fitted HMM model as equations.

5. Plot the predicted values versus year. See slide 50 of the HMM lecture for an example.

6. Plot the posterior probability of being in the various states from your best model (e.g. probability of being in state 1 over time)   

7. What is the long-run probability that the PDO is in state 1 versus state 2? You can calculate this from the transition matrix. There is an analytical solution for this (a bit of googling will find it). Or you can run a `for` loop to find it. Let $p_1$ be the probability that PDO is in state 1 and $p_2$ be the probability that PDO is in state 2. Note $p_1 + p_2 = 1$. If $P$ is the transition matrix (in Q3),

$$\begin{bmatrix}p_1&p_2\end{bmatrix}_n = \begin{bmatrix}p_1&p_2\end{bmatrix}_{n-1} P$$
Note this is a 1x2 matrix times a 2x2 matrix on the right. Start with $p_1=1$ and $p_2=0$, say. Run a `for` loop until 

$$\begin{bmatrix}p_1&p_2\end{bmatrix}_n \approx \begin{bmatrix}p_1&p_2\end{bmatrix}_{n-1}$$
That $\begin{bmatrix}p_1&p_2\end{bmatrix}_n$ is the long-run probability in each state.

**Some ideas for optional extra analyses**

* Change the model to a 3-state model. Using AIC as a model selection metric, does the 3-state model perform better (lower AIC) compared to the 2-state model? What about a 1-state model?

* If you include time varying parameters (e.g. year) in the means of each state, or state transition probabilities, does the model do any better?
 
* If you include time varying parameters (e.g. year) in the means of each state, or state transition probabilities, does the model do any better?

* Run diagnostics on the best model. Any problems?

* Compare the transition matrices for fits with different random starting conditions. Are the transition matrices stable?

