---
title: "Intro to ARMA models"
subtitle: "FISH 507 – Applied Time Series Analysis"
author: "Mark Scheuerell"
date: "15 Jan 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(123)
```

## Topics for today

Review

* White noise  

* Random walks  

Autoregressive (AR) models

Moving average (MA) models

Autoregressive moving average (ARMA) models

Using ACF & PACF for model ID


## White noise (WN)

A time series $\{w_t\}$ is discrete white noise if its values are

1. independent  

2. identically distributed with a mean of zero


## White noise (WN)

A time series $\{w_t\}$ is discrete white noise if its values are

1. independent  

2. identically distributed with a mean of zero

Note that distributional form for $\{w_t\}$ is flexible


## White noise (WN)

```{r, fig.cap="$w_t = 2e_t - 1; e_t \\sim \\text{Bernoulli}(0.5)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
tt <- rbinom(100, 1, 0.5) * 2 - 1
plot.ts(tt, ylab = expression(italic(w[t])))
acf(tt)
```


## Gaussian white noise

We often assume so-called _Gaussian white noise_, whereby

$$
w_t \sim \text{N}(0,\sigma^2)
$$


## Gaussian white noise

We often assume so-called _Gaussian white noise_, whereby

$$
w_t \sim \text{N}(0,\sigma^2)
$$

and the following apply as well

&nbsp; &nbsp; autocovariance:&nbsp; $\gamma_k =
    \begin{cases}
      \sigma^2 & \text{if } k = 0 \\
      0 & \text{if } k \geq 1
    \end{cases}$

&nbsp; &nbsp; autocorrelation: &nbsp; $\rho_k =
    \begin{cases}
      1 & \text{if } k = 0 \\
      0 & \text{if } k \geq 1
    \end{cases}$


## Gaussian white noise

```{r ex_gaussian_wn, fig.cap="$w_t \\sim \\text{N}(0,1)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
tt <- rnorm(100)
plot.ts(tt, ylab = expression(italic(w[t])))
acf(tt)
```


## Random walk (RW)

A time series $\{x_t\}$ is a random walk if

1. $x_t = x_{t-1} + w_t$  

2. $w_t$ is white noise


## Random walk (RW)

The following apply to random walks

&nbsp; &nbsp; mean: &nbsp; $\mu_x = 0$

&nbsp; &nbsp; autocovariance: &nbsp; $\gamma_k(t) = t \sigma^2$

&nbsp; &nbsp; autocorrelation: &nbsp; $\rho_k(t) = \frac{t \sigma^2}{\sqrt{t \sigma^2(t + k) \sigma^2}}$


## Random walk (RW)

The following apply to random walks

&nbsp; &nbsp; mean: &nbsp; $\mu_x = 0$

&nbsp; &nbsp; autocovariance: &nbsp; $\gamma_k(t) = t \sigma^2$

&nbsp; &nbsp; autocorrelation: &nbsp; $\rho_k(t) = \frac{t \sigma^2}{\sqrt{t \sigma^2(t + k) \sigma^2}}$

_Note_: Random walks are not stationary


## Random walk (RW)

```{r ex_rw, fig.cap="$x_t = x_{t-1} + w_t; w_t \\sim \\text{N}(0,1)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
tt <- cumsum(rnorm(100))
plot.ts(tt, ylab = expression(italic(x[t])))
acf(tt)
```


## Biased random walk

A _biased random walk_ (or _random walk with drift_) is written as

$$
x_t = x_{t-1} + u + w_t
$$  

where $u$ is the bias (drift) per time step and $w_t$ is white noise


## Biased random walk

```{r ex_biased_rw, fig.cap="$x_t = x_{t-1} + 1 + w_t; w_t \\sim \\text{N}(0,1)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
xx <- ww <- rnorm(100)
uu <- 1
for(t in 2:100) {
  xx[t] <- xx[t-1] + uu + ww[t]
}
plot.ts(xx, ylab = expression(italic(x[t])))
acf(tt)
```


## Differencing a biased random walk

First-differencing a biased random walk yields white noise with a non-zero mean $u$

$$
\begin{align}
  \nabla x_t &= x_{t-1} + u + w_t \\
  x_t - x_{t-1} &= x_{t-1} + u + w_t - x_{t-1} \\
  x_t - x_{t-1} &= u + w_t
\end{align}
$$


## Differencing a biased random walk

```{r ex_diff_biased_rw, fig.cap="$x_t = x_{t-1} + 1 + w_t; w_t \\sim \\text{N}(0,1)$"}
par(mfrow = c(1,2), mai = c(1.5,0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(diff(xx), ylab = expression(nabla~italic(x[t])))
acf(tt)
```


## {.flexbox .vcenter}

<font size="10">LINEAR STATIONARY MODELS</font>


## Linear stationary models

We saw last week that linear filters are a useful way of modeling time series

Here we extend those ideas to a general class of models call _autoregressive moving average_ (ARMA) models


## Autoregressive (AR) models

An _autoregressive_ model of order _p_, or AR(_p_), is defined as

$$
x_t = \phi_1 x_{t-1} + \phi_2 x_{t-2} + \dots + \phi_p x_{t-p} + w_t
$$

where we assume

1. $w_t$ is white noise

2. $\phi_p \neq 0$ for an order-_p_ process


## Examples of AR(_p_) models

AR(1)

$x_t = 0.5 x_{t-1} + w_t$

<br>
AR(1) with $\phi_1 = 1$ (random walk)

$x_t = x_{t-1} + w_t$

<br>
AR(2)

$x_t = -0.2 x_{t-1} + 0.4 x_{t-2} + w_t$


## Examples of AR(_p_) models

```{r}
set.seed(123)
## the 4 AR coefficients
ARp <- c(0.7, 0.2, -0.1, -0.3)
## empty list for storing models
AR_mods <- vector("list", 4L)

par(mfrow = c(2,2), mai = c(0.7,0.7,0.3,0.1), omi = c(0,0,0,0))
## loop over orders of p
for(p in 1:4) {
  ## assume SD=1, so not specified
  AR_mods[[p]] <- arima.sim(n=50, list(ar=ARp[1:p]))
  plot.ts(AR_mods[[p]], las = 1,
          ylab = expression(italic(x[t])))
  mtext(side = 3, paste0("AR(",p,")"),
        line = 0.5, adj = 0)
}
```


## Stationary AR(_p_) models

Recall that _stationary_ processes have the following properties

1. no systematic change in the mean or variance  
2. no systematic trend  
3. no periodic variations or seasonality

We seek a means for identifying whether our AR(_p_) models are also stationary


## Stationary AR(_p_) models

We can write out an AR(_p_) model using the backshift operator $(\mathbf{B})$, such that


$$
  x_t = \phi_1 x_{t-1} + \phi_2 x_{t-2} + \dots + \phi_p x_{t-p} + w_t \\
  \Downarrow \\
\begin{align}
  x_t - \phi_1 x_{t-1} - \phi_2 x_{t-2} - \dots - \phi_p x_{t-p} &= w_t \\
  (1 - \phi_1 \mathbf{B} - \phi_2 \mathbf{B}^2 - \dots - \phi_p \mathbf{B}^p) x_t &= w_t \\
  \phi_p (\mathbf{B}) x_t &= w_t \\
\end{align}
$$

## Stationary AR(_p_) models

If we treat $\mathbf{B}$ as a number (or numbers), we can out write the _characteristic equation_ as

$$
\phi_p (\mathbf{B}) x_t = w_t \\
\Downarrow \\
\phi_p (\mathbf{B}) = 0
$$

To be stationary, __all roots__ of the characteristic equation __must exceed 1 in absolute value__


## Stationary AR(_p_) models

For example, consider this AR(1) model from earlier

$$
\begin{align}
  x_t &= 0.5 x_{t-1} + w_t \\
  x_t - 0.5 x_{t-1} &= w_t \\
  (1 - 0.5 \mathbf{B})x_t &= w_t \\
\end{align}
$$


## Stationary AR(_p_) models

For example, consider this AR(1) model from earlier

$$
\begin{align}
  x_t &= 0.5 x_{t-1} + w_t \\
  x_t - 0.5 x_{t-1} &= w_t \\
  (1 - 0.5 \mathbf{B})x_t &= w_t \\
  \Downarrow \\
  1 - 0.5 \mathbf{B} &= 0 \\
  -0.5 \mathbf{B} &= -1 \\
  \mathbf{B} &= 2 \\
\end{align}
$$

This model is indeed stationary because $\mathbf{B} > 1$


## Stationary AR(_p_) models

What about this AR(2) model from earlier?

$$
\begin{align}
  x_t &= -0.2 x_{t-1} + 0.4 x_{t-2} + w_t \\
  x_t + 0.2 x_{t-1} - 0.4 x_{t-2} &= w_t \\
  (1 + 0.2 \mathbf{B} - 0.4 \mathbf{B}^2)x_t &= w_t \\
\end{align}
$$


## Stationary AR(_p_) models

What about this AR(2) model from earlier?

$$
\begin{align}
  x_t &= -0.2 x_{t-1} + 0.4 x_{t-2} + w_t \\
  x_t + 0.2 x_{t-1} - 0.4 x_{t-2} &= w_t \\
  (1 + 0.2 \mathbf{B} - 0.4 \mathbf{B}^2)x_t &= w_t \\
  \Downarrow \\
  1 + 0.2 \mathbf{B} - 0.4 \mathbf{B}^2 &= 0 \\
  \Downarrow \\
  \mathbf{B} \approx -1.35 ~ \text{and}& ~ \mathbf{B} \approx 1.85
\end{align}
$$

This model is _not_ stationary because only one $\mathbf{B} > 1$


## What about random walks?

Consider our random walk model

$$
\begin{align}
  x_t &= x_{t-1} + w_t \\
  x_t - x_{t-1} &= w_t \\
  (1 - 1 \mathbf{B})x_t &= w_t \\
\end{align}
$$


## What about random walks?

Consider our random walk model

$$
\begin{align}
  x_t &= x_{t-1} + w_t \\
  x_t - x_{t-1} &= w_t \\
  (1 - 1 \mathbf{B})x_t &= w_t \\
  \Downarrow \\
  1 - 1 \mathbf{B} &= 0 \\
  -1 \mathbf{B} &= -1 \\
  \mathbf{B} &= 1 \\
\end{align}
$$

Random walks are __not__ stationary because $\mathbf{B} = 1 \ngtr 1$


## Stationary AR(1) models

We can define a space over which all AR(1) model are stationary

For $x_t = \phi x_{t-1} + w_t$, we have

$$
\begin{align}
  1 - \phi \mathbf{B} &= 0 \\
  -\phi \mathbf{B} &= -1 \\
  \mathbf{B} &= \frac{1}{\phi} > 1 \Rightarrow 0 < \phi < 1\\
\end{align}
$$


## Stationary AR(1) models

And for $x_t = -\phi x_{t-1} + w_t$, we have

$$
\begin{align}
  1 + \phi \mathbf{B} &= 0 \\
  \phi \mathbf{B} &= -1 \\
  \mathbf{B} &= \frac{-1}{\phi} > 1 \Rightarrow -1 < \phi < 0\\
\end{align}
$$
<br>
Thus, AR(1) models are stationary only if $\lvert \phi \rvert  < 1$  


## Coefficients of AR(1) models

```{r ar_comp_pos_neg, fig.height=4}
set.seed(123)
## list description for AR(1) model with small coef
AR_pos <- list(order=c(1,0,0), ar=0.7, sd=0.1)
## list description for AR(1) model with large coef
AR_neg <- list(order=c(1,0,0), ar=-0.7, sd=0.1)
## simulate AR(1)
AR1_pos <- arima.sim(n=500, model=AR_pos)
AR1_neg <- arima.sim(n=500, model=AR_neg)

## get y-limits for common plots
ylm1 <- c(min(AR1_pos[1:50],AR1_neg[1:50]), max(AR1_pos[1:50],AR1_neg[1:50]))

## set the margins & text size
par(mfrow=c(1,2), mai=c(0.8,0.8,0.3,0.2), oma=c(0,0,0,0))
## plot the ts
plot.ts(AR1_pos[1:50], ylim=ylm1, las = 1,
        ylab=expression(italic(x)[italic(t)]),
        main = "")
mtext(side = 3, expression(paste(phi[1]," = 0.7")),
      line = 0.4, adj = 0)
plot.ts(AR1_neg[1:50], ylim=ylm1, las = 1,
        ylab=expression(italic(x)[italic(t)]),
        main = "")
mtext(side = 3, expression(paste(phi[1]," = -0.7")),
      line = 0.4, adj = 0)
```

Same value, but different sign


## Coefficients of AR(1) models

```{r ar_comp_sm_big, fig.height=4}
set.seed(123)
## list description for AR(1) model with small coef
AR_bg <- list(order=c(1,0,0), ar=0.9, sd=0.1)
## list description for AR(1) model with large coef
AR_sm <- list(order=c(1,0,0), ar=0.1, sd=0.1)
## simulate AR(1)
AR1_bg <- arima.sim(n=500, model=AR_bg)
AR1_sm <- arima.sim(n=500, model=AR_sm)

## get y-limits for common plots
ylm2 <- c(min(AR1_bg[1:50],AR1_sm[1:50]), max(AR1_bg[1:50],AR1_sm[1:50]))

## set the margins & text size
par(mfrow = c(1,2), mai = c(0.8,0.8,0.3,0.2), oma = c(0,0,0,0))
## plot the ts
plot.ts(AR1_bg[1:50], ylim = ylm2, las = 1,
        ylab = expression(italic(x)[italic(t)]),
        main = "")
mtext(side = 3, expression(paste(phi[1]," = 0.9")),
      line = 0.4, adj = 0)
plot.ts(AR1_sm[1:50], ylim = ylm2, las = 1,
        ylab = expression(italic(x)[italic(t)]),
        main = "")
mtext(side = 3, expression(paste(phi[1]," = 0.1")),
      line = 0.4, adj = 0)
```

Both positive, but different magnitude


## Autocorrelation function (ACF)

Recall that the _autocorrelation function_ ($\rho_k$) measures the correlation between $\{x_t\}$ and a shifted version of itself $\{x_{t+k}\}$ 


## ACF for AR(1) models

```{r}
## set the margins & text size
par(mfrow=c(2,2), mai=c(0.8,0.8,0.3,0.2), oma=c(0,0,0,0))
## plot the ts
plot.ts(AR1_pos[1:50], ylim=ylm1, las = 1,
        ylab=expression(italic(x)[italic(t)]),
        main = "")
mtext(side = 3, expression(paste(phi[1]," = 0.7")),
      line = 0.4, adj = 0)
acf(AR1_pos, lag.max = 20, las = 1)
plot.ts(AR1_neg[1:50], ylim=ylm1, las = 1,
        ylab=expression(italic(x)[italic(t)]),
        main = "")
mtext(side = 3, expression(paste(phi[1]," = -0.7")),
      line = 0.4, adj = 0)
acf(AR1_neg, lag.max = 20, las = 1)
```

ACF oscillates for model with $-\phi$

## ACF for AR(1) models

```{r}
## set the margins & text size
par(mfrow = c(2,2), mai = c(0.8,0.8,0.3,0.2), oma = c(0,0,0,0))
## plot the ts
plot.ts(AR1_bg[1:50], ylim = ylm2, las = 1,
        ylab = expression(italic(x)[italic(t)]),
        main = "")
mtext(side = 3, expression(paste(phi[1]," = 0.9")),
      line = 0.4, adj = 0)
acf(AR1_bg, lag.max = 20, las = 1)
plot.ts(AR1_sm[1:50], ylim = ylm2, las = 1,
        ylab = expression(italic(x)[italic(t)]),
        main = "")
mtext(side = 3, expression(paste(phi[1]," = 0.1")),
      line = 0.4, adj = 0)
acf(AR1_sm, lag.max = 20, las = 1)
```

For model with large $\phi$, ACF has longer tail

## Partial autocorrelation funcion (PACF)

Recall that the _partial autocorrelation function_ ($\phi_k$) measures the correlation between $\{x_t\}$ and a shifted version of itself $\{x_{t+k}\}$, with the linear dependence of $\{x_{t-1},x_{t-2},\dots,x_{t-k-1}\}$ removed


## ACF & PACF for AR(_p_) models

```{r}
## set 3 AR coefficients
ARp3 <- list(c(0.7, 0.2, -0.1), c(-0.7, 0.2, 0.1))

expr <- list(expression(paste("AR(3) with ", phi[1], " = 0.7, ",
                              phi[2], " = 0.2, ", phi[3], " = -0.1")),
             expression(paste("AR(3) with ", phi[1], " = -0.7, ",
                              phi[2], " = 0.2, ", phi[3], " = 0.1")))

## empty list for storing models
AR3_mods <- vector("list", 2L)

par(mfrow = c(2,3), mai = c(0.7,0.7,0.3,0.1), omi = c(0,0,0,0))
## loop over orders of p
for(p in 1:2) {
  ## assume SD=1, so not specified
  AR3_mods[[p]] <- arima.sim(n=5000, list(ar=ARp3[[p]]))
  plot.ts(AR3_mods[[p]][1:50], las = 1,
          ylab = expression(italic(x[t])))
  acf(AR3_mods[[p]], lag.max = 20,
      las = 1, main = "")
  mtext(side = 3, expr[[p]],
        line = 0.5, adj = 0.5)
  pacf(AR3_mods[[p]], lag.max = 20,
       las = 1, main = "")
}
```


## PACF for AR(_p_) models

```{r}
## empty list for storing models
pacf_mods <- vector("list", 4L)

par(mfrow = c(2,2), mai = c(0.7,0.7,0.3,0.1), omi = c(0,0,0,0))
## loop over orders of p
for(p in 1:4) {
  pacf_mods[[p]] <- arima.sim(n=5000, list(ar=ARp[1:p]))
  pacf(pacf_mods[[p]], lag.max = 15,
       las = 1, main = "")
  mtext(side = 3, paste0("AR(",p,")"),
        line = 0.5, adj = 0)
}

```

Do you see the link between the order _p_ and lag _k_?


## Using ACF & PACF for model ID

|         | ACF               | PACF                    |
|:-------:|:-----------------:|:-----------------------:|
| AR(_p_) | Tails off slowly  | Cuts off after lag _p_  |


## Topics for today

Review

* White noise  

* Random walks  

Autoregressive (AR) models

Moving average (MA) models

Autoregressive moving average (ARMA) models

Using ACF & PACF for model ID
