---
title: "Regression versus State-Space"
author: "Eli Holmes"
date: "2 Mar 2021"
output:
  ioslides_presentation:
    css: lecture_slides.css
  beamer_presentation: default
subtitle: FISH 507 – Applied Time Series Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

## Fixed & random effects

Let's go back to Mark's [lecture on DFA](https://atsa-es.github.io/atsa/Lectures/Week%204/lec_08_intro_to_DFA.html#25)

$$
y_t = \underbrace{\alpha + \beta x_t}_{\text{fixed}} + ~ \underbrace{f_t + e_t}_{\text{random}}
$$

Regression (linear or non-linear) with correlated errors gets your the fixed effects while properly taking into account the correlated random effects.

State-space model allows you to model the $f_t$.

## Things to think about

$$
y_t = \underbrace{\alpha + \beta x_t}_{\text{fixed}} + ~ \underbrace{f_t + e_t}_{\text{random}}
$$

* What if your want $f_t$, the hidden random walk
* What if you want $E[\alpha + \beta x_t + f_t]$
* What if you want the $E[y_t | y_{1:t-1}]$
* What if we want to forecast $y_t$?

## Let's simulate some data

$$
\begin{gathered}
x_t = t \\
f_t = \phi f_{t-1} + w_t, \, w_t \sim N(0,1)\\
y_t = \beta x_t + f_t + v_t, \, v_t \sim N(0,\sqrt{0.2})
\end{gathered}
$$


```{r sim.fun, echo=FALSE}
sim.data <- function(N, phi = 1, theta = 1, beta = 0.2, intercept = 0,
                     r = 0.2, h=50){
x <- 1:N
e <- rep(NA, N); e[1] <- 0
iid.e <- 0
if(r != 0) iid.e <- rnorm(N, 0, sqrt(r))
for(t in 2:N) e[t] <- phi*e[t-1] + rnorm(1)
y <- intercept + beta*x + theta*e + iid.e

# forecasts
y.fr <- matrix(NA, 1000, h)
for(i in 1:1000){
  e2 <- rep(NA, h); e2[1] <- e[N]/theta
  for(t in 2:h) e2[t] <- phi*e2[t-1]+rnorm(1)
  x2 <- (N+1):(N+h)
  if(r != 0) iid.e <- rnorm(1, 0, sqrt(r))
  y.fr[i,] <- intercept + beta*x2 + theta*e2 + iid.e
}
return(list(y=y, y.fr=y.fr))
}
```

## Simulated data

```{r, echo=TRUE}
set.seed(123)
N <- 100; h <- 100; x <- 1:N
sim1 <- sim.data(N, h=h, phi=0.8)
sim2 <- sim.data(N, h=h, phi=1)
```

```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(sim1$y, main="phi=0.8")
plot(sim2$y, main="phi=1")
```

## Fit with arima(y, xreg=x)

Fits a linear regression with ARMA errors.

```{r, echo=TRUE}
library(forecast)
fit <- auto.arima(sim1$y, xreg=x)
fr1 <- forecast(fit, xreg=(N+1):(N+h))
fit <- auto.arima(sim2$y, xreg=x)
fr2 <- forecast(fit, xreg=(N+1):(N+h))
```

## Plot forecasts versus true

I created simulations from the true process to get truth.

Plot shows prediction intervals (future y). Red lines are true 80% intervals.

```{r, echo=FALSE}
par(mfrow=c(1,2))
y <- sim1$y
y.fr <- sim1$y.fr
y.fr.lims <- apply(y.fr, 2, quantile, probs=c(.1, .9))
fr <- fr1
ymin <- min(fr$lower,y.fr.lims[1,], y)
ymax <- max(fr$upper,y.fr.lims[2,], y)
plot(fr, main="LR w corr errors\nphi=0.8", ylim=c(ymin, ymax))
for(i in 1:2) lines((N+1):(N+h), y.fr.lims[i,], lty=2, col="red")

y <- sim2$y
y.fr <- sim2$y.fr
y.fr.lims <- apply(y.fr, 2, quantile, probs=c(.1, .9))
fr <- fr2
ymin <- min(fr$lower,y.fr.lims[1,], y)
ymax <- max(fr$upper,y.fr.lims[2,], y)
plot(fr, main="LR w corr errors\nphi=1 (random walk)", ylim=c(ymin, ymax))
for(i in 1:2) lines((N+1):(N+h), y.fr.lims[i,], lty=2, col="red")
```

## Fitted

You need to be careful to think about what you mean by `fitted()`

* `gls(y~x)` and similar would return $E[\alpha+\beta x_t]$
* `arima(y, xreg=x)` returns $E[\alpha+\beta x_t + e_t|y_{1:t-1}]$
* state-space model would get you $E[\alpha+\beta x_t + f_t|y_{1:N}]$

##

```{r, echo=FALSE}
par(mfrow=c(1,2))
library(nlme)
y <- sim1$y
fit.gls <- gls(y ~ x, correlation=corARMA(p=1, q=1))
plot(y, main="gls fitted\nfixed effects")
lines(fitted(fit.gls), col="red")

fit.lr <- auto.arima(y, xreg=x)
plot(y, main="arima fitted\nE[y|y(1:t-1]")
lines(fitted(fit.lr), col="red")
```


## Regression w ARMA errors vs ARMAX

These are different models.

Regression w AR1 errors

$$
\begin{gathered}
e_t = \phi e_{t-1} + w_t \\
y_t = \alpha + \beta x_t + e_t
\end{gathered}
$$

ARMAX: In this case, AR(1)-X

$$
y_t = \phi y_{t-1} + \alpha + \beta x_t + w_t
$$