---
title: "Holt and Holt-Winters Models -- Model Evaluation"
subtitle: FISH 550 – Applied Time Series Analysis
author: "Eli Holmes"
date: "15 April 2025"
output:
  ioslides_presentation:
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml-full.js"
    widescreen: true
    css: lecture_slides.css
---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
knitr::opts_chunk$set(fig.height=5, fig.align="center")
library(huxtable)
data(greeklandings, package="atsalibrary")
```

## Exponential Smoothing Models

[ETS Models in Forecasting: Principles and Practice](https://otexts.com/fpp2/expsmooth.html)

**Exponential smoothing models** are a family of forecasting tools that help us make smarter predictions by giving **more weight to recent observations** while not completely ignoring the past.

**Why "exponential"?**
The weights assigned to past observations decrease exponentially — meaning the most recent data gets the most attention, and older data fades out gradually.


## Simple Level Model

My prediction for today is an exponentially weighted average of past values.

$$
\hat{y}_{t+1} = \ell_t
$$

$$
\ell_t = \alpha y_t + \alpha(1 - \alpha) y_{t-1} + \alpha(1 - \alpha)^2 y_{t-2} + \dots
$$

## Simple Trend + Level Model (Holt)

My prediction includes both an exponentially weighted average of past values **and** a trend. 

"trend" = tendency to increase or decrease.

$$
\hat{y}_{t+1} = \ell_t + b_t
$$

Where:

* $\ell_t$: exponentially weighted average of past values  
* $b_t$: exponentially weighted average of past trends ($\ell_{t} - \ell_{t-1}$)


## Expanded Form

**Level:**
$$
\ell_t = \alpha y_t + \alpha(1 - \alpha)(y_{t-1} + b_{t-1}) + \alpha(1 - \alpha)^2(y_{t-2} + b_{t-2}) + \dots
$$

**Trend:**
$$
b_t = \beta (\ell_t - \ell_{t-1}) + \beta (1 - \beta) (\ell_{t-1} - \ell_{t-2}) + (1-\beta)^2 b_{t-2}
$$

## Holt-Winters Model

My forecast includes a weighted average of past values, a trend, **and** a seasonal component:


$$
\hat{y}_{t+1} = \ell_t + b_t + s_{t - m}
$$

Where:
- $\ell_t$: smoothed level (local average)
- $b_t$: smoothed trend (local slope)
- $s_{t - m}$: seasonal adjustment from one full cycle ago

## Expanded form of seasonality:

$$
\begin{aligned}
s_t &= \gamma (y_t - \ell_t) + (1 - \gamma) s_{t - m} \\
    &= \gamma (y_t - \ell_t) + \gamma(1 - \gamma)(y_{t - m} - \ell_{t - m}) \\
    &\quad + \gamma(1 - \gamma)^2(y_{t - 2m} - \ell_{t - 2m}) + \gamma(1 - \gamma)^3(y_{t - 3m} - \ell_{t - 3m}) + \dots
\end{aligned}
$$

This lets  the **seasonal pattern** evolve gradually over time.

## Fitting Holt and Holt-Winters

The forecast package provides:

* `holt(dat)`, `ets(dat, model="AAN")` Holt
* `hw(dat)`, `ets(dat, model="AAA")`  Holt-Winters
* `ets(dat, model="ZZZ")` others

## Fit Holt model

Evolving level and trend.

Load the data and **forecast** package.

```{r load_data_exp_smoothing}
data(greeklandings, package="atsalibrary")
anchovy <- subset(greeklandings, 
                  Species=="Anchovy" & Year<=1987)$log.metric.tons
anchovy <- ts(anchovy, start=1964)
library(forecast)
```

```{r fit.ann}
fit <- forecast::holt(anchovy)
fr <- forecast(fit, h=5)
```

##

```{r}
plot(fr)
```


## Validation

```{r load_packages, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)
library(reshape2)
library(tseries)
library(forecast)
```

Once we have a model(s), we will want to evaluate the performance of the model. We'll see examples using the anchovy landings.

```{r load_data2, message=FALSE, warning=FALSE}
spp <- "Anchovy"
training = subset(greeklandings, Year <= 1987)
test = subset(greeklandings, Year >= 1988 & Year <= 1989)
traindat <- subset(training, Species==spp)$log.metric.tons
testdat <- subset(test, Species==spp)$log.metric.tons
```

## Measures of forecast fit

To measure the forecast fit, we fit a model to training data and test a forecast against data in a test set.  We 'held out' the test data and did not use it for fitting.

```{r echo=FALSE, out.width="70%"}
plot(1964:1987, traindat, xlim=c(1964,1989))
points(1988:1989, testdat, pch=2, col="red")
legend("topleft", c("Training data","Test data"), pch=c(1,2), col=c("black", "red"))
```

##

We will fit to the training data and make a forecast.

```{r}
fit_arima <- auto.arima(traindat)
fit_arima
```
## Forecast

```{r}
fr <- forecast(fit_arima, h=2)
plot(fr)
```

## Compare to test

```{r echo=FALSE, out.width="70%"}
plot(fr)
points(25:26, testdat, pch=2, col="red")
legend("topleft", c("forecast","actual"), pch=c(20,2), col=c("blue","red"))
```

##

How do we quantify the difference between the forecast and the actual values?

```{r}
fr.err <- testdat - fr$mean
fr.err
```

There are many metrics.  The `accuracy()` function in forecast provides many different metrics: mean error, root mean square error, mean absolute error, mean percentage error, mean absolute percentage error.

##

### ME Mean error

```{r}
me <- mean(fr.err); me
```

### RMSE Root mean squared error

```{r}
rmse <- sqrt(mean(fr.err^2)); rmse
```

### MAE Mean absolute error

```{r}
mae <- mean(abs(fr.err)); mae
```

##

### MPE Mean percentage error

```{r}
fr.pe <- 100*fr.err/testdat
mpe <- mean(fr.pe); mpe
```

### MAPE Mean absolute percentage error

```{r}
mape <- mean(abs(fr.pe)); mape
```

##

```{r}
accuracy(fr, testdat)[,1:5]
```

```{r}
c(me, rmse, mae, mpe, mape)
```

## Test all the models in your candidate set

Compute metrics for all the models in your candidate set.

```{r}
# The model picked by auto.arima
fit1 <- Arima(traindat, order=c(0,1,1))
fr1 <- forecast(fit1, h=2)
test1 <- accuracy(fr1, testdat)[2,1:5]

# AR-1
fit2 <- Arima(traindat, order=c(1,1,0))
fr2 <- forecast(fit2, h=2)
test2 <- accuracy(fr2, testdat)[2,1:5]

# Holt
fit3 <- holt(traindat)
fr3 <- forecast(fit3, h=2)
test3 <- accuracy(fr3, testdat)[2,1:5]
```

## Show a summary

```{r results='asis', echo=FALSE}
sum.tests <- rbind(test1, test2, test3)
row.names(sum.tests) <- c("(0,1,1)","(1,1,0)","Holt")
sum.tests <- format(sum.tests, digits=3)
knitr::kable(sum.tests, format="html")
```

## Cross-validation

An alternate approach to testing a model's forecast accuracy is to use windows.  

```{r cv.sliding, echo=FALSE, out.width="70%"}
library(ggplot2)
p <- list()
for(i in 1:9){
  p[[i]]<-ggplot(subset(greeklandings, Species=="Anchovy"&Year<1974+i), aes(x=Year, y=log.metric.tons))+geom_point()+ylab("landings")+xlab("")+xlim(1964,1990)+ylim(8,12)+
    geom_point(data=subset(greeklandings, Species=="Anchovy"&Year==1974+i),aes(x=Year,y=log.metric.tons),color="red") +
    ggtitle(paste0("forecast ",i))
}
gridExtra::grid.arrange(
  p[[1]],p[[2]],p[[3]],p[[4]],p[[5]],p[[6]],p[[7]],p[[8]],p[[9]],nrow=3,
  top = grid::textGrob("Cross-validation: sliding window", gp=grid::gpar(fontsize=20,font=3))
)
```

##

Another approach uses a fixed window.  For example, a 10-year window.

```{r cv.fixed, echo=FALSE, out.width="70%"}
p <- list()
for(i in 1:9){
  p[[i]]<-ggplot(subset(greeklandings, Species=="Anchovy"&Year>=1964+i-1&Year<1974+i), aes(x=Year, y=log.metric.tons))+geom_point()+ylab("landings")+xlab("")+xlim(1964,1990)+ylim(8,12)+
    geom_point(data=subset(greeklandings, Species=="Anchovy"&Year==1974+i),aes(x=Year,y=log.metric.tons),color="red") +
    ggtitle(paste0("forecast ",i))
}
gridExtra::grid.arrange(
  p[[1]],p[[2]],p[[3]],p[[4]],p[[5]],p[[6]],p[[7]],p[[8]],p[[9]],nrow=3,
  top = grid::textGrob("Cross-validation: fixed window", gp=grid::gpar(fontsize=20,font=3))
)
```

## Cross-validation with the forecast package

```{r}
far2 <- function(x, h, order){
  forecast(Arima(x, order=order), h=h)
  }
e <- tsCV(traindat, far2, h=1, order=c(0,1,1))
tscv1 <- c(ME=mean(e, na.rm=TRUE), RMSE=sqrt(mean(e^2, na.rm=TRUE)), 
           MAE=mean(abs(e), na.rm=TRUE))
tscv1
```

Compare to RMSE from just the 2 test data points.
```{r}
test1[c("ME","RMSE","MAE")]
```

## Cross-validation farther in future

```{r cv.sliding.4plot, echo=FALSE, out.width="70%"}
p <- list()
for(i in 1:9){
  p[[i]]<-ggplot(subset(greeklandings, Species=="Anchovy"&Year<1974+i), aes(x=Year, y=log.metric.tons))+geom_point()+ylab("landings")+xlab("")+xlim(1964,1990)+ylim(8,12)+
    geom_point(data=subset(greeklandings, Species=="Anchovy"&Year==1974+i+3),aes(x=Year,y=log.metric.tons),color="red") +
    ggtitle(paste0("forecast ",i))
}
gridExtra::grid.arrange(
  p[[1]],p[[2]],p[[3]],p[[4]],p[[5]],p[[6]],p[[7]],p[[8]],p[[9]],nrow=3,
  top = grid::textGrob("Cross-validation: 4 step ahead forecast", gp=grid::gpar(fontsize=20,font=3))
)
```

##

Compare accuracy of forecasts 1 year out versus 4 years out.  If `h` is greater than 1, then the errors are returned as a matrix with each `h` in a column.  Column 4 is the forecast, 4 years out.

```{r cv.sliding.4}
e <- tsCV(traindat, far2, h=4, order=c(0,1,1))[,4]
#RMSE
tscv4 <- c(ME=mean(e, na.rm=TRUE), RMSE=sqrt(mean(e^2, na.rm=TRUE)), 
           MAE=mean(abs(e), na.rm=TRUE))
rbind(tscv1, tscv4)
```

##

Compare accuracy of forecasts with a fixed 10-year window and 1-year out forecasts.

```{r fixed.cv.1}
e <- tsCV(traindat, far2, h=1, order=c(0,1,1), window=10)
#RMSE
tscvf1 <- c(ME=mean(e, na.rm=TRUE), RMSE=sqrt(mean(e^2, na.rm=TRUE)), 
            MAE=mean(abs(e), na.rm=TRUE))
tscvf1
```

##

```{r results='asis'}
comp.tab <- rbind(test1=test1[c("ME","RMSE","MAE")],
      slide1=tscv1,
      slide4=tscv4,
      fixed1=tscvf1)
knitr::kable(comp.tab, format="html")
```

## Summary forecast performance

We can evaluate the forecast performance with forecasts of our test data or we can use all the data and use time-series cross-validation.

Let's start with the former.

```{r echo=FALSE}
spp <- "Anchovy"
training = subset(greeklandings, Year <= 1987)
test = subset(greeklandings, Year >= 1988 & Year <= 1989)

traindat <- ts(subset(training, Species==spp)$log.metric.tons, start=1964)
testdat <- ts(subset(test, Species==spp)$log.metric.tons, start=1988)
```

## Test forecast performance against a test data set

We will fit Holt model to the training data and make a forecast for the years that we 'held out'. 

```{r fig.height=4, echo=FALSE}
fit1 <- holt(traindat)
h=length(testdat)
fr <- forecast(fit1, h=h)
plot(fr)
points(end(traindat)[1]+1:h, testdat, pch=2, col="red")
legend("topleft", c("forecast","actual"), pch=c(20,2), col=c("blue","red"))
```

##

We can calculate a variety of forecast error metrics with

```{r}
accuracy(fr, testdat)
```

We would now repeat this for all the models in our candidate set and choose the model with the best forecast performance.

## Forecast performance with time-series cross-validation

We can use `tsCV()` as we did for ARIMA models.  We will redefine `traindat` as all our Anchovy data.

```{r subset.anch, echo=FALSE}
spp <- "Anchovy"
training = subset(greeklandings, Year <= 1989)
traindat <- subset(training, Species==spp)$log.metric.tons
```

## tsCV() function

We will use the `tsCV()` function.  We need to define a function that returns a forecast.

```{r def.far2}
far2 <- function(x, h){
  fit <- holt(x)
  forecast(fit, h=h)
  }
```

##

Now we can use `tsCV()` to run our `far2()` function to a series of training data sets.  We will specify that a NEW ets model be estimated for each training set.  We are not using the weighting estimated for the whole data set but estimating the weighting new for each set.

The `e` are our forecast errors for all the forecasts that we did with the data.

```{r}
e <- tsCV(traindat, far2, h=1)
c(ME=mean(e, na.rm=TRUE), RMSE=sqrt(mean(e^2, na.rm=TRUE)), 
           MAE=mean(abs(e), na.rm=TRUE))
```


## Seasonal Holt-Winters

```{r}
data("chinook", package="atsalibrary")
head(chinook.month)
```

##

The data are monthly and start in January 1990.  To make this into a ts object do

```{r}

df <- chinook.month %>% subset(State == "WA")
chinookts <- ts(df$log.metric.tons, start=c(1990,1), frequency=12)
```
`start` is the year and month and frequency is the number of months in the year.  

## Plot seasonal data


```{r out.width="70%"}
plot(chinookts)
```

## Holt-Winters


```{r}
library(forecast)
traindat <- window(chinookts, c(1990,1), c(1999,12))
fit <- forecast::hw(traindat)
```

##

```{r}
fr <- forecast(fit, h=24)
plot(fr)
points(window(chinookts, c(1996,1), c(1996,12)))
```

## Compare to a seasonal ARIMA model

`auto.arima()` will recognize that our data has season and fit a seasonal ARIMA model to our data.  Let's use the data that `ets()` used.  This is shorter than our training data.  The data used by `hw()` is returned in `fit$x`.

##

```{r out.width="70%"}
no_miss_dat <- fit$x
fit <- auto.arima(no_miss_dat)
fr <- forecast(fit, h=12)
plot(fr)
points(window(chinookts, c(1996,1), c(1996,12)))
```

## Missing values

Missing values are ok when fitting a seasonal ARIMA model, unlike an ETS model.

```{r out.width="70%"}
fit <- auto.arima(traindat)
fr <- forecast(fit, h=12)
plot(fr)
```

## Forecast evaluation

We can compute the forecast performance metrics as usual. 

```{r}
fit <- hw(traindat)
fr <- forecast(fit, h=12)
```
Look at the forecast so you know what years and months to include in your test data.  Pull those 12 months out of your data using the `window()` function.

```{r}
testdat <- window(traindat, c(1996,1), c(1996,12))
```

##

Use `accuracy()` to get the forecast error metrics.
```{r}
accuracy(fr, testdat)
```

##

We can do the same for the ARIMA model.

```{r}
no_miss_dat <- fit$x
fit <- auto.arima(no_miss_dat)
fr <- forecast(fit, h=12)
accuracy(fr, testdat)
```

