---
title: "MARSS model specification"
subtitle: "FISH 507 – Applied Time Series Analysis"
author: "Eli Holmes"
date: "22 Jan 2019"
output:
  ioslides_presentation:
  css: lecture_slides.css
---


## MARSS Package

We will be using the MARSS package to fit univariate and multivariate state-space models. 

The main function is `MARSS()`:

```
fit <- MARSS(data, model=list())
```

data are a vector or a matrix with time going along the columns.

model list is a list with the structure of all the parameters.

## MARSS model notation

$$x_t = \mathbf{B}x_{t-1} + \mathbf{U} + w_t, \,\,\, w_t \sim N(0,\mathbf{Q})$$
$$y_t = \mathbf{Z}x_{t} + \mathbf{A} + v_t, \,\,\, v_t \sim N(0,\mathbf{R})$$
  
The MARSS model list follows this notation one-to-one.

## Example 2 x time series

$$\begin{bmatrix}x_1 \\ x_2\end{bmatrix}_t = \begin{bmatrix}1 & 0 \\ 0 & 1\end{bmatrix}\begin{bmatrix}x_1 \\ x_2\end{bmatrix}_{t-1} + \begin{bmatrix}u \\ u\end{bmatrix}+\begin{bmatrix}w_1 \\ w_2\end{bmatrix}_t$$

$$\begin{bmatrix}w_1 \\ w_2\end{bmatrix}_t \sim N\left( 0, \begin{bmatrix}q & 0 \\ 0 & q\end{bmatrix} \right)$$
  
## Let's write out the x parameters

$$\mathbf{B} = \begin{bmatrix}1 & 0 \\ 0 & 1\end{bmatrix}$$
```{r}
B <- diag(1,2)
B
```

##

$$\mathbf{U} = \begin{bmatrix}u \\ u\end{bmatrix}$$

```{r}
U <- matrix("u",2,1)
U
```

A value in quotes is a character vector and is the name of a parameter to be estimated.

##

$$\mathbf{Q} = \begin{bmatrix}q & c \\ c & q\end{bmatrix}$$

```{r}
Q <- matrix(c("q","c","c","q"),2,2)
Q
```

or use
```
Q <- "equalvarcov"
```

## Example 2 y time series

$$\begin{bmatrix}y_1 \\ y_2\end{bmatrix}_t = \begin{bmatrix}1 & 0 \\ 0 & 1\end{bmatrix}\begin{bmatrix}x_1 \\ x_2\end{bmatrix}_t + \begin{bmatrix}0 \\ 0\end{bmatrix}+\begin{bmatrix}v_1 \\ v_2\end{bmatrix}_t$$

$$\begin{bmatrix}v_1 \\ v_2\end{bmatrix}_t \sim N\left( 0, \begin{bmatrix}r & 0 \\ 0 & r\end{bmatrix} \right)$$
  
## Let's write out the y parameters

The $\mathbf{Z}$ and $\mathbf{A}$ are just like $\mathbf{B}$ and $\mathbf{U}$, but $\mathbf{R}$ is different.

$$\mathbf{R} = \begin{bmatrix}r & 0 \\ 0 & r\end{bmatrix}$$

```{r}
R <- matrix(list(0),2,2)
diag(Q) <- "r"
R
```

or use
```
R <- "diagonal and equal"
```

$\mathbf{R}$ has both fixed values (the 0s) and estimated values (the $r$), so we need both numbers and charaters in our matrix.
  