---
title: "A gentle introduction to hidden markov models"
author: "Eric Ward"
date: "14 Feb 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
  beamer_presentation: default
subtitle: FISH 507 – Applied Time Series Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, 
                      message=FALSE, results='hide',
                      fig.align = 'center')
library(ggplot2)
library(atsar)
library(datasets)
library(MARSS)
library(dplyr)
```

## Overview of today's material

* Gentle introduction to HMMs

* Theory and notation

* Examples of univariate HMMs in R

* Examples of multivariate HMMs in R

## State space models 

We've already discussed state space models. These models include

* a latent process model (we don't directly observe)

* a data model (we've used normal data for this class)

## State space models 

Process model: 
$${ x }_{ t }={ x }_{ t-1 }+{ \varepsilon  }_{ t-1 }$$

Observation model: 
$${ y }_{ t }={ x }_{ t }+{ \delta  }_{ t }$$

where ${ \varepsilon  }_{ t } \sim Normal\left( 0,{ \sigma  }_{ \varepsilon  } \right)$ and ${ \delta  }_{ t } \sim Normal\left( 0,{ \sigma  }_{ \delta  } \right)$

## State space models

Adding AR coefficients can make these models stationary with respect to the mean,

$${ x }_{ t }={ p\cdot x }_{ t-1 }+{ \varepsilon  }_{ t-1 }$$

however they may not be able to explain some datasets very well. 

* Specifically, these models are not well designed to model regimes

## Regimes

Many examples of time series in ecology & fisheries may alternate between multiple states (2+)

* Vert-pre et al. (2013) https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3562848/

* Francis et al. (2012) https://onlinelibrary.wiley.com/doi/full/10.1111/j.1365-2486.2012.02702.x

## Regimes

Lots of non-HMM approaches for detecting regimes

* STARS algorithm
    + Sequential t-test approach for detecting changes in the mean
    + Rodionov (2015) https://www.mdpi.com/2225-1154/3/3/474

* Brute force model selection approach
    + iterate through change points, evaluating data support for each
    + how do we do change points with regression? ${ Y }_{ t }=B{ X }_{ t }+{ \varepsilon  }_{ t }$
    
## Regimes: simulating data

```{r}
set.seed(123)
state = 1
probs = matrix(c(0.95,0.05,0.05,0.95),2,2)
for(i in 2:100) {
  state[i] = sample(c(1,2), prob = probs[state[i-1],])
}
mu = c(1,3)
y = rnorm(length(state), mu[state], 0.5)
plot(1:100, y, xlab="Time", ylab = "Simulated data", type="b")
```

## Limitations of state space models

They can actually fit the data from a regime model quite well.

* Via MARSS,
```{r}

plot(1:100, y, xlab="Time", ylab = "Simulated data", type="b")
lines(MARSS(y)$states[1,], col="blue")

```

## Limitations of state space models

What's lacking is inference about:

* What's the probability of transition between regimes?
* How long are regimes expected to last?
* What regimes might we expect in the future?

Lots of applications: speech recognition, bioinformatics, animal movement, environmental data (rain), finance

## HMM: theory

Markov Process

* time may be discrete or continuous (we'll focus on discrete)
* Markov if at each time step, the state $x_{t}$ is only dependent on the previous state $x_{t-1}$
* $x_{t}$ does not depend on future values

Entire history can be written as $$\left\{ { x }_{ 1 },{ x }_{ 2 },{ x }_{ 3 },...,{ x }_{ T } \right\}$$

## HMM: theory

A key piece of the Markov process are the transition probabilities.

* The probability of transitioning from state $j$ to $i$ given the current state is 
$$P\left( { x }_{ t+1 }=j | { x }_{ t }=i \right) ={ \gamma  }_{ ij }$$

And then these probabilities can be summarized in a transition matrix, 
$$\Gamma =\left[ \begin{matrix} { \gamma  }_{ 11 } & { \gamma  }_{ 12 } & { \gamma  }_{ 13 } \\ { \gamma  }_{ 21 } & { \gamma  }_{ 22 } & { \gamma  }_{ 23 } \\ { \gamma  }_{ 31 } & { \gamma  }_{ 32 } & { \gamma  }_{ 33 } \end{matrix} \right]$$

## HMM theory

Matrix $\Gamma$ is the 1-step transitions. However k-step transition probabilities can be generated, 

$\Gamma(k) = \Gamma^{k}$

* From this, we can also calculate the stationary distribution of the chain
    + See Zucchini et al. (2006) Chapter 2

## HMM theory

There are two general flavours of transition matrices:

* Homogenous (or stationary)
    + transition probabilities don't depend on $t$

* Non-homogeneous
    + transition probabilities are time-varying

* In this class, we'll only consider **homogeneous** cases

## HMM: theory

* Observations: observable data ${Y}_{ i=1,...,N }$

* States: latent (discrete) variables that are not directly observed
    + ${x}_{ t=1,...,T }$
    + $N$ is the number of states possible

* Transition probabilities: transition matrix representing the probability of transitioning between states in the Markov chain
    + $\Gamma$ and ${ \gamma  }_{ ij }$
    
* Emission probabilities: how likely the states are at any particular timestep
    + ${ \theta  }_{ i=1,...,N }$
