---
title: "Bayesian stan exercises"
output:
  html_document: default
  pdf_document: default
date: "2023-05-02"
---

## Fitting a univariate model in Stan

The `atsar` package contains a bunch of bundled Stan files for fitting univariate state space models in Stan, for those who want to try those methods. There's also a bunch of options that we didn't talk about in class, but can explore those in more detail here.

As a first example, we'll use the MARSS gray whale dataset.

```{r}
library(ggplot2)
library(shinystan)
library(rstan)
library(broom.mixed)
library(bayesplot)
```

```{r}
whale <- as.data.frame(MARSS::graywhales)

ggplot(whale, aes(Year,Count)) + 
  geom_point() + geom_smooth()
```

An interesting question might be whether we can fit a Gompertz a population model and estimate the long term trend. To do this, we'll use the `atsar` package.

```{r warning=FALSE, message=FALSE}
remotes::install_github("atsa-es/atsar")
```

It might be helpful to look at the documentation to see what kind of univariate models are available,

```{r}
?atsar::fit_stan
```

For the whale example, let's start with the `ss_rw` model, a state space univariate model. We will fit the model in log-space and estimate drift.

```{r}
library(atsar)
```

We will start with a few hundred samples (normally you want the burn in period to be 1000-2000)

```{r message=FALSE, warning=FALSE, results='hide'}
set.seed(123)
fit <- fit_stan(y = log(whale$Count),
                model_name = "ss_rw",
                est_drift = TRUE,
                list(n_mcmc = 500, 
                     n_burn = 100, 
                     n_chain = 3))
```

Now let's look at the R-hat values. Generally we want these to be \< 1.1. Does it look like the model converges ok?

```{r}
fit
```

It might be good to look at the posterior plots with ShinyStan. Specifically, let's look at the posterior distributions and trace plots of the two variance parameters. Ask:

-   are posterior distributions multi-modal?

-   do MCMC trace plots look stationary?

```{r}
library(shinystan)
shinystan::launch_shinystan(fit)
```

Let's try to look at the posterior of the trend -- is it generally overlapping 0? Do you think the drift term should be kept in the model or not?

```{r}
pars <- as.data.frame(rstan::extract(fit))

ggplot(pars, aes(mu)) + 
  geom_histogram()
```

### Can we estimate density dependence?

Density dependence is commonly interpreted as the autoregression term,

$$
x_t = \phi*x_{t-1}+\mu+d_t; d_t \sim N(0,\sigma)
$$

We can estimate the autoregression term by changing our model to `ss_ar` ,

```{r message=FALSE, warning=FALSE}
set.seed(123)
```

```{r message=FALSE, warning=FALSE, results='hide'}
fit2 <- fit_stan(y = log(whale$Count),
                model_name = "ss_ar",
                est_drift = TRUE,
                list(n_mcmc = 500, 
                     n_burn = 100, 
                     n_chain = 3))
```

Questions:

-   Did this model appear to converge? Why or why not?

-   Does the `phi` parameter appear well estimated?

-   Extract the states, and plot them with the original data. For bonus points, include 95% Credible intervals

### Changing the family

In the above model(s) and all of MARSS, we have been modeling log-transformed data. Instead, the Stan code allows us to easily change the family. Given that the original whale dataset consists of counts, let's try to change the family

-   <div>

    ```{r message=FALSE, warning=FALSE, results='hide'}
    set.seed(123)
    fit3 <- fit_stan(y = log(whale$Count),
                    model_name = "ss_rw",
                    est_drift = TRUE,
                    list(n_mcmc = 500, 
                         n_burn = 100, 
                         n_chain = 3),
                    family="poisson")
    ```

    </div>

Questions:

-   How do the estimates of the trend differ between this approach

-   Calculate the correlation between the mean state estimates in this model, versus the model we used initially (`fit`). The Poisson model uses a log-link -- and so the predicted states are in log-space. Is this relationship expected?

### Bonus

1.  Fit MARSS models to any of the above, and plot the state estimates versus the estimates from the Stan models. How similar are they?

```{=html}
<!-- -->
```
2.  Play with broom.mixed -- this is often a more efficient way to extract things like predicted values or parameter estimates,

```{r}
broom.mixed::tidy(fit, pars="pred")
```

3.  Make some alternative plots of distributions and traceplots using `bayesplot`. For example,

```{r}
bayesplot::mcmc_hist_by_chain(fit, pars = "sigma_obs")
```

4.  If you want to look at the source code (in Stan) for any of the above models, you can do something like this

```{r}
cat( get_stancode(fit) ) # cat makes this readable

```

2.  
