---
title: "A gentle introduction to hidden markov models"
author: "Eric Ward"
date: "14 Feb 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
  beamer_presentation: default
subtitle: FISH 507 – Applied Time Series Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, 
                      message=FALSE, results='hide',
                      fig.align = 'center')
library(ggplot2)
library(atsar)
library(datasets)
library(MARSS)
library(dplyr)
```

## Overview of today's material

* Gentle introduction to HMMs

* Theory and notation

* Examples of univariate HMMs in R

* Examples of multivariate HMMs in R

## State space models 

We've already discussed state space models. These models include

* a latent process model (we don't directly observe)

* a data model (we've used normal data for this class)

## State space models 

Process model: 
$${ x }_{ t }={ x }_{ t-1 }+{ \varepsilon  }_{ t-1 }$$

Observation model: 
$${ y }_{ t }={ x }_{ t }+{ \delta  }_{ t }$$

where ${ \varepsilon  }_{ t } \sim Normal\left( 0,{ \sigma  }_{ \varepsilon  } \right)$ and ${ \delta  }_{ t } \sim Normal\left( 0,{ \sigma  }_{ \delta  } \right)$

## State space models

Adding AR coefficients can make these models stationary with respect to the mean,

$${ x }_{ t }={ p\cdot x }_{ t-1 }+{ \varepsilon  }_{ t-1 }$$

however they may not be able to explain some datasets very well. 

* Specifically, these models are not well designed to model regimes

## Regimes

Many examples of time series in ecology & fisheries may alternate between multiple states (2+)

* Vert-pre et al. (2013) https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3562848/

* Francis et al. (2012) https://onlinelibrary.wiley.com/doi/full/10.1111/j.1365-2486.2012.02702.x

## Regimes

Lots of non-HMM approaches for detecting regimes

* STARS algorithm
    + Sequential t-test approach for detecting changes in the mean
    + Rodionov (2015) https://www.mdpi.com/2225-1154/3/3/474

* Brute force model selection approach
    + iterate through change points, evaluating data support for each
    + how do we do change points with regression? ${ Y }_{ t }=B{ X }_{ t }+{ \varepsilon  }_{ t }$
    
## Regimes: simulating data

```{r}
set.seed(123)
state = 1
probs = matrix(c(0.95,0.05,0.05,0.95),2,2)
for(i in 2:100) {
  state[i] = sample(c(1,2), prob = probs[state[i-1],])
}
mu = c(1,3)
y = rnorm(length(state), mu[state], 0.5)
plot(1:100, y, xlab="Time", ylab = "Simulated data", type="b")
```

## Limitations of state space models

They can actually fit the data from a regime model quite well.

* Via MARSS,
```{r}

plot(1:100, y, xlab="Time", ylab = "Simulated data", type="b")
lines(MARSS(y)$states[1,], col="blue")

```

## Limitations of state space models

What's lacking is inference about:

* What's the probability of transition between regimes?
* How long are regimes expected to last?
* What regimes might we expect in the future?

