---
title: "Multivariate Generalized Additive Models (GAMs)"
subtitle: FISH 550 – Applied Time Series Analysis
author: ""
date: ""
output:
  ioslides_presentation:
    css: lecture_slides.css
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, 
                      message=FALSE, results='hide',
                      fig.align = 'center')
library(ggplot2)
library(datasets)
library(MARSS)
library(tidyr)
library(dplyr)
library(mgcv)
library(lubridate)
library(viridisLite)
library(here)
library(mvgam)
```

## Overview of today's material

-   Multivariate models in `mgcv` (Pederson et al. 2019)[https://doi.org/10.7717/peerj.6876]

-   Flexible dynamic models with `mvgam` (Clark et al. 2022)[https://doi.org/10.1111/2041-210X.13974]

## Multivariate models in `mgcv`

- 

## Introducing factor smooths

- **What is a factor smooth?** 
- A factor smooth is a smooth term in a model where a categorical variable interacts with a continuous predictor
- This type of smooth allows modeling of varying relationships between continuous predictors across different levels of a factor

## How do we implement these in `mgcv`?

- Use the `s()` function to specify smooth terms.
- To specify a factor smooth, we include the factor variable in the formula:

```{r echo = TRUE, eval = FALSE}
gam(response ~ s(predictor, by = factor_variable), 
    data = dataset)
```

## Illustrating different models with the harbor seal dataset

```{r echo = TRUE, eval = TRUE}
data("harborSeal")
long_seal <- as.data.frame(harborSeal) |>
  pivot_longer(cols = starts_with("CoastalEstuaries"):
                 starts_with("Georgia.Strait"), 
               names_to = "location", 
               values_to = "value")

long_seal$location <- as.factor(long_seal$location)
```

## Model G

* Global smooth (shared trend)
```{r echo = TRUE}
fit_G <- gam(value ~ s(Year, k=8, bs="cr"), #<<
           data=long_seal, 
           family="gaussian")
```

## Model G fits

```{r}
my_cols <- viridis(3, begin = 0.2, end = 0.8)
long_seal$pred_G <- predict(fit_G, long_seal)

ggplot(long_seal, aes(Year, pred_G)) + 
  geom_line(col = my_cols[1]) + 
  geom_point(aes(Year, value), col = "red") + 
  facet_wrap(~ location, scale="free_y") + 
  ylab("Predicted") + theme_bw() + ylim(c(5.5,11))
```

## Model G

* Global smooth (shared trend)
* Time series specific intercepts (n = 12)
```{r echo = TRUE}
fit_G <- gam(value ~ s(Year, k=8, bs="cr") + 
             s(location, k=12, bs="re"), #<<
           data=long_seal, 
           family="gaussian")
```

## Model G fits

```{r}
my_cols <- viridis(3, begin = 0.2, end = 0.8)
long_seal$pred_G <- predict(fit_G, long_seal)

ggplot(long_seal, aes(Year, pred_G)) + 
  geom_line(col = my_cols[1]) + 
  geom_point(aes(Year, value), col = "red") + 
  facet_wrap(~ location, scale="free_y") + 
  ylab("Predicted") + theme_bw() + ylim(c(5.5,11))
```

## Model GS

* Global smooth (shared trend)
* Time series specific smooths
```{r echo = TRUE}
fit_GS <- gam(value ~ s(Year, k=8, bs="cr") + 
             s(Year, by = location, k = 8, bs="fs") + #<<
               s(location, k=12, bs="re"), 
           data=long_seal, 
           family="gaussian")
```

## Model GS fits

```{r}
long_seal$pred_GS <- predict(fit_GS, long_seal)

ggplot(long_seal, aes(Year, pred_GS)) + 
  geom_line(aes(Year, pred_G), col = my_cols[1]) +
  geom_point(aes(Year, value), col = "red") + 
  facet_wrap(~ location, scale="free_y") + 
  ylab("Predicted") + theme_bw() + 
  geom_line(col = my_cols[2]) + ylim(c(5.5,11))
```

## Model S

* Time series specific smooths
* No global smooth
```{r echo = TRUE}
fit_S <- gam(value ~  
             s(Year, by = location, k = 8, bs="fs") + 
               s(location, k=12, bs="re"), 
           data=long_seal, 
           family="gaussian")
```

## Model S fits

```{r}
long_seal$pred_S <- predict(fit_S, long_seal)

ggplot(long_seal, aes(Year, pred_S)) + 
  geom_line(aes(Year, pred_G), col = my_cols[1]) +
  geom_line(aes(Year, pred_GS), col = my_cols[2]) +
  geom_point(aes(Year, value), col = "red") + 
  facet_wrap(~ location, scale="free_y") + 
  ylab("Predicted") + theme_bw() + 
  geom_line(col = my_cols[3]) + 
  ylim(c(5.5,11))
```

## Factor Smooths

* Add flexibility for time - series specific smooths

* Similar to MARSS model with either shared (single trend) or 
MARSS model where each time series is a different trend

* Wiggliness of both global and factor smooths can be specified

## Summarizing these approaches in `mgcv`

* Lots of flexibility 

* We could have used `gamm` for additional flexibility in random effects

* Implements mixed models 

## Transitioning to DGAMs in `mvgam`

- **DGAMs** are an extension of GAMs designed to handle time-varying and dynamic data
- Particularly useful for modeling temporal dependencies / nonlinear relationships 
- Can be used with multivariate time series data
- DGAMs incorporate dynamic smooth terms that evolve over time

## Why use DGAMs? 

- GAMs are great, but uncertainty blows up as we get further from data (forecast)

- Random walk / AR1 processes also have uncertainty balloon

- DGAMs represent a hybrid approach that work particularly well for forecasting

## Mathematically

Remember that GAMs are typically represented as
$$ E[Y_t] = \beta_0 + \sum_{i=1}^{I} s(x_{i,t})$$
where $Y_t$ is the response variable at time $t$, $\beta_0$ is the intercept, and $s(x_{i,t})$ are smooth functions of the predictors

## Mathematically

Because the response can be non-Gaussian we use a link function $g^{=1}()$ to relate the mean of the response to the linear predictor

$$ E[Y_t] = g^{=1} \left( \beta_0 + \sum_{i=1}^{I} s(x_{i,t}) \right)$$
where $g$ is the link function (e.g., logit, log, identity)

## Mathematically

DGAMs add a dynamic component to this equation, allowing for time-varying smooths

$$ E[Y_t] = g^{-1} \left( \beta_0 + \sum_{i=1}^{I} s(x_{i,t}) + z_{t} \right)$$
where $z_{t} = \psi_{0} + \psi_{1}z_{t-1} + \epsilon_t$ and $\epsilon_t \sim N(0, \sigma_{\epsilon})$ and $g^{-1}$ is the inverse link function

## Complex latent processes

Random walks can be specified with any dynamic model that we've already discussed (AR, MA processes, etc)

$z_{t} = \psi_{0} + \psi_{1}z_{t-1} + \epsilon_t$

$z_{t} = \psi_{0} + \psi_{1}z_{t-1} + \psi_{2}z_{t-2} + \epsilon_t$

## Dynamic factor processes

For multiple time series, there can be > 1 latent trend, and we can estimate factor loadings (DFA). For time series $j$

$$ E[Y_{j,t}] = g^{-1} \left( \beta_0 + \sum_{i=1}^{I} s(x_{i,j,t}) + \sum_{m=1}^{M} z_{m,t}\theta_{j} \right)$$

## The mvgam Package

- The **`mvgam`** package provides tools for fitting DGAMs in R
- Bayesian estimation with Stan backend
- Supports multivariate time series modeling with dynamic smooths
- Provides an interface to easily define dynamic terms in the model

## Harbor seal example with `mvgam`

* Before, we were using factor smooths to model harbor seal trends
* We can fit these exact same models (without trend) in `mvgam`
```{r eval = FALSE, echo=TRUE, message=FALSE, warning=FALSE}
fit_GS <- mvgam(value ~ s(Year, k=8, bs="cr") +#<< 
             s(Year, by = location, k = 8, bs="fs") + 
               s(location, k=12, bs="re"), 
           data=long_seal, 
           family="gaussian")
```

## 


## More examples: factor analysis

## Forecasting comparison vs GAMs


