---
title: "Multivariate Generalized Additive Models (GAMs)"
subtitle: FISH 550 – Applied Time Series Analysis
author: ""
date: ""
output:
  ioslides_presentation:
    css: lecture_slides.css
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, 
                      message=FALSE, results='hide',
                      fig.align = 'center')
library(ggplot2)
library(datasets)
library(MARSS)
library(tidyr)
library(dplyr)
library(mgcv)
library(lubridate)
library(viridisLite)
library(here)
library(mvgam)
```

## Overview of today's material

-   Multivariate models in `mgcv` (Pederson et al. 2019)[https://doi.org/10.7717/peerj.6876]

-   Flexible dynamic models with `mvgam` (Clark et al. 2022)[https://doi.org/10.1111/2041-210X.13974]

## Multivariate models in `mgcv`

- Often we have multiple related response variables

- Examples: different locations, species, or strata

- Goal: model shared and individual trends across responses while accounting for correlation and partial pooling

## Multivariate models in `mgcv`

- Lots of ways we can think to model variation by time series

- Different intercepts / trends (harbor seal example in MARSS)

- Site:year random effects (GLMMs)

- Factor smooths (specific to GAMs)

## Introducing factor smooths

- **What is a factor smooth?** 
- A factor smooth is a smooth term in a model where a categorical variable interacts with a continuous predictor
- This type of smooth allows modeling of varying relationships between continuous predictors across different levels of a factor

## How do we implement these in `mgcv`?

- Use the `s()` function to specify smooth terms.
- To specify a factor smooth, we include the factor variable in the formula:

```{r echo = TRUE, eval = FALSE}
gam(response ~ s(predictor, by = factor_variable), 
    data = dataset)
```

- When `by = factor`, a separate smooth is estimated for each level

## Illustrating different models with the harbor seal dataset

```{r echo = TRUE, eval = TRUE}
data("harborSeal")
long_seal <- as.data.frame(harborSeal) |>
  pivot_longer(cols = starts_with("CoastalEstuaries"):
                 starts_with("Georgia.Strait"), 
               names_to = "location", 
               values_to = "value")

long_seal$location <- as.factor(long_seal$location)
```

## Three models for comparison

- Model G: global smooth

- Model GS: global smooth, time series - specific smooths

- Model S: time series - specific smooths

## Model G

* Global smooth (shared trend)
```{r echo = TRUE}
fit_G <- gam(value ~ s(Year, k=8, bs="cr"), #<<
           data=long_seal, 
           family="gaussian")
```

## Model G fits

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}
my_cols <- viridis(3, begin = 0.2, end = 0.8)
long_seal$pred_G <- predict(fit_G, long_seal)

ggplot(long_seal, aes(Year, pred_G)) + 
  geom_line(col = my_cols[1]) + 
  geom_point(aes(Year, value), col = "red") + 
  facet_wrap(~ location, scale="free_y") + 
  ylab("Predicted") + theme_bw() + ylim(c(5.5,11))
```

## Model G

* Global smooth (shared trend)
* Time series specific intercepts (n = 12)
```{r echo = TRUE}
fit_G <- gam(value ~ s(Year, k=8, bs="cr") + 
             s(location, k=12, bs="re"), #<<
           data=long_seal, 
           family="gaussian")
```

## Model G fits

```{r}
my_cols <- viridis(3, begin = 0.2, end = 0.8)
long_seal$pred_G <- predict(fit_G, long_seal)

ggplot(long_seal, aes(Year, pred_G)) + 
  geom_line(col = my_cols[1]) + 
  geom_point(aes(Year, value), col = "red") + 
  facet_wrap(~ location, scale="free_y") + 
  ylab("Predicted") + theme_bw() + ylim(c(5.5,11))
```

## Model GS

* Global smooth (shared trend)
* Time series specific smooths
```{r echo = TRUE}
fit_GS <- gam(value ~ s(Year, k=8, bs="cr") + 
             s(Year, by = location, k = 8, bs="fs") + #<<
               s(location, k=12, bs="re"), 
           data=long_seal, 
           family="gaussian")
```

## Model GS fits

```{r}
long_seal$pred_GS <- predict(fit_GS, long_seal)

ggplot(long_seal, aes(Year, pred_GS)) + 
  geom_line(aes(Year, pred_G), col = my_cols[1]) +
  geom_point(aes(Year, value), col = "red") + 
  facet_wrap(~ location, scale="free_y") + 
  ylab("Predicted") + theme_bw() + 
  geom_line(col = my_cols[2]) + ylim(c(5.5,11))
```

## Model S

* Time series specific smooths
* No global smooth
```{r echo = TRUE}
fit_S <- gam(value ~  
             s(Year, by = location, k = 8, bs="fs") + 
               s(location, k=12, bs="re"), 
           data=long_seal, 
           family="gaussian")
```

## Model S fits

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=3, fig.width=7}
long_seal$pred_S <- predict(fit_S, long_seal)

ggplot(long_seal, aes(Year, pred_S)) + 
  geom_line(aes(Year, pred_G), col = my_cols[1]) +
  geom_line(aes(Year, pred_GS), col = my_cols[2]) +
  geom_point(aes(Year, value), col = "red") + 
  facet_wrap(~ location, scale="free_y") + 
  ylab("Predicted") + theme_bw() + 
  geom_line(col = my_cols[3]) + 
  ylim(c(5.5,11))
```

## Comparing fits

```{r echo=TRUE}
AIC(fit_G, fit_GS, fit_S)
```

## Factor Smooths

* Add flexibility for time - series specific smooths

* Similar to MARSS model with either shared (single trend) or 
MARSS model where each time series is a different trend

* Wiggliness of both global and factor smooths can be specified

## Summarizing these approaches in `mgcv`

* Lots of flexibility 

* We could have used `gamm` for additional flexibility in random effects

* Implements mixed models (year:site random effects, etc)

## Transitioning to DGAMs in `mvgam`

* **DGAMs** are an extension of GAMs designed to handle time-varying and dynamic data

* Particularly useful for modeling temporal dependencies / nonlinear relationships 

* Can be used with multivariate time series data

* DGAMs incorporate dynamic smooth terms that evolve over time

## Why use DGAMs? 

* GAMs are great, but uncertainty blows up as we get further from data (forecast)

* Random walk / AR1 processes also have uncertainty balloon

* DGAMs represent a hybrid approach that can work particularly well for forecasting

## GAM forecast uncertainty

* `y ~ s(t, k = 12)`
* low uncertainty, poor fit

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=3, fig.width=7}
library(mgcv)
set.seed(42)

# Simulate a seasonal time series
n <- 100
x <- 1:n
y <- sin(2 * pi * x / 12) + rnorm(n, 0, 0.3)
df <- data.frame(x = x, y = y)

# Fit GAM
fit <- gam(y ~ s(x, k = 12), data = df)

# Create a prediction dataset
x_future <- data.frame(x = 1:130)  # extend 30 steps ahead
pred <- predict(fit, newdata = x_future, se.fit = TRUE)

# Plot
df_pred <- data.frame(
  x = x_future$x,
  fit = pred$fit,
  lwr = pred$fit - 2 * pred$se.fit,
  upr = pred$fit + 2 * pred$se.fit
)

library(ggplot2)
ggplot(df_pred, aes(x, fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "grey80") +
  geom_line() +
  geom_point(data = df, aes(x, y), col = "blue", size = 1) +
  labs(y = "Response", title = "Forecasting with a non-linear GAM") +
  geom_vline(xintercept = 100, linetype = "dashed", col = "red") +
  annotate("text", x = 115, y = 2, label = "Forecast horizon", angle = 90, vjust = -0.5) +
  theme_minimal()
```

## GAM forecast uncertainty

* `y ~ s(t, k = 50)`
* good fit, high uncertainty

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=3, fig.width=7}
set.seed(42)

# Simulate a seasonal time series
n <- 100
x <- 1:n
y <- sin(2 * pi * x / 12) + rnorm(n, 0, 0.3)
df <- data.frame(x = x, y = y)

# Fit GAM
fit <- gam(y ~ s(x, k = 50), data = df)

# Create a prediction dataset
x_future <- data.frame(x = 1:130)  # extend 30 steps ahead
pred <- predict(fit, newdata = x_future, se.fit = TRUE)

# Plot
df_pred <- data.frame(
  x = x_future$x,
  fit = pred$fit,
  lwr = pred$fit - 2 * pred$se.fit,
  upr = pred$fit + 2 * pred$se.fit
)

library(ggplot2)
ggplot(df_pred, aes(x, fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "grey80") +
  geom_line() +
  geom_point(data = df, aes(x, y), col = "blue", size = 1) +
  labs(y = "Response", title = "Forecasting with a non-linear GAM") +
  geom_vline(xintercept = 100, linetype = "dashed", col = "red") +
  annotate("text", x = 115, y = 2, label = "Forecast horizon", angle = 90, vjust = -0.5) +
  theme_minimal()
```

## Mathematically

Remember that GAMs are typically represented as
$$ E[Y_t] = \beta_0 + \sum_{i=1}^{I} s(x_{i,t})$$
where $Y_t$ is the response variable at time $t$, $\beta_0$ is the intercept, and $s(x_{i,t})$ are smooth functions of the predictors

## Mathematically

Because the response can be non-Gaussian we use a link function $g^{=1}()$ to relate the mean of the response to the linear predictor

$$ E[Y_t] = g^{=1} \left( \beta_0 + \sum_{i=1}^{I} s(x_{i,t}) \right)$$
where $g$ is the link function (e.g., logit, log, identity)

## Mathematically

DGAMs add a dynamic component to this equation, allowing for time-varying smooths

$$ E[Y_t] = g^{-1} \left( \beta_0 + \sum_{i=1}^{I} s(x_{i,t}) + z_{t} \right)$$
where $z_{t} = \psi_{0} + \psi_{1}z_{t-1} + \epsilon_t$ and $\epsilon_t \sim N(0, \sigma_{\epsilon})$ and $g^{-1}$ is the inverse link function

## Complex latent processes

Random walks can be specified with any dynamic model that we've already discussed (AR, MA processes, etc)

$z_{t} = \psi_{0} + \psi_{1}z_{t-1} + \epsilon_t$

$z_{t} = \psi_{0} + \psi_{1}z_{t-1} + \psi_{2}z_{t-2} + \epsilon_t$

## Dynamic factor processes

For multiple time series, there can be > 1 latent trend, and we can estimate factor loadings (DFA). For time series $j$

$$ E[Y_{j,t}] = g^{-1} \left( \beta_0 + \sum_{i=1}^{I} s(x_{i,j,t}) + \sum_{m=1}^{M} z_{m,t}\theta_{j} \right)$$

## The mvgam Package

* The **`mvgam`** package provides tools for fitting DGAMs in R

* Bayesian estimation with Stan backend

* Supports multivariate time series modeling with dynamic smooths

* Provides an interface to easily define dynamic terms in the model

## Simulated data example with `mvgam`

* As expected, this is giving the same result as `mgcv`

* `fit <- mvgam(y ~ s(time, k = 50), 
             data = df, family = "gaussian")`

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=3, fig.width=7}
set.seed(42)

# Simulate a seasonal time series
n <- 100
x <- 1:n
y <- sin(2 * pi * x / 12) + rnorm(n, 0, 0.3)
df <- data.frame(time = x, y = y)

# Fit GAM
fit <- mvgam(y ~ s(time, k = 50), 
             data = df, family = "gaussian",
             chains = 1)

# Create a prediction dataset
x_future <- data.frame(time = 1:130)  # extend 30 steps ahead
pred <- predict(fit, newdata = x_future, se.fit = TRUE)
colnames(pred)[3:4] <- c("Q25","Q75")
pred <- as.data.frame(pred)
# Plot
df_pred <- data.frame(
  x = x_future$time,
  fit = pred$Estimate,
  lwr = pred$Q25,
  upr = pred$Q75
)

library(ggplot2)
ggplot(df_pred, aes(x, fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "grey80") +
  geom_line() +
  geom_point(data = df, aes(x, y), col = "blue", size = 1) +
  labs(y = "Response", title = "Forecasting with a non-linear GAM") +
  geom_vline(xintercept = 100, linetype = "dashed", col = "red") +
  #annotate("text", x = 115, y = 2, label = "Forecast horizon", angle = 90, vjust = -0.5) +
  theme_minimal()
```

## Harbor seal example with `mvgam`

* Before, we were using factor smooths to model harbor seal trends

* We can fit these exact same models (without trend) in `mvgam`

```{r eval = TRUE, echo=TRUE, message=FALSE, warning=FALSE}
fit_GS <- mvgam(value ~ s(Year, k=8, bs="cr") +#<< 
             s(Year, by = location, k = 8, bs="fs") + 
               s(location, k=12, bs="re"), 
           data=long_seal, 
           family="gaussian", chains = 1)
```

## Plotting the main trend

```{r echo = TRUE}
plot_mvgam_smooth(fit_GS)
```

## Including dynamic components: coastal upwelling

*Jacox MG, CA Edwards, EL Hazen, SJ Bograd (2018) Coastal upwelling revisited: Ekman, Bakun, and improved upwelling indices for the U.S. west coast, J. Geophysical Research, 123(10), 7332-7350.*

- CUTI estimates of vertical transport near the coast

- BEUTI estimates of vertical nitrate flux

## Including dynamic components: coastal upwelling

- CUTI collected at monthly time steps

- 1 deg latitude resolution (31 - 47) -- we'll use every other

## Including dynamic components: coastal upwelling

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=3, fig.width=7}
cuti <- read.csv("CUTI_monthly.csv")

library(dplyr)
library(tidyr)
library(stringr)

# Subset every other column (X31N, X33N, etc.)
cuti_subset <- cuti %>%
  select(year, month, matches("^X\\d+N")) %>%
  select(year, month, which(!duplicated(colnames(.)))[c(TRUE, FALSE)])  # keep every other variable column

# Pivot to long format and clean up variable names
cuti_long <- cuti_subset %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "latitude",
    values_to = "cuti"
  ) %>%
  mutate(latitude = as.integer(str_extract(latitude, "\\d+")))  # Extract latitude number
cuti_long$latitude <- as.factor(cuti_long$latitude)
cuti_long$time <- sort(rep(1:447, 9))
cuti_long$series <- cuti_long$latitude
```

```{r echo=TRUE, message = TRUE, results = 'asis'}
head(cuti_long)
```


## Including dynamic components: coastal upwelling

* for speed, we'll filter the data to be 2023--2024 (train) and use 2025 for testing

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=3, fig.width=7}
cuti2324 <- dplyr::filter(cuti_long, year %in% 2023:2024)
ggplot(cuti2324, aes(year+month/12, cuti)) + 
  geom_line()+
  facet_wrap(~ latitude) + theme_bw() + xlab("") + ylab("CUTI")
```

## Including dynamic components: coastal upwelling

* initially no latent trend -- what are all the pieces doing??
* estimation is slooooow -- so we use data 2023--2025

```{r eval = TRUE, echo=TRUE, message=FALSE, warning=FALSE}
fit_trend <- mvgam(cuti ~ s(time, bs="cr") + 
             s(time, by = latitude, bs="fs") + 
               s(latitude, k=9, bs="re") + 
                s(month, k = 12, bs="cc"),#<<,
           data=cuti2324, 
           share_obs_params = TRUE,#<<,
           family="gaussian", chains = 1)
```

## Including dynamic components: coastal upwelling

* Smooths

```{r eval = TRUE, echo=TRUE, message=FALSE, warning=FALSE}
plot_mvgam_smooth(fit_trend, smooth = "month")
```

## Including dynamic components: coastal upwelling

* Time

```{r eval = TRUE, echo=TRUE, message=FALSE, warning=FALSE}
plot_mvgam_smooth(fit_trend, smooth = "time")
```

## Including dynamic components: coastal upwelling

* Predictions (last 3 points are forecasts)

```{r echo=FALSE, fig.height=3, fig.width=7}
newdata <- dplyr::filter(cuti_long, year > 2022)
newdata$pred <- predict(fit_trend, newdata = newdata)[,"Estimate"]

ggplot(newdata, aes(time, pred)) + 
  geom_line() + 
  geom_point(aes(time, cuti), col="red", alpha = 0.5) + 
  facet_wrap(~ latitude) + theme_bw()
```

## Including dynamic components: coastal upwelling

* Possibly some problematic residuals (autocorrelated)
```{r echo=TRUE, fig.height=3, fig.width=7}
plot(fit_trend)
```

## Including dynamic components: coastal upwelling

* Next we'll add a trend as a random walk

* By default this is creating a separate trend for each time series

* May have too many competing smooths etc

```{r eval = TRUE, echo=TRUE, message=FALSE, warning=FALSE}
fit_trend_rw <- mvgam(cuti ~ s(time, bs="cr") + 
             s(time, by = latitude, bs="fs") + 
               s(latitude, k=9, bs="re") + 
                s(month, k = 12, bs="cc"),
             trend_model = RW(),
           data=cuti_long[cuti_long$year>2022,], 
           share_obs_params = TRUE,#<<,
           family="gaussian", chains = 1)
```

## Including dynamic components: coastal upwelling

* ACF now looks a little better (and QQ plot too)
```{r echo=TRUE, fig.height=3, fig.width=7}
plot(fit_trend_rw)
```

## Including dynamic components: coastal upwelling

```{r echo=TRUE, fig.height=3, fig.width=7}
plot_mvgam_smooth(fit_trend_rw)
```

## Including dynamic components: coastal upwelling

* Each of the trends (here for Latitude 31) is ~ flat 

```{r echo=TRUE, fig.height=3, fig.width=7}
plot_mvgam_trend(fit_trend_rw)
```

## Including dynamic components: coastal upwelling

* Some fits to forecasts are better -- but training data actually seems worse

```{r echo=FALSE, fig.height=3, fig.width=7}
newdata <- dplyr::filter(cuti_long, year > 2022)
newdata$pred <- predict(fit_trend_rw, newdata = newdata)[,"Estimate"]

ggplot(newdata, aes(time, pred)) + 
  geom_line() + 
  geom_point(aes(time, cuti), col="red", alpha = 0.5) + 
  facet_wrap(~ latitude) + theme_bw()
```

## Including dynamic components: coastal upwelling

* Our smooths and the trend are competing to explain the process

* Let's drop out the main smooth now and let that be explained by the trend

```{r echo=TRUE, fig.height=3, fig.width=7}
fit_trend_ar1 <- mvgam(cuti ~ s(time, bs="cr") + 
               s(latitude, k=9, bs="re") + 
                s(month, k = 12, bs="cc"),
             trend_model = 'AR1',
           data=cuti_long[cuti_long$year>2022,], 
           share_obs_params = TRUE,
           family="gaussian", chains = 1)
```

## Including dynamic components: coastal upwelling

* Trends more flexible than the random walk appear to fit data a little better
```{r echo=FALSE, fig.height=3, fig.width=7}
newdata <- dplyr::filter(cuti_long, year > 2022)
newdata$pred <- predict(fit_trend_ar1, newdata = newdata)[,"Estimate"]

ggplot(newdata, aes(time, pred)) + 
  geom_line() + 
  geom_point(aes(time, cuti), col="red", alpha = 0.5) + 
  facet_wrap(~ latitude) + theme_bw()
```

## Including dynamic components: coastal upwelling

* But maybe we're trying to estimate too many trends

* Instead, model trends as dynamic factors (n = 2)

```{r echo=TRUE, fig.height=3, fig.width=6.5}
fit_trend_fac <- mvgam(cuti ~ s(time, bs="cr") + 
               s(latitude, k=9, bs="re") + 
                s(month, k = 12, bs="cc"),
             trend_model = 'AR1',
             use_lv = TRUE,#<<,
             n_lv = 2,#<<,
           data=cuti_long[cuti_long$year>2022,], 
           share_obs_params = TRUE,
           family="gaussian", chains = 1)
```


## Including dynamic components: coastal upwelling

* What do factors look like? 

* By default mvdlm models these as AR1

* DFA model with AR1 trend + seasonal smooth
```{r echo=TRUE, fig.height=3, fig.width=6.5}
plot(fit_trend_fac, type = 'factors')
```

## Including dynamic components: coastal upwelling

* Factor model seems to be struggling (maybe more so in middle latitudes)

* Factors for this application probably not flexible enough

```{r echo=FALSE, fig.height=3, fig.width=6.5}
newdata <- dplyr::filter(cuti_long, year > 2022)
newdata$pred <- predict(fit_trend_fac, newdata = newdata)[,"Estimate"]

ggplot(newdata, aes(time, pred)) + 
  geom_line() + 
  geom_point(aes(time, cuti), col="red", alpha = 0.5) + 
  facet_wrap(~ latitude) + theme_bw()
```