---
title: "Generalized Additive Models & Time Series"
author: ''
date: ''
output:
  slidy_presentation: default
  beamer_presentation: default
  ioslides_presentation:
    css: lecture_slides.css
subtitle: FISH 550 – Applied Time Series Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, 
                      message=FALSE, results='hide',
                      fig.align = 'center')
library(ggplot2)
library(MARSS)
library(dplyr)
library(forecast)
library(mgcv)
library(lubridate)
#library(tsDyn)
library(viridis)
```

## Overview of today's material

-   **What are Generalized Additive Models (GAMs)?**
-   **Smooth Functions, Splines, Basis Functions, and Knots**
-   **Key Differences and Similarities with Dynamic Linear Models (DLMs)**
-   **GAMs in Time Series Analysis**
-   **Applications of GAMs**

## What is a Generalized Additive Model (GAM)?

-   **GAMs** are a class of statistical models used to model complex, non-linear relationships between the response and the predictors
-   **Key idea**: GAMs model the mean of the response variable as a sum of smooth functions of the predictors

## GAM Formula

-   The expected value of the response variable is modeled as:

    $$ E[Y] = \beta_0 + f_1(x_1) + f_2(x_2) + \dots + f_k(x_k) $$

-   Where:

    -   $E[Y]$ is the expected value (mean) of the response variable $Y$.
    -   $\beta_0$ is the intercept.
    -   $f_1(x_1), f_2(x_2), \dots$ are smooth functions of the predictors $x_1, x_2, \dots$.

## Why GAMs (mgcv) are very powerful

-   **Accessible**: Formula interface similar to lm(), gam(), etc
-   **Smooths**: Non-linear smooths, etc
-   **Families**: GAMs can use non-normal families for responses
-   **Complexity**: Automates model fit and smoothness via penalization
-   **GAMM**: random effects, autocorrelation, etc

## Smooth Functions in GAMs

-   A smooth function allows us to model **non-linear relationships** in the data without needing to specify a fixed functional form
-   Smoothness is controlled by a penalty term, helping to avoid overfitting.
-   Common types of smooth functions used in GAMs:
    -   **Splines** (B-splines, cubic splines, thin-plate splines)
    -   **Kernel smoothing**
    -   **Local regression (LOESS)**
    -   **Gaussian processes**

## Smooth Functions vs. Linear Models

-   **Linear Models**: Assume a linear relationship between predictors and the response $$ E[Y] = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_k x_k $$

-   **GAMs**: Allow for flexible, non-linear relationships by fitting smooth functions: $$ E[Y] = \beta_0 + f_1(x_1) + f_2(x_2) + \dots + f_k(x_k) $$

-   **Key Advantage**: GAMs are more flexible in capturing complex trends in data that cannot be adequately modeled with a simple linear relationship (and have flexible families)

## Smooth Functions in `mgcv`

Very powerful R package for fitting GAMs

Univariate smooth terms are expressed with `s()`

```{r echo = TRUE, eval = FALSE}
gam(y ~ s(x1, k = 10, bs = "cr") + 
      s(x2, k = 10, bs = "cr"), data = data)
```

We'll cover more complicated smooths later

## Splines and Their Role in GAMs

-   **Splines**: Mathematical functions that are piecewise polynomial and used to fit smooth curves to data
-   **Why Splines?**:
    -   Allow for flexible, smooth curves that can adapt to data patterns
    -   Avoid overfitting by controlling the complexity of the curve
-   **Some Types of Splines**:
    -   **B-splines (Basis Splines)**: Commonly used due to computational efficiency and flexibility
    -   **Cubic Splines**: Splines with cubic polynomials between each pair of adjacent knots
    -   **Thin-plate Splines**: A generalization of B-splines, often used for higher-dimensional data

## What Are Knots in Splines?

-   **Knots**: Points where the pieces of the spline curve join
-   **Importance**: Knots define the structure of the spline -- more knots allows for more wiggliness
-   **Placement**: Knots can be placed:
    -   **Evenly**: Common approach, but not always optimal
    -   **Data-driven placement**: Placed where data points are dense or where significant changes in the response occur

## Visualizing B-Splines and Their Role in GAMs

To better understand how B-splines work, let’s visualize basis functions and see how they combine to form a smooth function

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}
library(ggplot2)
library(splines)

# Generate a sequence of x values
x <- seq(0, 100, by=1)

# Create B-splines with 6 degrees of freedom
spl <- splines::bs(x, df = 6)

# Visualizing B-spline basis functions
df_splines <- data.frame(x = rep(x, ncol(spl)), 
                         "b" = sort(rep(1:ncol(spl), nrow(spl))), 
                         "y" = c(spl))

# Plot individual basis functions
p1 <- ggplot(df_splines, aes(x, y, group = b, col = b)) + 
  geom_line(size = 1.2) + 
  theme_bw() + 
  xlab("X") + 
  ylab("Basis Function Value") + 
  scale_color_viridis(end = 0.8) + 
  theme(legend.position = "none") + 
  ggtitle("B-Spline Basis Functions")
print(p1)
```

## Basis Functions and B-Splines

-   **Basis Functions**: Functions that form the building blocks of splines (used to construct the overall smooth function)

-   **B-splines** are defined using a set of basis functions, each associated with a knot. The smooth curve is formed by a weighted sum of these basis functions.

-   **B-spline Basis**: The basis functions for B-splines have the following properties:

    -   They are piecewise-defined polynomials
    -   They are non-negative and have local support (i.e., each basis function is non-zero only in a limited range)

## How Basis Functions Build Splines

- **Weighted Sum of Basis Functions**: The final spline is created by summing these basis functions, each weighted by a coefficient. This allows the spline to adapt to the data and capture non-linear relationships.

    $$ f(x) = \sum_{i=1}^{n} w_i \cdot B_i(x) $$

    - \( f(x) \) is the smooth function (spline) estimated by the GAM.
    - \( B_i(x) \) are the individual basis functions.
    - \( w_i \) are the coefficients (weights) that scale each basis function to fit the data.

## How Basis Functions Build Splines

To understand how the weighted sum of basis functions creates a smooth function, let's visualize how different basis functions combine to form the final curve. 

`weights <- c(0.4, -0.5, 0.3, 0.1, 0.6, -0.2)`
```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}

# Define weights for the basis functions
weights <- c(0.4, -0.5, 0.3, 0.1, 0.6, -0.2)  # Example weights for each basis function

# Calculate the weighted sum of basis functions
weighted_spline <- spl %*% matrix(weights, ncol=1)

# data frame for plotting the basis functions and the weighted spline
df_splines_weighted <- data.frame(x = rep(x, ncol(spl)), 
                                  "b" = sort(rep(1:ncol(spl), nrow(spl))), 
                                  "basis_function" = c(spl))

# Data frame for the weighted spline
df_weighted_spline <- data.frame(x = x, 
                                  weighted_spline = weighted_spline)

# Plot both individual basis functions and the weighted sum
p3 <- ggplot() +
  geom_line(data = df_splines_weighted, aes(x, basis_function, group = b, col = as.factor(b)), size = 1.2) +
  geom_line(data = df_weighted_spline, aes(x, weighted_spline), col = "blue", size = 1.2) +
  theme_bw() +
  xlab("X") +
  ylab("Function Value") +
  scale_color_viridis_d(end = 0.8) +  
  theme(legend.position = "none") +
  ggtitle("Basis Functions and Weighted Spline")
print(p3)
```

## What if we make the basis dimension smaller?

`weights <- c(0.2, -0.1, 0.3)`

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}

# Define weights for the basis functions
weights <- c(0.2, -0.1, 0.3)  # Example weights for each basis function
spl <- splines::bs(x, df = 3)

# Calculate the weighted sum of basis functions
weighted_spline <- spl %*% matrix(weights, ncol=1)

# data frame for plotting the basis functions and the weighted spline
df_splines_weighted <- data.frame(x = rep(x, ncol(spl)), 
                                  "b" = sort(rep(1:ncol(spl), nrow(spl))), 
                                  "basis_function" = c(spl))

# Data frame for the weighted spline
df_weighted_spline <- data.frame(x = x, 
                                  weighted_spline = weighted_spline)

# Plot both individual basis functions and the weighted sum
p3 <- ggplot() +
  geom_line(data = df_splines_weighted, aes(x, basis_function, group = b, col = as.factor(b)), size = 1.2) +
  geom_line(data = df_weighted_spline, aes(x, weighted_spline), col = "blue", size = 1.2) +
  theme_bw() +
  xlab("X") +
  ylab("Function Value") +
  scale_color_viridis_d(end = 0.8) +  
  theme(legend.position = "none") +
  ggtitle("Basis Functions and Weighted Spline")
print(p3)
```

## What if we make the basis dimension smaller?

`weights <- c(0.6, -0.1, 0.01)`

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}

# Define weights for the basis functions
weights <- c(0.6, -0.1, 0.01)  # Example weights for each basis function
spl <- splines::bs(x, df = 3)

# Calculate the weighted sum of basis functions
weighted_spline <- spl %*% matrix(weights, ncol=1)

# data frame for plotting the basis functions and the weighted spline
df_splines_weighted <- data.frame(x = rep(x, ncol(spl)), 
                                  "b" = sort(rep(1:ncol(spl), nrow(spl))), 
                                  "basis_function" = c(spl))

# Data frame for the weighted spline
df_weighted_spline <- data.frame(x = x, 
                                  weighted_spline = weighted_spline)

# Plot both individual basis functions and the weighted sum
p3 <- ggplot() +
  geom_line(data = df_splines_weighted, aes(x, basis_function, group = b, col = as.factor(b)), size = 1.2) +
  geom_line(data = df_weighted_spline, aes(x, weighted_spline), col = "blue", size = 1.2) +
  theme_bw() +
  xlab("X") +
  ylab("Function Value") +
  scale_color_viridis_d(end = 0.8) +  
  theme(legend.position = "none") +
  ggtitle("Basis Functions and Weighted Spline")
print(p3)
```

## What if we make the basis dimension larger?

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}

# Define weights for the basis functions
weights <- c(0.4, -0.5, 0.3, 0.1, 0.6, -0.2, 0.4, -0.1, 0.3, -0.3, 0.2, 0.1)  # Example weights for each basis function
spl <- splines::bs(x, df = 12)

# Calculate the weighted sum of basis functions
weighted_spline <- spl %*% matrix(weights, ncol=1)

# data frame for plotting the basis functions and the weighted spline
df_splines_weighted <- data.frame(x = rep(x, ncol(spl)), 
                                  "b" = sort(rep(1:ncol(spl), nrow(spl))), 
                                  "basis_function" = c(spl))

# Data frame for the weighted spline
df_weighted_spline <- data.frame(x = x, 
                                  weighted_spline = weighted_spline)

# Plot both individual basis functions and the weighted sum
p3 <- ggplot() +
  geom_line(data = df_splines_weighted, aes(x, basis_function, group = b, col = as.factor(b)), size = 1.2) +
  geom_line(data = df_weighted_spline, aes(x, weighted_spline), col = "blue", size = 1.2) +
  theme_bw() +
  xlab("X") +
  ylab("Function Value") +
  scale_color_viridis_d(end = 0.8) +  
  theme(legend.position = "none") +
  ggtitle("Basis Functions and Weighted Spline")
print(p3)
```

## Adjusting in `mgcv`

Setting `k` controls the number of basis functions, and number of knots is generally closely related (e.g. k-2 for `cr`)
```{r echo = TRUE, eval = FALSE}
gam(y ~ s(x1, k = 10, bs = "cr") + 
      s(x2, k = 10, bs = "cr"), data = data)
```

## Extracting from mgcv

-   Basis function matrices can be extracted 
```{r eval=FALSE, echo=TRUE}
x <- seq(0, 1, length.out = 100)
sc <- smoothCon(s(x, bs = "cr", k = 10), 
      data = data.frame(x = x), 
      knots = NULL)
X_spline <- sc[[1]]$X
```

## Extracting from mgcv

- Spline matrix can be passed to lm (spline regression)
- Not exactly the same as mgcv because mgcv also uses penalty matrix

```{r eval=FALSE, echo=TRUE}
# simulate sinusoidal data + obs error
y <- sin(2 * pi * x) + rnorm(100, sd = 0.2)
# spline regression
fit <- lm(y ~ X_spline)
```

## Key Take Homes

- **It doesn't take many basis functions to create a flexible spline**

- **Flexibility is controlled by**:
  - The number of knots and how they are placed 
  - The number of basis functions used in the spline construction
    - More basis functions (higher df) -> more flexibility
  - We used 6 basis functions (df) in the last example

- **Important Considerations**:
  - Too few basis functions may lead to underfitting (not capturing the complexity of the data)
  - Too many basis functions (too high df) -> overfitting, as spline becomes too wiggly

- **Knots and Basis Functions**:
  - Knots define where the individual pieces of the spline connect
  - The placement of knots (evenly spaced vs. data-driven) can significantly impact the flexibility and accuracy of the model

## What are Knots in Splines?

- **Knots**: Points where the individual segments of the spline curve meet
- **Function of Knots**: 
  - The placement of knots determines flexibility
- **More Knots**:
  - Allow for more flexibility, creating more complex curves
  - Increase the risk of overfitting
- **Fewer Knots**:
  - Lead to smoother curves, but may underfit the data
  - Lack the ability to capture complex non-linear patterns

- **Summary**: The number and position of knots affect how well the spline can match patterns in the data

## Example: Effect of Knot Placement on Spline Fit

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}
# Simulate some data
set.seed(123)
x <- exp(rnorm(100, sd = 0.7))
y <- sin(x) + rnorm(100, sd = 0.2)

ggplot(data = data.frame(x = x, y = y), 
       aes(x, y)) + 
      xlab("Temperature") + 
      ylab("Log-odds survival") + 
  geom_point(size = 2, alpha = 0.5, col = viridis(1)) + 
  theme_bw()
```

## `mgcv` default is to distribute them based on data

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}
gam_uneven <- mgcv::gam(y ~ s(x, k = 10, bs = "cr"), data = data.frame(x = x, y = y))
knots_uneven <- gam_uneven$smooth[[1]]$xp

# Plot the data
ggplot(data = data.frame(x = x, y = y), aes(x, y)) + 
  geom_point(size = 2, alpha = 0.5, col = viridis(1)) + 
  xlab("Temperature") + 
  ylab("Log-odds survival") + 
  theme_bw() +
  #geom_vline(xintercept = knots_even, col = "red", linetype = "dashed", size = 1.2) + 
  geom_vline(xintercept = knots_uneven, col = "blue", linetype = "solid", size = 1.2) + 
  ggtitle("Knots Placement for GAMs")
```

## `mgcv` default is to distribute them based on data

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}
set.seed(123)
x <- exp(rnorm(100, sd = 0.7))
y <- sin(x) + rnorm(100, sd = 0.2)

# Fit the two GAM models
gam_uneven <- mgcv::gam(y ~ s(x, k = 10, bs = "cr"), data = data.frame(x = x, y = y))

# Fit the GAM model with custom evenly spaced knots
knots_even <- seq(min(x), max(x), length.out = 10)
gam_even <- mgcv::gam(y ~ s(x, k = 10, bs = "cr"), 
                      data = data.frame(x = x, y = y), 
                      knots = list(x = knots_even))

knots_uneven <- gam_uneven$smooth[[1]]$xp  
knots_even <- seq(min(x), max(x), length.out = 10)

new_data <- data.frame(x = seq(min(x), max(x), length.out = 100))

pred_uneven <- predict(gam_uneven, newdata = new_data, se.fit = TRUE)
pred_even <- predict(gam_even, newdata = new_data, se.fit = TRUE)

ci_uneven <- data.frame(
  x = new_data$x,
  fit = pred_uneven$fit,
  lower = pred_uneven$fit - 1.96 * pred_uneven$se.fit,
  upper = pred_uneven$fit + 1.96 * pred_uneven$se.fit
)

ci_even <- data.frame(
  x = new_data$x,
  fit = pred_even$fit,
  lower = pred_even$fit - 1.96 * pred_even$se.fit,
  upper = pred_even$fit + 1.96 * pred_even$se.fit
)

ggplot() + 
  geom_point(data = data.frame(x = x, y = y), aes(x, y), size = 2, alpha = 0.5, col = viridis(1)) +
  geom_ribbon(data = ci_uneven, aes(x = x, ymin = lower, ymax = upper), fill = "blue", alpha = 0.2) +
  geom_line(data = ci_uneven, aes(x = x, y = fit), color = "blue", size = 1.2) +
  geom_ribbon(data = ci_even, aes(x = x, ymin = lower, ymax = upper), fill = "red", alpha = 0.2) +
  geom_line(data = ci_even, aes(x = x, y = fit), color = "red", size = 1.2) +
  geom_vline(xintercept = knots_uneven, col = "blue", linetype = "solid", size = 1.2) +
  geom_vline(xintercept = knots_even, col = "red", linetype = "dashed", size = 1.2) +
  xlab("Temperature") + 
  ylab("Log-odds survival") + 
  theme_bw() + 
  ggtitle("Knots Placement and Predictions")
```

## When does knot placement matter?

-   **When the data is unevenly distributed**: If the data is concentrated in certain regions, more knots can be placed in high density regions
-   **When the relationship is non-linear**: If the relationship between the predictor and response variable is highly non-linear, adding knots can help the spline fit better
-  **When the response data has abrupt changes**: If the data is affected by regimes / change points / big changes or discontinuities

## Common Smooth Types in GAMs

- **Thin Plate Splines (`tp`)**: Flexible, non-linear smooths for complex, irregular data
- **Cubic Regression Splines (`cr`)**: Piecewise cubic polynomials, flexible with knots for general trends
- **Circular Splines (`cc`)**: For periodic data (e.g., angles, time-of-day), handles cyclical patterns
- **Gaussian Process (`gp`)**: Smooths with uncertainty estimates, used for spatial/time series models
- **B-Splines (`bs`)**: Piecewise polynomials, flexible but computationally efficient
- **P-Splines (`ps`)**: Combination of B-splines and a penalty for smoothness, good for large datasets


## GAMs in Time Series Analysis

- **Why Use GAMs for Time Series?**
  - complex, non-linear trends and seasonality
  - GAMs can capture these patterns by modeling trends as smooth functions of time
  - Aren't structured, can handle very patchy data

- **Handling Seasonality**
  - Seasonal effects (e.g., monthly, yearly) can be smoothly incorporated into the model

- **Incorporating Trend and Noise**
  - Smooth functions can represent underlying trends, while residuals capture random noise or irregularities

## GAMs and DLMs: 

* Example: daily shad counts on the Columbia River (Bonneville)

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=7}
shad <- read.csv("shad.csv")
shad$year <- lubridate::year(shad$date)

shad22_24 <- dplyr::filter(shad, year > 2021)
shad22_24$jday <- seq(1, nrow(shad22_24))
shad22_24 |>
ggplot(aes(date, log(shad))) + geom_point() + 
  ggtitle("Shad counts 2022 - 2024") + theme_bw() + ylab("ln (Shad)")
```

## GAMs and DLMs: 

* Use smooth to fill in the gaps
* `cor(obs, pred)` ~ 0.86
```{r}
fit <- gam(log(shad) ~ as.factor(year) + s(jday, bs = "cr"),
           data = shad22_24)
```

## GAMs and DLMs: 

`economics` dataset

- `psavert`: Personal savings rate

- `unemploy`: Unemployment rate

- `date`: Date of observation
```{r echo = TRUE, eval = TRUE, fig.height=3, fig.width=6}
data("economics")
economics$time_num <- as.numeric(economics$date)
economics$ln_unemploy <- log(economics$unemploy)
```

## GAMs and DLMs: 

```{r echo = FALSE, eval = TRUE, fig.height=3, fig.width=6}
p1 <- ggplot(economics, aes(time_num, psavert)) +
  geom_point() + 
  geom_smooth(method = "gam", formula = y ~ s(x, k = 10), col = "blue") +
  theme_bw() + 
  xlab("Time") + 
  ylab("Personal Savings Rate") + theme_bw()

p2 <- ggplot(economics, aes(time_num, unemploy)) +
  geom_point() + 
  geom_smooth(method = "gam", formula = y ~ s(x, k = 10), col = "blue") +
  theme_bw() + 
  xlab("Time") + 
  ylab("Unemployment") + theme_bw()
gridExtra::grid.arrange(p1, p2, ncol = 1)
```


## GAMs and DLMs: 

* What is this model doing? Where have we seen something similar??
```{r echo = TRUE, eval = TRUE, fig.height=3, fig.width=6}
# personal savings = s(time)
fit <- gam(psavert ~ s(time_num, bs = "cr"),
           data = economics)
```

## GAMs and DLMs: 

* This is fitting a smooth to describe changes in the mean, similar to an intercept-only Dynamic Linear Model 

```{r echo = TRUE, eval = TRUE, fig.height=3, fig.width=6}
plot(fit)
```

## GAMs and DLMs: 

* DLMs are flexible and allow a random walk in the intercept, covariates or both. Presumably we can do the same with a GAM. 

* How about a smooth / random walk on the covariate. Is this correct and or why not? 

```{r echo=TRUE, eval=TRUE}
# savings rate ~ s(unemployment)
fit <- gam(psavert ~ s(ln_unemploy, bs = "cr"),
           data = economics)
```

## GAMs and DLMs: 

* This is fitting a flexible non-linear model

* However, non-linear smooth of unemployment totally ignores time aspect

```{r echo = TRUE, eval = TRUE, fig.height=3, fig.width=6}
plot(fit)
```

## GAMs and DLMs: 

* If you thought we needed to be including time, you are correct

* The prior model was fitting a non-linear smooth of the covariate
 
* What about a smooth of the covariate and time?

* Here we add a 2D smooth, `s(ln_unemploy, time_num)`
```{r echo=TRUE, eval=TRUE}
fit <- gam(psavert ~ s(ln_unemploy, time_num),
           data = economics)
```

## GAMs and DLMs: 

* What is the 2D smooth of ln_unemploy and time doing?

* There's maybe some slight non-linearity here 

```{r echo = TRUE, eval = TRUE, fig.height=3, fig.width=6}
vis.gam(fit, view = c("time_num", "ln_unemploy"), plot.type = "contour", color = "heat")
```

## GAMs and DLMs

* In some cases, the 2D smooth is more complciated 

```{r echo = TRUE, eval = TRUE, fig.height=3, fig.width=6}

# Simulate predictors
set.seed(123)
n <- 400
x1 <- runif(n, 0, 10)
x2 <- runif(n, 0, 10)

# Create a response with interaction
# The response surface is non-additive: x1's effect depends on x2
y <- sin(x1) * cos(x2) + rnorm(n, sd = 0.3)

dat <- data.frame(x1 = x1, x2 = x2, y = y)

model <- gam(y ~ s(x1, x2), data = dat)

vis.gam(model, view = c("x1", "x2"), plot.type = "contour", color = "topo")
```

## GAMs and DLMs

* `s(ln_unemploy, time_num)` is fitting a 2D smooth that includes the interaction between the covariate and time

* This is a flexible model, but not the same as a time varying slope

## GAMs and DLMs:

* Another way to model the interaction is with the `by` covariate

* This lets the smooth vary / creates separate smooths for different values of the `by` covariate

```{r echo=TRUE, eval=FALSE}
s(predictor, by = covariate)
```

## GAMs and DLMs: 

* For time series data, a common mistake is to use 

```{r echo=TRUE, eval=TRUE}
fit <- gam(psavert ~ s(ln_unemploy, by = time_num),
           data = economics)
```

## GAMs and DLMs: 

* Why? 

* `s(ln_unemploy, by = time_num)` is letting the shape / smooth effect of `ln_unemploy` vary by time step
* But ordering of years isn't preserved
* Neighboring years should be similar

## GAMs and DLMs: 

```{r echo = TRUE, eval = TRUE, fig.height=3, fig.width=6}
plot(fit)
```

## GAMs and DLMs: 

* Instead we can swap order of `time_num` and `psavert` 
* This allows the effect of time on the response to differ across different levels of the covariate 

```{r echo=TRUE, eval=TRUE}
fit <- gam(psavert ~ s(time_num, by = ln_unemploy),
           data = economics)
```

* Creates a smooth that is very similar to a DLM

## GAMs and DLMs: 

```{r echo = FALSE, eval = TRUE, fig.height=3, fig.width=6}
df <- data.frame("Goal" = c("Effect of x changes over time", "Effect of time changes with x"), "Formula" = c("s(time, by = x)", "s(x, by = time)"))
knitr::kable(df)
```

## GAMs and DLMs: 

* Now we have a time-varying slope
```{r echo = TRUE, eval = TRUE, fig.height=3, fig.width=6}
plot(fit)
```

## Forecasts

* Let's think through how each of these models would do predictions 

## Non-linearity vs non-stationarity

* These models have very different assumptions
```{r}
fit_nonlinear <- gam(psavert ~ s(time_num) + s(ln_unemploy),
           data = economics)
fit_dlm <- gam(psavert ~ s(time_num, by = ln_unemploy),
           data = economics)
```

## Non-linearity vs non-stationarity

* But they give nearly the exact same predictions (rho > 0.99)

```{r echo = FALSE, fig.height=3, fig.width=6}
dat1 <- economics
dat1$Model <- "Non-linear"
dat1$pred <- predict(fit_nonlinear, newdata = dat1)

dat2 <- economics
dat2$Model <- "Time-varying"
dat2$pred <- predict(fit_dlm, newdata = dat2)

rbind(dat1, dat2) |>
  ggplot(aes(time_num, pred, col = Model)) +
  geom_line() + theme_bw()
```

## Non-linearity vs non-stationarity

* What about forecasts?
* Let's hold out the last 20 observations and try to forecast those

```{r echo = FALSE, fig.height=3, fig.width=6}
fit_nonlinear <- gam(psavert ~ s(time_num) + s(ln_unemploy),
           data = economics[1:(dim(economics)[1]-20),])
fit_dlm <- gam(psavert ~ s(time_num, by = ln_unemploy),
           data = economics[1:(dim(economics)[1]-20),])

dat1$pred <- predict(fit_nonlinear, newdata = dat1)
dat2$pred <- predict(fit_dlm, newdata = dat2)

dat1$pred_se <- predict(fit_nonlinear, newdata = dat1)
dat2$pred_se <- predict(fit_dlm, newdata = dat2)

rbind(dat1[554:574,], dat2[554:574,]) |>
  ggplot(aes(time_num, pred, col = Model)) +
  geom_line() +
  geom_point(aes(y = psavert), col = "black") +
  ggtitle("Forecasts from GAMs") + theme_bw()
```

## Non-linearity vs non-stationarity

* What about forecast uncertainty?
* What's similar / different?
```{r echo = FALSE, fig.height=3, fig.width=6}
fit_nonlinear <- gam(psavert ~ s(time_num) + s(ln_unemploy),
           data = economics[1:(dim(economics)[1]-20),])
fit_dlm <- gam(psavert ~ s(time_num, by = ln_unemploy),
           data = economics[1:(dim(economics)[1]-20),])

dat1$pred <- predict(fit_nonlinear, newdata = economics)
dat2$pred <- predict(fit_dlm, newdata = economics)

dat1$pred_se <- predict(fit_nonlinear, newdata = dat1, se.fit = TRUE)$se.fit
dat2$pred_se <- predict(fit_dlm, newdata = dat2, se.fit = TRUE)$se.fit

rbind(dat1[554:574,], dat2[554:574,]) |>
  ggplot(aes(time_num, pred, col = Model, fill = Model)) +
  geom_ribbon(aes(time_num, ymin = pred - 1.96 * pred_se, ymax = pred + 1.96 * pred_se), alpha = 0.2) +
  geom_line() +
  geom_point(aes(y = psavert), col = "black") +
  ggtitle("Forecasts from GAMs")
```

## Non-linearity vs non-stationarity

* How to decide between these 2 approaches?

* Diagnostics: R2, RMSE, AIC, etc

* `compare_gams` Available in slide Rmd

```{r echo=FALSE, warning=FALSE, message=FALSE}
compare_gams <- function(gam1, gam2, data, response, model_names = c("Non-linear GAM", "Time-varying GAM")) {
    models <- list(gam1, gam2)
    results <- list()
    
    for (i in 1:2) {
        model <- models[[i]]
        name <- model_names[i]
        
        fitted_vals <- fitted(model)
        residuals <- resid(model)
        observed <- data[[response]]
        
        # Metrics
        aic_val <- AIC(model)
        gcv_val <- model$gcv.ubre
        r2 <- summary(model)$r.sq
        rmse <- sqrt(mean((observed - fitted_vals)^2))
        
        # Normality test
        shapiro_p <- shapiro.test(residuals)$p.value
        
        # Store results
        results[[i]] <- data.frame(
            Model = name,
            AIC = round(aic_val, 2),
            GCV = round(gcv_val, 4),
            R2 = round(r2, 4),
            RMSE = round(rmse, 4),
            Shapiro_p = round(shapiro_p, 4),
            stringsAsFactors = FALSE
        )
    }
    
    # Combine into a single table
    result_table <- do.call(rbind, results)
    return(result_table)
}
```

## Non-linearity vs non-stationarity

```{r echo=TRUE, message=FALSE, warning=FALSE, results='asis'}
k <- compare_gams(fit_nonlinear, fit_dlm, response="psavert", data = economics)
print(knitr::kable(k, digits = 3, caption = "Comparison of GAMs"))
```
## Time-varying slope vs intercept

* Like with DLM, it might be hard to fit a model with a time-varying slope and intercept
* This is because the model is trying to fit a smooth function of time and a smooth function of the covariate

## Example: Time-varying slope vs intercept

```{r echo = TRUE}
data("SalmonSurvCUI")
# time varying intercept model
g1 <- gam(logit.s ~ s(year, k = 10), 
          data = SalmonSurvCUI)
# time varying slope model
g2 <- gam(logit.s ~ s(year, by = CUI.apr, k = 10), 
          data = SalmonSurvCUI)
```

## Example: Time-varying slope vs intercept

* Very different interpretations
* Similar flexibility (though here, intercept model does better)
```{r echo=FALSE, eval = TRUE, fig.height=3, fig.width=6}
p1 <- SalmonSurvCUI
p1$pred <- predict(g1)
p1$Model <- "Time varying intercept"

p2 <- SalmonSurvCUI
p2$pred <- predict(g2)
p2$Model <- "Time varying slope"

ggplot(rbind(p1, p2), aes(year, pred, col = Model)) +
  geom_line(size = 1.2) + 
  geom_point(aes(y = logit.s), col = "black") +
  theme_bw() + 
  xlab("Year") + 
  ylab("Log-odds survival")
```

## Example: Time-varying slope vs intercept

* Residuals from time-varying intercept look ok

```{r echo=FALSE, eval = TRUE, fig.height=3, fig.width=6}
plot(resid(g1))
abline(0,0,col="red")
```

## Example: Time-varying slope vs intercept

* Residuals from time-varying slope look like they're still pretty autocorrelated

```{r echo=FALSE, eval = TRUE, fig.height=3, fig.width=6}
plot(resid(g2))
abline(0,0,col="red")
```

## Summary: Time-varying slope vs intercept

* Time varying intercepts more often supported vs time varying slopes. Why?
- They capture the overall trend (level) in the data
- Covariate may not be variable enough 
- Intercepts may not be identifiable without priors, etc

* This will vary case by case, and in general you want to be careful in comparing / constructing models 

## Lots of references on GAMs

There's lots of great references out there, including:

-   Gavin Simpson's work with GAMs and time series [here](https://www.biorxiv.org/content/10.1101/322248v1)
-   Simon Wood's [book](https://www.amazon.com/Generalized-Additive-Models-Introduction-Statistical/dp/1498728332/ref=asc_df_1498728332/?tag=hyprod-20&linkCode=df0&hvadid=312065538926&hvpos=1o1&hvnetw=g&hvrand=11863417304955012193&hvpone=&hvptwo=&hvqmt=&hvdev=c&hvdvcmdl=&hvlocint=&hvlocphy=9033315&hvtargid=pla-305424873573&psc=1)
