---
title: "Fitting univariate Bayesian time series models"
author: "Eric Ward"
date: "09 Feb 2021"
output:
  ioslides_presentation:
    css: lecture_slides.css
  beamer_presentation: default
subtitle: FISH 507 – Applied Time Series Analysis
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, 
                      message=FALSE, results='hide',
                      fig.align = 'center')
library(ggplot2)
library(rstan)
library(bayesdfa)
library(atsar)
library(broom.mixed)
library(datasets)
library(MARSS)
library(dplyr)
```

## Overview of today's material

* Using STAN for MAP estimation
* Multivariate time series models
* DFA models
* Writing our own Stan code

## MAP (maximum a posteriori) estimation

* Point estimate of unknown parameters
* Optimization algorithms: BFGS, Newton, etc
* Estimates similar to maximum likelihood, but also incorporate prior
* Quick for checking models, etc

## MAP estimation in Stan

* We left off with a Bayesian DLM with random walk in intercept
```{r cache=TRUE, echo=TRUE, warning=FALSE, results=FALSE}
bayes_fit = fit_stan(y = SalmonSurvCUI$logit.s, 
               model_name="dlm-intercept")
```

## MAP estimation in Stan

* Switching this to MAP estimation
```{r cache=TRUE, echo=TRUE, warning=FALSE, results=FALSE}
map_fit = fit_stan(y = SalmonSurvCUI$logit.s, 
               model_name="dlm-intercept", map_estimation=TRUE)
```

## MAP estimation in Stan

```{r echo=TRUE, results='markdown'}
print(map_fit)
```

## MAP estimation in Stan

* Check that model converged
```{r echo=TRUE, results='markdown'}
map_fit$return_code
```

* MAP value when converged
```{r echo=TRUE, results='markdown'}
map_fit$value
```

## MAP estimation in Stan

* grep or other string matching functions needed
```{r echo=TRUE, results='markdown'}
grep("pred",names(map_fit$par))
pred = map_fit$par[grep("pred",names(map_fit$par))]
```

## MAP estimation in Stan

* grep or other string matching functions needed
```{r echo=TRUE, results='markdown'}
grep("pred",names(map_fit$par))
pred = map_fit$par[grep("pred",names(map_fit$par))]
```

## MAP estimation in Stan

* grep or other string matching functions needed
```{r echo=TRUE, results='markdown'}
grep("pred",names(map_fit$par))
pred = map_fit$par[grep("pred",names(map_fit$par))]
```

## MAP estimation in Stan

* include SEs of estimates, via Hessian
```{r echo=TRUE, results='markdown', eval=FALSE}
map_fit = fit_stan(y = SalmonSurvCUI$logit.s, 
     model_name="dlm-intercept", 
     map_estimation=TRUE, 
     hessian=TRUE)
```
* or change algorithm ("LBFGS" default) to "BFGS" or "Newton"
```{r echo=TRUE, results='markdown', eval=FALSE}
map_fit = fit_stan(y = SalmonSurvCUI$logit.s, 
     model_name="dlm-intercept", 
     map_estimation=TRUE, 
     hessian=TRUE, 
     algorithm="BFGS")
```

## MAP estimation in Stan
* Posterior means, 95% CIs
```{r}
pars = rstan::extract(bayes_fit)
df = data.frame("Time"=1:42,
                y = SalmonSurvCUI$logit.s,
                pred = apply(pars$pred,2,mean),
                lower=apply(pars$pred,2,quantile,0.025),
                upper=apply(pars$pred,2,quantile,0.975))
ggplot(df, aes(Time, pred)) + 
  geom_ribbon(aes(ymin=lower, ymax=upper),alpha=0.3) +
  theme_bw() + 
  geom_line() + ylab("Estimate") + 
  geom_point(aes(Time,y),col="red")
```

## MAP estimation in Stan
* MAP estimates = posterior
```{r}
df$map = pred
ggplot(df, aes(Time, pred)) + 
  geom_ribbon(aes(ymin=lower, ymax=upper),alpha=0.3) +
  theme_bw() + 
  geom_line() + ylab("Estimate") + 
  geom_point(aes(Time,y),col="red") + 
  geom_line(aes(Time,pred),col="blue")
```

## Multivariate time series models

* Multivariate state space models



## Bayesian DFA models

* DFA models from `atsar` bundled with other DFA code we've developed
```{r echo=TRUE}
devtools::install_github("fate-ewi/bayesdfa")
```

## Bayesian DFA models

* DFA poses interesting identifiability challenges
* Bayesian models generally involve > 1 MCMC chain
* Direction of DFA loadings / trends don't have meaning

## Bayesian DFA models

```{r}
df = data.frame("iter"=rep(1:1000,3), "chain" = sort(rep(as.factor(1:3),1000)),
                "Parameter"=c(rnorm(1000,-3), rnorm(1000,-3), rnorm(1000,3)))
ggplot(df, aes(iter,Parameter,group=chain,col=chain)) + 
  geom_line(alpha=0.7) + theme_bw() + xlab("MCMC iteration")
```

## Bayesian DFA models

* Solution 1: priors for identifiability

* Solution 2: post-hoc 'chain flipping' before convergence tests run

## Bayesian DFA models

* Fitting DFA models will give very similar answers to using MARSS
```{r echo=TRUE}
data("harborSealWA")
```
```{r fig.height=4}
matplot(harborSealWA[,-1], 
        type="l", 
        lwd=3, 
        ylab="Ln (abundance)")
```

## Bayesian DFA models

We'll extract predictions from the best model, 
```{r echo=TRUE, cache=TRUE, fig.height=2, fig.align="center",eval=TRUE}
fit = bayesdfa::fit_dfa(y = t(harborSealWA[,-1]), num_trends = 1)
```
And as an `stanfit` object, we can extract summaries of states `x`
```{r echo=TRUE}
pars = extract(fit$model)
```


## Bayesian DFA models
```{r echo=TRUE}
bayesdfa::plot_fitted(fit) + theme_bw()
```

## Bayesian DFA models

* DFA extension 1: extreme events

## Bayesian DFA models

* DFA extension 2: increased flexibility in trends

## Bayesian DFA models

* DFA extension 3: alternate constraints on loadings matrix
