---
title: "HMMs and Bayesian State-Space models"
author: "Eli Holmes, Eric Ward"
date: "Due March 1, 11:59pm 2021"
output:
  html_document:
    theme: cosmo
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r dlm-setup, include=FALSE, purl=FALSE}
#in case you forget to add a chunk label
knitr::opts_knit$set(unnamed.chunk.label = "hw6-",
                     messages=FALSE)
```

<br>


## Part 2. Fitting Bayesian State-Space Models with Stan

In the first part, you will use the wild dogs data set in the MARSS package. You will fit a random walk with drift model. The drift term ($u$ or `mu` in the code) is the rate of population growth (or decline).

```{r echo=FALSE}
data(wilddogs, package="MARSS")
plot(wilddogs[,1], wilddogs[,2])
```

Set up the data.
```{r}
data(wilddogs, package="MARSS")
```

Set up your Stan model code for a state-space random walk with drift $x_t = x_{t-1} + u + w_t$ where $w_t \sim N(0, q)$. The observation model is $y_t = x_t + v_t$ where $v_t \sim N(0, r)$. This is [ss_rw_drift.stan](https://github.com/atsa-es/atsar/blob/master/inst/stan/ss_rw_drift.stan) (with a small change in the `pos_indx` specification). Note that in the atsar code, $x$ is called `pred`, $u$ is `mu`, $\sqrt{q}$ is `sigma_process` and $\sqrt{r}$ is `sigma_obs`.

```{r}
scode <- "
data {
  int<lower=0> N;
  int<lower=0> n_pos;
  vector[n_pos] y;
  int pos_indx[n_pos];
}
parameters {
  real x0;
  real mu;
  vector[N-1] pro_dev;
  real<lower=0> sigma_process;
  real<lower=0> sigma_obs;
}
transformed parameters {
  vector[N] pred;
  pred[1] = x0;
  for(i in 2:N) {
    pred[i] = pred[i-1] + mu + sigma_process*pro_dev[i-1];
  }
}
model {
  x0 ~ normal(0,10);
  mu ~ normal(0,2);
  sigma_process ~ student_t(3,0,2);
  sigma_obs ~ student_t(3,0,2);
  pro_dev ~ std_normal();//normal(0, sigma_process);
  for(i in 1:n_pos) {
    y[i] ~ normal(pred[pos_indx[i]], sigma_obs);
  }
}
generated quantities {
  vector[n_pos] log_lik;
  for (n in 1:n_pos) log_lik[n] = normal_lpdf(y[n] | pred[pos_indx[n]], sigma_obs);
}
"
```

Fit the model. Stan does not like NAs in the data thus we use the index trick in the code. You may get a number of warnings; ignore for the homework.
```{r results="hide"}
y <- log(y <- wilddogs[,"Count"])
# We pass in the non-NA ys as vector
ypos <- y[!is.na(y)]
n_pos <- sum(!is.na(y))  # number on non-NA ys
pos_indx <- which(!is.na(y))  # index on the non-NAs
mod <- rstan::stan(model_code = scode, 
                   data = list(y = ypos, N = length(y), n_pos = n_pos, 
                               pos_indx = pos_indx), 
                   pars = c("sigma_process", "sigma_obs", "mu", "pred"), 
                   chains = 1, iter = 5000, thin = 10)
```

8. Evaluate whether the model converged using R-hat values. See slides 40 & 41 in the [Bayesian lecture](https://atsa-es.github.io/atsa/Lectures/Week%206/lec_11_bayes.html).
    
9. Make a histogram of the posterior distribution of `mu` (population growth rate). You can extract the posterior draws using `pars <- rstan::extract(mod, pars="mu")`. 
10. Using the posterior distribution of `mu`, what is the probability that the dog population is declining, i.e. the probability that `mu` is less than 0. Hint, you can use the fraction of the posterior draws that are less than 0. 

10. Make a plot of the estimated wild dog population size with 95% credible intervals. The estimated population size is the estimated $x$, called `pred` in the code. This is a plot like [slide 49](https://atsa-es.github.io/atsa/Lectures/Week%206/lec_11_bayes.html#49) in the Bayesian lecture. If you are getting stuck, look at the code that produced the figure on slide 49 by looking at the [Rmd file](https://atsa-es.github.io/atsa/Lectures/Week%206/lec_11_bayes.Rmd).

**Some ideas for optional extra analyses**

* Use the bayesplot package to make a histogram of the posteriors. See the [bayesplot documentation for examples](https://mc-stan.org/bayesplot/).

* Use the bayesplot package to make some diagnostic plots. 

* Using the interactive 'shinystan' package, does it appear the model converged? You can be qualitative and brief, with a focus just on the mu and variance parameters. `shinystan::launch_shinystan(mod)` is the code to launch shinystan.`

* The wild dog counts are discrete and the normality assumption for the observations errors for log of the counts is an approximation. Change the Stan code (the `scode` code above) to use a log linear poisson errors model. A few tips:
    + Look at the code for [dlm models](https://github.com/atsa-es/atsar/blob/master/inst/stan/dlm.stan) in the atsar package for an example.
    + You will need to change the code in 4 places: The data, to make `y` an integer, the parameters to remove `sigma_obs`, the model to use `poisson_log()`, and the likelihood to use `poisson_lpmf()`
    + When you pass in `y` to `stan()`, you will pass in `wilddogs[,"Count"]` not the log of that.
    
* For the log linear poisson model, make a plot of the posterior of `mu`. Does your conclusions about the decline rate of the dogs change?
    
* For the log linear poisson model, what is the error model? Note, it's a little tricky.





