---
title: "Dynamic Factor Analysis"
subtitle: "FISH 507 – Applied Time Series Analysis"
author: "Mark Scheuerell"
date: "31 Jan 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
set.seed(123)
```

## Topics for today

Deterministic vs stochastic elements

Dynamic Factor Analysis (DFA)

DFA with covariates


## A very simple model

Consider this simple model, consisting of a mean $\mu$ plus error

$$
y_i = \mu + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$


## A very simple model

The righthand side of the equation is composed of _deterministic_ and _stochastic_ pieces

$$
y_i = \underbrace{\mu}_{\text{deterministic}} + \underbrace{e_i}_{\text{stochastic}}
$$


## A very simple model

Sometime these pieces are referred to as _fixed_ and _random_

$$
y_i = \underbrace{\mu}_{\text{fixed}} + \underbrace{e_i}_{\text{random}}
$$


## A very simple model

This can also be seen by rewriting the model 

$$
y_i = \mu + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$

as

$$
y_i \sim \text{N}(\mu,\sigma^2)
$$


## Simple linear regression

We can expand the deterministic part of the model, as with linear regression

$$
y_i = \underbrace{\alpha + \beta x_i}_{\text{mean}} + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$

so

$$
y_i \sim \text{N}(\alpha + \beta x_i,\sigma^2)
$$


## A simple time series model

Consider a simple model with a mean $\mu$ plus white noise

$$
y_t = \mu + e_t ~ \text{with} ~ e_t \sim \text{N}(0,\sigma^2)
$$


## Time series model with covariates

We can expand the deterministic part of the model, as before with linear regression

$$
y_t = \underbrace{\alpha + \beta x_t}_{\text{mean}} + e_t ~ \text{with} ~ e_t \sim \text{N}(0,\sigma^2)
$$

so

$$
y_t \sim \text{N}(\alpha + \beta x_t,\sigma^2)
$$


## Example of linear model

```{r ts_with_AR_errors}
set.seed(123)
tt <- 30
ee <- rnorm(tt)
ww <- rnorm(tt)
phi <- 0.7
for(t in 2:tt) {
  ee[t] <- phi * ee[t-1] + ww[t]
}

aa <- 5
bb <- 1
xx <- rnorm(tt) + 2

yy <- aa + bb * xx + ee

par(mai = c(0.9,0.9,0.3,0.1), omi = c(0,0,0,0))

plot.ts(yy, ylim = c(min(xx,yy), max(xx, yy)), las = 1,
        col = "dodgerblue", lwd = 2,
        ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(xx, col = "blue", lwd = 2)
text(xx[1], expression(italic(x[t])), pos = 3, col = "blue")
text(yy[1], expression(italic(y[t])), pos = 3, col = "dodgerblue")
mtext(side = 3, expression(italic(y[t]) == 0.5~+~5~italic(x[t])~+~italic(e[t])),
      line = 0.4, adj = 0)
```


## Model residuals

```{r model_residuals}

par(mai = c(0.9,0.9,0.3,0.1), omi = c(0,0,0,0))

plot.ts(ee, col = "darkred", lwd = 2,
        ylab = expression(italic(e[t])))
```

These do _not_ look like white noise!


## ACF of model residuals

```{r ACF_model_residuals}

par(mai = c(0.9,0.9,0.3,0.1), omi = c(0,0,0,0))

acf(ee)
```

There is significant autocorrelation at lag = 1


## Model with autocorrelated errors

We can expand the stochastic part of the model to have autocorrelated errors

$$
y_t = \alpha + \beta x_t + e_t \\
e_t = \phi e_{t-1} + w_t~ \text{with} ~ w_t \sim \text{N}(0,\sigma^2)
$$


## Model with autocorrelated errors

We can also expand the stochastic part of the model to have autocorrelated errors

$$
y_t = \alpha + \beta x_t + e_t \\
e_t = \phi e_{t-1} + w_t~ \text{with} ~ w_t \sim \text{N}(0,\sigma^2)
$$

We can write this model as our standard state-space model


## Model with autocorrelated errors

Beginning with the observation equation

$$
\begin{align}
  y_t &= \alpha + \beta x_t + e_t \\
      &= e_t + \alpha + \beta x_t\\
      &\Downarrow \\
  y_t &= Z x_t + a + D d_t + v_t\\
\end{align}
$$

with

$Z = 1$, $x_t = e_t$, $a = \alpha$, $D = \beta$, $d_t = x_t$ & $v_t = 0$


## Model with autocorrelated errors

And the state equation

$$
\begin{align}
  e_t &= \phi e_{t-1} + w_t \\
      &\Downarrow \\
  x_t &= B x_t + w_t\\
\end{align}
$$

with

$x_t = e_t$ and $B = \phi$


## State-space form

Putting both parts together yields

$$
  y_t = Z x_t + a + D d_t + v_t\\
  x_t = B x_t + w_t
$$


## Topics for today

Deterministic vs stochastic elements

Dynamic Factor Analysis (DFA)

DFA with covariates
