---
title: "Dynamic Factor Analysis"
subtitle: "FISH 507 – Applied Time Series Analysis"
author: "Mark Scheuerell"
date: "31 Jan 2019"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
set.seed(123)
```

## Topics for today

Deterministic vs stochastic elements

Dynamic Factor Analysis (DFA)

DFA with covariates


## A very simple model

Consider this simple model, consisting of a mean $\mu$ plus error

$$
y_i = \mu + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$


## A very simple model

The righthand side of the equation is composed of _deterministic_ and _stochastic_ pieces

$$
y_i = \underbrace{\mu}_{\text{deterministic}} + \underbrace{e_i}_{\text{stochastic}}
$$


## A very simple model

Sometime these pieces are referred to as _fixed_ and _random_

$$
y_i = \underbrace{\mu}_{\text{fixed}} + \underbrace{e_i}_{\text{random}}
$$


## A very simple model

This can also be seen by rewriting the model 

$$
y_i = \mu + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$

as

$$
y_i \sim \text{N}(\mu,\sigma^2)
$$


## Simple linear regression

We can expand the deterministic part of the model, as with linear regression

$$
y_i = \underbrace{\alpha + \beta x_i}_{\text{mean}} + e_i ~ \text{with} ~ e_i \sim \text{N}(0,\sigma^2)
$$

so

$$
y_i \sim \text{N}(\alpha + \beta x_i,\sigma^2)
$$


## A simple time series model

Consider a simple model with a mean $\mu$ plus white noise

$$
y_t = \mu + e_t ~ \text{with} ~ e_t \sim \text{N}(0,\sigma^2)
$$


## Time series model with covariates

We can expand the deterministic part of the model, as before with linear regression

$$
y_t = \underbrace{\alpha + \beta x_t}_{\text{mean}} + e_t ~ \text{with} ~ e_t \sim \text{N}(0,\sigma^2)
$$

so

$$
y_t \sim \text{N}(\alpha + \beta x_t,\sigma^2)
$$


## Example of linear model

```{r ts_with_AR_errors}
set.seed(123)
tt <- 30
ee <- rnorm(tt)
ww <- rnorm(tt)
phi <- 0.7
for(t in 2:tt) {
  ee[t] <- phi * ee[t-1] + ww[t]
}

aa <- 5
bb <- 1
xx <- rnorm(tt) + 2

yy <- aa + bb * xx + ee

par(mai = c(0.9,0.9,0.3,0.1), omi = c(0,0,0,0))

plot.ts(yy, ylim = c(min(xx,yy), max(xx, yy)), las = 1,
        col = "dodgerblue", lwd = 2,
        ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(xx, col = "blue", lwd = 2)
text(xx[1], expression(italic(x[t])), pos = 3, col = "blue")
text(yy[1], expression(italic(y[t])), pos = 3, col = "dodgerblue")
mtext(side = 3, expression(italic(y[t]) == 0.5~+~5~italic(x[t])~+~italic(e[t])),
      line = 0.4, adj = 0)
```


## Model residuals

```{r model_residuals}

par(mai = c(0.9,0.9,0.3,0.1), omi = c(0,0,0,0))

plot.ts(ee, col = "darkred", lwd = 2,
        ylab = expression(italic(e[t])))
```

These do _not_ look like white noise!


## ACF of model residuals

```{r ACF_model_residuals}

par(mai = c(0.9,0.9,0.3,0.1), omi = c(0,0,0,0))

acf(ee)
```

There is significant autocorrelation at lag = 1


## Model with autocorrelated errors

We can expand the stochastic part of the model to have autocorrelated errors

$$
y_t = \alpha + \beta x_t + e_t \\
e_t = \phi e_{t-1} + w_t
$$

with $w_t \sim \text{N}(0,\sigma^2)$


## Model with autocorrelated errors

We can expand the stochastic part of the model to have autocorrelated errors

$$
y_t = \alpha + \beta x_t + e_t \\
e_t = \phi e_{t-1} + w_t
$$

with $w_t \sim \text{N}(0,\sigma^2)$

We can write this model as our standard state-space model


## State-space model | Observation equation

$$
\begin{align}
  y_t &= \alpha + \beta x_t + e_t \\
      &= e_t + \alpha + \beta x_t\\
      &\Downarrow \\
  y_t &= x_t + a + D d_t + v_t\\
\end{align}
$$

with

$x_t = e_t$, $a = \alpha$, $D = \beta$, $d_t = x_t$, $v_t = 0$


## State-space model | State equation


$$
\begin{align}
  e_t &= \phi e_{t-1} + w_t \\
      &\Downarrow \\
  x_t &= B x_t + w_t\\
\end{align}
$$

with

$x_t = e_t$ and $B = \phi$


## State-space model | Full form

$$
y_t = \alpha + \beta x_t + e_t \\
e_t = \phi e_{t-1} + w_t \\
\Downarrow \\
y_t = a + D d_t + x_t\\
x_t = B x_t + w_t
$$


## State-space model | In full MARSS form

$$
y_t = a + D d_t + x_t \\
x_t = B x_t + w_t \\
\Downarrow \\
y_t = Z x_t + a + D d_t + v_t \\
x_t = B x_t + u + C c_t + w_t
$$


## State-space model | Observation model in `MARSS()`

$$
y_t = a + D d_t + x_t \\
\Downarrow \\
y_t = Z x_t + a + D d_t + v_t
$$

```{r, echo=TRUE, eval=FALSE}
y = data         ## [1 x T] matrix of data
a = matrix("a")  ## intercept
D = matrix("D")  ## slope
d = covariate    ## [1 x T] matrix of measured covariate
Z = matrix(1)    ## no multiplier on x 
R = matrix(0)    ## v_t ~ N(0,R); want y_t = 0 for all t
```


## State-space model | State model in `MARSS()`

$$
x_t = B x_t + w_t \\
\Downarrow \\
x_t = B x_t + u + C c_t + w_t
$$

```{r, echo=TRUE, eval=FALSE}
B = matrix("b")  ## AR(1) coefficient for model errors
Q = matrix("q")  ## w_t ~ N(0,Q); var for model errors
u = matrix(0)    ## u = 0
C = matrix(0)    ## C = 0
c = matrix(0)    ## c_t = 0 for all t
```


## {.flexbox .vcenter}

<font size="10">MORE RANDOM EFFECTS</font>


## Expanding the random effect

Recall our simple model

$$
y_t = \underbrace{\mu}_{\text{fixed}} + \underbrace{e_t}_{\text{random}}
$$


## Expanding the random effect

We can expand the random portion

$$
y_t = \underbrace{\mu}_{\text{fixed}} + ~ \underbrace{e_t + f_t}_{\text{random}}
$$

$$
e_t \sim \text{N}(0, \sigma) \\
f_t \sim \text{N}(f_{t-1}, \gamma)
$$


## Expanding the random effect

We can expand the random portion

$$
y_t = \underbrace{\mu}_{\text{fixed}} + ~ \underbrace{e_t + f_t}_{\text{random}}
$$

$$
e_t \sim \text{N}(0, \sigma) \\
f_t \sim \text{N}(x_{t-1}, \gamma)
$$

This is simply a random walk observed with error


## Random walk observed with error

$$
y_t = x_t + a + v_t \\
x_t = x_{t-1} + w_t
$$

with

$v_t \sim \text{N}(0, R)$

$w_t \sim \text{N}(0, Q)$


## Expanding fixed & random effects

We can expand the random portion

$$
y_t = \underbrace{\alpha + \beta x_t}_{\text{fixed}} + ~ \underbrace{e_t + f_t}_{\text{random}}
$$

$$
e_t \sim \text{N}(0, \sigma) \\
f_t \sim \text{N}(f_{t-1}, \gamma)
$$


## Fixed & random effects | In familiar state-space form

$$
y_t = x_t + a + D d_t + v_t \\
x_t = x_{t-1} + w_t
$$

with

$v_t \sim \text{N}(0, R)$

$w_t \sim \text{N}(0, Q)$


## {.flexbox .vcenter}

<font size="10">MULTIPLE TIME SERIES</font>


## Simple model for 2+ time series | Random walk observed with error

$$
y_{i,t} = x_{i,t} + a_i + v_{i,t} \\
x_{i,t} = x_{i,t-1} + w_{i,t}
$$

with

$v_{i,t} \sim \text{N}(0, R)$

$w_{i,t} \sim \text{N}(0, Q)$


## Random walk observed with error

$$
y_{1,t} = x_{1,t} + a_1 + v_{1,t} \\
y_{2,t} = x_{2,t} + a_2 + v_{2,t} \\
\vdots \\
y_{n,t} = x_{n,t} + a_2 + v_{n,t} \\
$$

$$
x_{1,t} = x_{1,t-1} + w_{1,t} \\
x_{2,t} = x_{2,t-1} + w_{2,t} \\
\vdots \\
x_{n,t} = x_{n,t-1} + w_{n,t}
$$


## Random walk observed with error | In matrix form

$$
\mathbf{y}_t = \mathbf{x}_t + \mathbf{a} + \mathbf{v}_t \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t
$$

with 

$\mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R})$

$\mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})$


##

```{r plot_many_ts, echo=FALSE, fig.align='center'}
NN <- 25
TT <- 30
MM <- 3
 
## MM x TT matrix of innovations
ww <- matrix(rnorm(MM*TT, 0, 1), MM, TT)
ww[,1] <- rnorm(MM, 0, sqrt(5))
## MM x TT matrix of scaled latent trends
xx <- t(scale(apply(ww,1,cumsum)))

## loadings matrix
ZZ <- matrix(runif(NN*MM, -1, 1), NN, MM)
diag(ZZ) <- rev(sort(abs(diag(ZZ))))
ZZ[upper.tri(ZZ)] <- 0
ZZ <- round(ZZ, 2)

## obs var
obs_var <- 0.2^2
## obs errors
ee <- t(MASS::mvrnorm(TT, matrix(0,NN,1), diag(obs_var,NN,NN)))
## NN x TT matrix of observed data
yy <- ZZ %*% xx + ee

clr <- viridis::plasma(NN, alpha=0.7, end=0.8)

vv <- sample(seq(NN), NN)

par(mfrow=c(5,5), mai=c(0.1,0.1,0,0), omi=c(0,0,0,0)) 

for(i in 1:NN) {
	plot.ts(yy[vv[i],], lwd=2,
	        xlab="", xaxt="n", ylab="", yaxt="n",
	        col=clr[i], bty="n")
}
```


## Environmental time series

We often observe covariance among environmental time series, especially for those close to one another


##

```{r, echo=FALSE, fig.align='center'}
par(mfrow=c(5,5), mai=c(0.1,0.1,0,0), omi=c(0,0,0,0)) 

for(i in 1:NN) {
	plot.ts(yy[vv[i],], lwd=2,
	        xlab="", xaxt="n", ylab="", yaxt="n",
	        col=clr[i], bty="n")
}

```

Are there some common patterns here?


## Common patterns in time series

```{r plot_dfa_trends, fig.align="center"}
## plot the trends
par(mfrow=c(1,3), mai=c(1.2,0,0,0), omi=rep(0.1,4))
clr <- viridis::plasma(MM, end=0.8)
for(i in 1:3) {
	plot.ts(xx[i,], lwd=3,
	        xlab="", xaxt="n", ylab="", yaxt="n",
	        col=clr[i], bty="n")
}
```



## Topics for today

Deterministic vs stochastic elements

Dynamic Factor Analysis (DFA)

DFA with covariates
